---
title: "generate data objects for publication"
author: "Caroline Wheeler"
date: "2023-10-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readxl)
```

Publication objects = cleaned R objects in the repo, without PHI, that are the inputs to ALL of the figures

file path 
```{r}
paths.tdrive <- "/Volumes/spakowicz/"
#paths.tdrive <- "T:/Labs/Spakowicz/"
```


read in necessary data
```{r}
# pt_log <- readxl::read_xlsx(file.path(paths.tdrive,"mitox/Patient Logs/OSU 19125 Patient Log.xlsx"), skip = 3) %>%
#   dplyr::select(-`Pt Name`, -MRN)
# 
irAE <- readxl::read_xlsx(file.path(paths.tdrive,"mitox/Clinical info/Mitoxstudy_patient_symptoms_AM_2.xlsx")) %>%
  dplyr::select(-`Patient Name`, - MRN)
# 

# new sample column to include in data 
recist.scan <- readxl::read_xlsx(file.path(paths.tdrive, "mitox/Clinical info/OSU-19125_patient_progression_RECIST_KK_v4.xlsx")) %>%
  dplyr::select(-Name, -MRN)
# 
# micSamp <- read.table(file.path(paths.tdrive, 'mitox/data/labii/mitox-samples-export_2022-10-21.tsv'), sep = '\t', header = TRUE, fill = NA)
# 
# blood.adj <- openxlsx::readWorkbook(xlsxFile =  file.path(paths.tdrive, "mitox/Clinical info/Patient Data 7-1-22.xlsm"), sheet = "Adj Datasheet", startRow = 2)
# blood.met <- openxlsx::readWorkbook(xlsxFile =  file.path(paths.tdrive, "mitox/Clinical info/Patient Data 7-1-22.xlsm"), sheet = "Met Datasheet", startRow = 2)
# blood.metj <- openxlsx::readWorkbook(xlsxFile =  file.path(paths.tdrive, "mitox/Clinical info/Patient Data 7-1-22.xlsm"), sheet = "Jason Met Cohort", startRow = 2)
# 
demo.dat <- read_excel(file.path(paths.tdrive, "mitox/Patient Logs/Demographics/Patient_demographics.xlsx")) %>%
   select(-MRN, -Name)
# 
# ae.stat <- read.csv(file.path(paths.tdrive, "mitox/data/table_patients-with-AE_baseline-samps.csv" )) %>%
#   select(-mrn)
# 
# ae_att <-read.csv(file.path(paths.tdrive, "mitox/data/mitox_table_fisherexact_test.csv"))
# 
# 
pfs <- readxl::read_xlsx(file.path(paths.tdrive, "mitox/Clinical info/OSU-19125_patient_progression_RECIST_KK_v4.xlsx"),
                               sheet = "Progression dates")
# 
# ## ?
# tabone <- read.csv("../data/mitox_table1.csv") 
```

Demographics (Input to Tables 1, S1)
Rows are pts
Age, stage, response, irae
```{r}
# variables used in table 1 
# table1(data=tabone, ~ Age + Sex + Race + irAE + Cohort + Treatment |Response)

# clean up demographic data
demographics <- demo.dat %>% 
  filter(!is.na(`Patient Id`)) %>%
  select(-c(No., Notes))

# get irAE and Response
recist.mic <- recist.scan %>%
  select(`Patient ID`, RECIST, Category, Samples) %>%
  rename("Patient Id" = `Patient ID`,
         "Mic.Samp" = Samples) %>%
  filter(!is.na(`Patient Id`))

pfs <- pfs %>%
  select(`Patient Id`, `12-month PFS`)

irae <- irAE %>%
  dplyr::select(`Patient Id`, `Attribution (Final)`) %>%
  rename(irAE = `Attribution (Final)`) %>%
  filter(irAE == "Yes") %>%
  distinct()

symptoms <- irAE %>%
  dplyr::select(`Patient Id`, `Attribution (Final)`, Symptoms) %>%
  rename(irAE = `Attribution (Final)`) %>%
  filter(irAE == "Yes")

# join 
demographics <- merge(demographics, recist.mic, all=T)
demographics <- merge(demographics, pfs, all=T)
demographics <- merge(demographics, irae, all=T)

# slight mutations 
demographics <- demographics %>%
  mutate(irAE = case_when(is.na(irAE)~ "No", .default = irAE))

# add irAE groupings and microbiome samples present 
save(demographics, symptoms, file = "../data/clinical.Rdata")
```


Response = ifelse(Category %in% c("PR", "CR"), "Responder", 
                           ifelse(Category %in% c("SD", "PD"), "Non-Responder", NA))


Events object (input to Figure 1C (timeline figure))
Rows are events
ptID, event, time from C1D1 
```{r}

```


irAE table
Rows are irAEs
ptID, grade, type (input to Table S2)
```{r}

```


Microbiome sample metadata
Rows are microbiome samples
ptID, category label [baseline, 12wk, irae, control]
```{r}

```


Microbiome data
Abundances 
Rows are microbiomes
```{r}

```


Taxonomy table
```{r}

```


Blood data
Rows are blood samples
ptID, genes
```{r}

```

