---
title: "irAE Distribution"
author: "Mitchell Muniak"
date: "6/18/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(zoo)
library(reshape2)
```


This script describes the process by which i pulled irAE data from Owen et al 2019 (full citation here) in order to estimate the frequency of irAEs with different timings.

Here I input the Owen data as weeks

```{r}

irAE_weeks <- c(0, 4, 4, 4, 4, 4, 8, 8, 8, 8, 8, 12, 12, 16, 16, 24, 24, 28, 32, 32, 44, 56, 56, 72, 72, 144, 168)

```
I created a histogram to visualize the distribution.


I fiddled around with breaks and range to make a representative chart.

```{r}

hist(irAE_weeks, breaks=50, col="red", xlim = range(0,170)) 

```

I plugged in the percentile to view how many weeks will be needed to reach it.  I cotinued to return to this function and change the quantile based on what we decided for sample size.

```{r}

quantile(irAE_weeks, .5350)

```


The Owen data represented as a density graph.  A line is included on the X axis to represent the total follow-up time that we eventually decided on.

```{r}
df <- data.frame(weeks = irAE_weeks,
                 patient = seq(length(irAE_weeks)))
ggplot(df, aes(x = weeks)) +
  geom_density(fill = "blue", alpha = 0.2) +
  theme_bw() +
  geom_vline(aes(xintercept = quantile(weeks, 0.5350)), color= "red", linetype= "dashed")+
  labs(y = "Density of irAEs observed")+
ggsave("Fig_2a_sample_size.png", width = 4, height = 3)

```




```{r}
irAE_rate <- (27/91)
timing_and_samplesize <- function(compliance, samplesize, irAE_number){
  irAE_polulation_fraction <- irAE_number/(samplesize * compliance * irAE_rate)
time <- quantile(irAE_weeks, irAE_polulation_fraction)

out <- data.frame(compliance = compliance,
                  ss = samplesize,
                  irAE_number = irAE_number,
                  time = time)

return(out)
}

```
Inital sample size calculation/time with 0.7 compliance and 8 irAEs as constants.

```{r}

ss <- seq(50, 60)
newlist <- list()
for (N in ss){
  newlist[[N]] <- timing_and_samplesize(0.7, N, 8)
}
newlist <- dplyr::bind_rows(newlist)
timing_and_samplesize(0.7, 60, 8)


```

Sample size/time calculation with compliance split into 3 groups (0.7, 0.8, and 0.9)

Data is represented as a line graph.

```{r}

comp.series <- seq(0.7, 0.9, by = 0.1)
comp.list <- list()
for (c in 1:length(comp.series)) {
  
  newlist <- list()
  
  for (N in ss){
    newlist[[N]] <- timing_and_samplesize(comp.series[c], N, 8)
  }
  
  comp.list[[c]] <- dplyr::bind_rows(newlist)
}

out <- dplyr::bind_rows(comp.list)

out %>%
  mutate(compliance = factor(compliance)) %>%
  ggplot(aes(x = time, y = ss)) +
  geom_point(aes(color = compliance)) +
  geom_line(aes(group= compliance, color = compliance))

```
The graph is reconsidered with sample size as the independent variable.  We found the graph easier to read and draw conclusions from.

A line representing the average global min of the 3 compliance groups is added as a line.




```{r}

ss2 <- seq(50, 80)
comp.series <- seq(0.7, 0.9, by = 0.1)
comp.list <- list()
for (c in 1:length(comp.series)) {
  
  newlist <- list()
  
  for (N in ss2){
    newlist[[N]] <- timing_and_samplesize(comp.series[c], N, 8)
  }
  
  comp.list[[c]] <- dplyr::bind_rows(newlist)
}

outall <- 
  dplyr::bind_rows(comp.list) %>%
  mutate(total.study.time = ss/3 + time)

glomin <- outall %>%
  group_by(compliance) %>%
summarise(av_glo_min = min(total.study.time))  

all_ave <- outall %>%
  select(-time) %>%
  spread(key = "compliance", value = "total.study.time") %>%
  mutate(ave.tst = (`0.7` + `0.8` + `0.9`)/3)
all_ave%>%
  ggplot(aes(x = ss, y = ave.tst)) +
  geom_line()


outall %>%
  mutate(compliance = factor(compliance)) %>%
  ggplot(aes(x = ss, y = total.study.time)) +
  geom_point(aes(color = compliance)) +
  geom_line(aes(group= compliance, color = compliance))+
  geom_hline(yintercept = mean(glomin$av_glo_min), color = "orange")
```



The geom_point function is removed for aesthetic purposes.  The average global min line is removed.

A line representing the average point of each line is added.
```{r}

ss2 <- seq(50, 80)
comp.series <- seq(0.7, 0.9, by = 0.1)
comp.list <- list()
for (c in 1:length(comp.series)) {
  
  newlist <- list()
  
  for (N in ss2){
    newlist[[N]] <- timing_and_samplesize(comp.series[c], N, 8)
  }
  
  comp.list[[c]] <- dplyr::bind_rows(newlist)
}

outall <- 
  dplyr::bind_rows(comp.list) %>%
  mutate(total.study.time = ss/3 + time)

all_ave <- outall %>%
  select(-time) %>%
  spread(key = "compliance", value = "total.study.time") %>%
  mutate(ave.tst = (`0.7` + `0.8` + `0.9`)/3)
all_ave%>%
  ggplot(aes(x = ss, y = ave.tst)) +
  geom_line()
allx2 <- all_ave %>%
  gather(`0.7`, `0.8`, `0.9`, ave.tst, key = "compliance", value = "total.study.time")

allx2 %>%
  mutate(compliance = factor(compliance)) %>%
  ggplot(aes(x = ss, y = total.study.time)) +
  geom_line(aes(group= compliance, color = compliance))+
  theme_bw()
```
The average point line is removed.

A floor line is added.

A vertical line is added to represent our targeted sample size.


```{r}

ss2 <- seq(50, 90)
comp.series <- seq(0.7, 0.9, by = 0.1)
comp.list <- list()
for (c in 1:length(comp.series)) {
  
  newlist <- list()
  
  for (N in ss2){
    newlist[[N]] <- timing_and_samplesize(comp.series[c], N, 8)
  }
  
  comp.list[[c]] <- dplyr::bind_rows(newlist)
}

outall <- 
  dplyr::bind_rows(comp.list) %>%
  mutate(total.study.time = ss/3 + time)

#all_ave <- outall %>%
#  select(-time) %>%
#  spread(key = "compliance", value = "total.study.time") %>%
#  mutate(ave.tst = (`0.7` + `0.8` + `0.9`)/3)
#all_ave%>%
#  ggplot(aes(x = ss, y = ave.tst)) +
#  geom_line()
#allx2 <- all_ave %>%
#  gather(`0.7`, `0.8`, `0.9`, ave.tst, key = "compliance", value = "total.study.time")

outall %>%
  mutate(compliance = factor(compliance)) %>%
  mutate(floor = ss/3+12) %>%
  mutate(total.t.floor = ifelse(total.study.time <= floor, floor, total.study.time))%>%
  ggplot(aes(x = ss, y = total.t.floor)) +
  geom_line(aes(x= ss, y= floor), color= 'gray') +
  geom_line(aes(group= compliance, color = compliance))+
  theme_bw()+
  geom_vline(aes(xintercept = 63), color= 'tan')
```
Appropriate labels are added to axises.

```{r}
ss2 <- seq(50, 90)
comp.series <- seq(0.7, 0.9, by = 0.1)
comp.list <- list()
for (c in 1:length(comp.series)) {
  
  newlist <- list()
  
  for (N in ss2){
    newlist[[N]] <- timing_and_samplesize(comp.series[c], N, 8)
  }
  
  comp.list[[c]] <- dplyr::bind_rows(newlist)
}

outall <- 
  dplyr::bind_rows(comp.list) %>%
  mutate(total.study.time = ss/3 + time)

outall %>%
  mutate(compliance = factor(compliance)) %>%
  mutate(floor = ss/3+12) %>%
  mutate(total.t.floor = ifelse(total.study.time <= floor, floor, total.study.time))%>%
  ggplot(aes(x = ss, y = total.t.floor)) +
  geom_line(aes(x= ss, y= floor), color= 'gray') +
  geom_line(aes(group= compliance, color = compliance))+
  theme_bw()+
  geom_vline(aes(xintercept = 63), color= 'tan') +
  labs(x = "Sample Size", y = "Total Study Duration")+
  ggsave("Fig_2b_sample_size.png", width = 4, height = 3)
  
```