---
title: "core microbiome analysis"
author: "C. Dravillas"
date: "2025-09-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(betareg)
```

# Load data
```{r}
load("../../data/counts-wide.rds")
load("../../data/ra-wide.rds")
load("../../data/microbiome_metadata.rds")
load("../../data/clinical.rds")
```

```{r}
pts <- mic.meta |>
  group_by(`Patient Id`) |>
  filter(n() > 2) |>
  ungroup()

# n = 26
resp <- demographics |>
  select(`Patient Id`, Response, irAE) |>
  filter(`Patient Id` %in% pts$`Patient Id`)

ra.spec <- as.matrix(ra.w |>
  filter(grepl("s__", microbe)) |>
  select(microbe, pts$sample) |>
  column_to_rownames("microbe"))

all.equal(colnames(ra.spec), pts$sample)
```

Binarize the abundance matrix at species level
For each pt with > 1 sample
Drop microbes with 0’s in each sample
What fraction of the microbes are found in both (all) samples
Core.fraction = Consistent microbes / total microbes

```{r}
# binarize
mat_bin <- (ra.spec > 0) * 1

# compute core fraction per patient
core_df <- lapply(unique(pts$`Patient Id`), function(pid) {
  # get samples for this patient
  cols <- which(pts$`Patient Id` == pid)
  
  submat <- mat_bin[, cols, drop = FALSE]
  
  # microbes present in ≥1 sample
  present_any <- rowSums(submat) > 0
  
  # microbes present in all samples
  present_all <- rowSums(submat) == length(cols)
  
  total_microbes <- sum(present_any)
  consistent_microbes <- sum(present_all)
  
  core_fraction <- consistent_microbes / total_microbes
  
  data.frame(
    `Patient Id` = pid,
    core.fraction = core_fraction
  )
})

core_df <- bind_rows(core_df) |>
  dplyr::rename(`Patient Id` = Patient.Id) |>
  left_join(resp)

save(core_df, file = "../../data/core-score-min3.rds")

#write.csv(core_df, "../../../data/core-score-min2.csv")
```

Closer to 1 = more stable microbiome 
```{r}
core_df |> 
  ggplot(aes(x = Response, y = core.fraction, fill = Response)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4) +
  theme_bw() +
  labs(x = "Response", y = "Core fraction of microbes") +
  stat_compare_means(method = "wilcox.test") +
  theme(legend.position = "none")

core_df |> 
  ggplot(aes(x = irAE, y = core.fraction, fill = irAE)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4) +
  theme_bw() +
  labs(x = "Response", y = "Core fraction of microbes") +
  stat_compare_means(method = "wilcox.test") +
  theme(legend.position = "none")
```

Beta regression more appropriate for value bounded by 0 and 1 ?

```{r}
mod <- betareg(core.fraction ~ Response, data = core_df)
summary(mod)
```

```{r}
mod <- betareg(core.fraction ~ irAE, data = core_df)
summary(mod)
```

```{r}
plogis(0.3295)
plogis(0.3295 - 0.6268)
```

We modeled the core fraction (0–1) using beta regression. Those who did not 
experience an irAE had a higher estimated mean core fraction (0.58) compared to 
who did (0.43), p = 0.06, suggesting greater within-patient microbial stability.
