---
title: "core microbiome analysis"
author: "C. Dravillas"
date: "2025-09-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(betareg)
library(ggplot2)
library(ggpubr)
```

# Load data
```{r}
load("../../data/counts-wide.rds")
load("../../data/ra-wide.rds")
load("../../data/microbiome_metadata.rds")
load("../../data/clinical.rds")
load("../../data/events.rds")
```

```{r}
irAE.time <- events %>%
  filter(Event == "irAE") %>%
  mutate(irAE.day = Days) %>%
  dplyr::select(`Patient Id`, irAE.day) 

# get irae samples 
meta.irae <- mic.meta %>%
  left_join(demographics) %>%
  left_join(irAE.time) %>%
  mutate(irAE.day = ifelse(irAE.day < 0 | irAE.day > 100, NA, irAE.day),
         irAE.sample = ifelse(is.na(irAE.day), 0,
                       ifelse(abs(Days - irAE.day) <= 10, 1, 0))) %>%
  group_by(sample) %>%
  slice_max(irAE.sample, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(sample, `Patient Id`, irAE.sample)

# pts with more than 2 samples 
pts <- mic.meta |>
  group_by(`Patient Id`) |>
  filter(n() > 2) |>
  ungroup() |>
  left_join(meta.irae) |>
    group_by(`Patient Id`) %>%
  # Step 1: Keep all irAE samples
  mutate(irae_keep = ifelse(irAE.sample == 1, TRUE, FALSE)) %>%
  # Step 2: Count how many slots are left per patient
  group_modify(~{
    patient_data <- .
    n_keep <- sum(patient_data$irae_keep)
    n_slots <- 3 - n_keep
    
    if (n_slots <= 0) {
      # already have 3 or more irAE samples, keep only these
      return(patient_data %>% filter(irae_keep))
    } else {
  remaining <- patient_data %>% filter(!irae_keep)

  if (nrow(remaining) <= n_slots) {

    selected_remaining <- remaining

  } else if (n_slots == 1) {

    ## selects max here b/c thats what is farther from the irAE samples we 
    ## have but this will need to be edited in future 
    selected_remaining <- remaining %>%
      slice_max(Days, n = 1, with_ties = FALSE)

  } else if (n_slots == 2) {

    selected_remaining <- remaining %>%
      slice_min(Days, n = 1, with_ties = FALSE) %>%
      bind_rows(
        remaining %>%
          slice_max(Days, n = 1, with_ties = FALSE)
      ) %>%
      distinct()

  } else if (n_slots == 3) {

    selected_remaining <- remaining %>%
      slice_min(Days, n = 1, with_ties = FALSE) %>%
      bind_rows(
        remaining %>%
          slice_max(Days, n = 1, with_ties = FALSE)
      ) %>%
      distinct()

    middle <- remaining %>%
      filter(!Days %in% selected_remaining$Days) %>%
      slice(1)

    selected_remaining <- bind_rows(selected_remaining, middle)
  }

  bind_rows(patient_data %>% filter(irae_keep), selected_remaining)
}
  }) %>%
  ungroup() %>%
  select(-irae_keep)

# n = 26
resp <- demographics |>
  select(`Patient Id`, Response, irAE) |>
  filter(`Patient Id` %in% pts$`Patient Id`)

ra.spec <- as.matrix(ra.w |>
  filter(grepl("s__", microbe)) |>
  select(microbe, pts$sample) |>
  column_to_rownames("microbe"))

all.equal(colnames(ra.spec), pts$sample)
```

Binarize the abundance matrix at species level
For each pt with > 1 sample
Drop microbes with 0’s in each sample
What fraction of the microbes are found in both (all) samples
Core.fraction = Consistent microbes / total microbes

```{r}
# binarize
mat_bin <- (ra.spec > 0) * 1

# compute core fraction per patient
core_df <- lapply(unique(pts$`Patient Id`), function(pid) {
  # get samples for this patient
  cols <- which(pts$`Patient Id` == pid)
  
  submat <- mat_bin[, cols, drop = FALSE]
  
  # microbes present in ≥1 sample
  present_any <- rowSums(submat) > 0
  
  # microbes present in all samples
  present_all <- rowSums(submat) == length(cols)
  
  total_microbes <- sum(present_any)
  consistent_microbes <- sum(present_all)
  
  core_fraction <- consistent_microbes / total_microbes
  
  data.frame(
    `Patient Id` = pid,
    core.fraction = core_fraction
  )
})

core_df <- bind_rows(core_df) |>
  dplyr::rename(`Patient Id` = Patient.Id) |>
  left_join(resp)

save(core_df, file = "../../data/core-score-min3.rds")
load("../../data/core-score-min3.rds")

#write.csv(core_df, "../../../data/core-score-min2.csv")
```

Closer to 1 = more stable microbiome 
```{r}
core_df |> 
  ggplot(aes(x = Response, y = core.fraction, fill = Response)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4) +
  theme_bw() +
  labs(x = "Response", y = "Core fraction of microbes") +
  stat_compare_means(method = "wilcox.test") +
  theme(legend.position = "none")

core_df |> 
  ggplot(aes(x = irAE, y = core.fraction, fill = irAE)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4) +
  theme_bw() +
  labs(x = "irAE", y = "Core fraction of microbes") +
  stat_compare_means(method = "wilcox.test") +
  theme(legend.position = "none")
```

Beta regression more appropriate for value bounded by 0 and 1 ?

```{r}
mod <- betareg(core.fraction ~ Response, data = core_df)
summary(mod)
```

```{r}
mod <- betareg(core.fraction ~ irAE, data = core_df)
summary(mod)
```
Where are these numbers from again???
```{r}
plogis(0.3295)
plogis(0.3295 - 0.6268)
```

We modeled the core fraction (0–1) using beta regression. Those who did not 
experience an irAE had a higher estimated mean core fraction (0.58) compared to 
who did (0.43), p = 0.06, suggesting greater within-patient microbial stability.
