---
title: "prediction using core score"
author: "Caroline Dravillas"
date: "2026-04-08"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggplot2)
library(randomForest)
library(ROCR)
library(caTools)
library(PRROC)
library(Boruta)
```

# Load data 
```{r}
source("rf-functions.R")

load("../../data/ra-long.rds")
load("../../data/clinical.rds")
load("../../data/blood.rds")
load("../../data/microbiome_metadata.rds")
load("../../data/phylum-labels.rds")
load("../../data/ANCOM-BC2_response_baseline.Rdata")
load("../../data/ANCOM-BC2_irae_baseline.RData") 

# core score
load("../../data/core-score-min3.rds")
```

```{r}
seed_list <- c(12345:12369)
```

#q value model objects
```{r, warning=FALSE}
#Object to limit microbiome sample to one ones taken within 10 days of treatment
baseline <- mic.meta %>%
  group_by(`Patient Id`) %>%
  arrange(Days) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  filter(abs(Days) <= 10) %>%
  #subset(`Patient Id` %in% blood$`Patient Id`) |>
  # filter to samples with core score value
  # note there is also one person w/ no baseline sample
  # final n = 12; r = 8, nr = 4 
  # n = 8 if also filter to samples that have blood data 
  filter(`Patient Id` %in% core_df$`Patient Id`)

phyl_labels <- phyl_labels %>%
  mutate(taxon = substr(microbe, 4, nchar(microbe)))

#creating object for use in biome model
base.micro <- ra.l %>%
  pivot_wider(names_from = microbe, values_from = RelAbun) %>%
  subset(sample %in% baseline$sample)

# CD note - how do we feel about these cut offs being different??
#note it is dropping bacteria and archaea becaise it isn't in phyl_labels
#making object for irAE where q < 0.01
irAE_q <- res.irae %>%
  left_join(phyl_labels) %>%
  dplyr::select(microbe, q_irAEYes)  %>%
  filter(q_irAEYes < 0.005) %>%
  replace_na(list( microbe = "k__Archaea"))

irae_list <- irAE_q$microbe %>%
  append("Patient Id", after = 0)

irae_list[[1]] <- "Patient Id"

#making object for Response where q < 0.01
Response_q <- res.recist %>%
  left_join(phyl_labels) %>%
  dplyr::select(microbe, q_ResponseResponder)  %>%
  filter(q_ResponseResponder < 0.01) %>%
  replace_na(list( microbe = "k__Archaea"))

response_list <- Response_q$microbe %>%
  append("Patient Id", after = 0)

response_list[[1]] <- "Patient Id"
```

# blood gene objects
```{r}
#limiting blood samples to ones also with microbiome sample and creating object
base.gene <- blood %>%
  subset(`Patient Id` %in% baseline$`Patient Id`) %>%
  dplyr::select(`Patient Id`:YWHAZ)
```

# core score
```{r}
corescore <- core_df |>
  select(`Patient Id`, core.fraction)
```

# clinical objects
```{r}
base.clinical <- demographics %>%
  dplyr::select(`Patient Id`, 
                # binary factors
                Sex,
                Treatment.bin,
                BRAF,
                NRAS,
                NF1,
                Antibiotics,
                Probiotics,
                Corticosteroids,
                PPI,
                `prior IO`,
                
                # categorical - many levels
                Stage,
                Mel.type,
                Recurrence.Site.Cat,
                Smoker,
                
                # ordinal
                ECOG,
                # could do stage and smoking status? leaving as unordered cat
                
                # continuous 
                Age,
               `BMI.(kg/m2)`,
               cci_total)

base.clinical <- base.clinical %>% 
  mutate_at(c("Sex",
              "Treatment.bin", 
              "BRAF",
              "NRAS",
              "NF1",
              "Antibiotics",
              "Probiotics",
              "Corticosteroids",
              "PPI",
              "prior IO",
              "Stage",
              "Mel.type",
              "Smoker",
              "Recurrence.Site.Cat"), as.factor) %>%
  mutate(ECOG = ordered(ECOG,
                   levels = c(0, 1, 2)))

# clinical alone
clinical_list <- colnames(base.clinical)
# core score alone
core_list <- c("Patient Id","core.fraction")

# irae mics + core
irae_core_list <- c(irae_list, core_list[core_list != "Patient Id"])

# response mics + core
response_core_list <- c(response_list, core_list[core_list != "Patient Id"])

# irae mics + clin 
irae_clinical_list <- c(irae_list, clinical_list[clinical_list != "Patient Id"])
# response mics + clin 
response_clinical_list <- c(response_list, clinical_list[clinical_list != "Patient Id"])

# core + clin
core_clinical_list <- c(core_list, clinical_list[clinical_list != "Patient Id"])

# irae mics + core + clin
irae_core_clinical_list <- c(irae_core_list, clinical_list[clinical_list != "Patient Id"])

# response mics + core + clin 
response_core_clinical_list <- c(response_core_list, clinical_list[clinical_list != "Patient Id"])
```

# Object with all factors and outcomes to pass to rf func
```{r}
outcomes <- demographics %>%
  filter(`Patient Id` %in% baseline$`Patient Id`) |>
  select(`Patient Id`, Response, irAE) |>
    slice(match(baseline$`Patient Id`, `Patient Id`))

in.df <- baseline %>%
  dplyr::select(sample, `Patient Id`) %>%
  left_join(base.micro) %>%
  left_join(base.clinical) %>%
  left_join(base.gene) %>%
  left_join(outcomes) %>%
  left_join(corescore) %>%
  dplyr::select(-sample)

outcomes <- outcomes |>
  filter(`Patient Id` %in% in.df$`Patient Id`) |>
  arrange(match(`Patient Id`, in.df$`Patient Id`))
```
### irAE 

# Clinical model
```{r}
rf_clinical_irAE_list <- kTest(seed_list,
                               in.df,
                               "No",
                               "Yes",
                               "irAE",
                               clinical_list)

clinical_irae_aucprs <- grabVals(rf_clinical_irAE_list,
                            seed_list)

clinical_irae_aucprs[[1]]$Predictors <- "Clinical"
clinical_irae_aucprs[[2]]$Predictors <- "Clinical"
```

# Limited biome model
```{r}
rf_q005_irAE_list <- kTest(seed_list,
                           in.df,
                          "No",
                          "Yes",
                          "irAE",
                          irae_list)

biome_irae_aucprs <- grabVals(rf_q005_irAE_list,
                         seed_list)

biome_irae_aucprs[[1]]$Predictors <- "Microbiome"
biome_irae_aucprs[[2]]$Predictors <- "Microbiome"
```

# Just corescore model
```{r}
rf_core_irAE_list <- kTest(seed_list,
                            in.df,
                            "No",
                            "Yes",
                            "irAE",
                            core_list)

core_irae_aucprs <- grabVals(rf_core_irAE_list,
                         seed_list)

core_irae_aucprs[[1]]$Predictors <- "Core Score"
core_irae_aucprs[[2]]$Predictors <- "Core Score"
```

# Biome and corescore model
```{r}
rf_biome_core_irAE_list <- kTest(seed_list,
                                  in.df,
                                  "No",
                                  "Yes",
                                  "irAE",
                                  irae_core_list)

biome_and_core_irae_aucprs <- grabVals(rf_biome_core_irAE_list,
                                   seed_list)

biome_and_core_irae_aucprs[[1]]$Predictors <- "Core Score + Microbiome"
biome_and_core_irae_aucprs[[2]]$Predictors <- "Core Score + Microbiome"
```

# Corescore and clinical model
```{r}
rf_core_clinical_irAE_list <- kTest(seed_list,
                                     in.df,
                                     "No",
                                     "Yes",
                                     "irAE",
                                     core_clinical_list)

core_and_clinical_irae_aucprs <- grabVals(rf_core_clinical_irAE_list,
                                   seed_list)

core_and_clinical_irae_aucprs[[1]]$Predictors <- "Clinical + Core Score"
core_and_clinical_irae_aucprs[[2]]$Predictors <- "Clinical + Core Score"
```

# Biome and clinical model
```{r}
rf_biome_clinical_irAE_list <- kTest(seed_list,
                                     in.df,
                                     "No",
                                     "Yes",
                                     "irAE",
                                     irae_clinical_list)

biome_and_clinical_irae_aucprs <- grabVals(rf_biome_clinical_irAE_list,
                                   seed_list)

biome_and_clinical_irae_aucprs[[1]]$Predictors <- "Clinical + Microbiome"
biome_and_clinical_irae_aucprs[[2]]$Predictors <- "Clinical + Microbiome"
```

# Biome core clinical model
```{r}
rf_biome_core_clinical_irAE_list <- kTest(seed_list,
                                           in.df,
                                           "No",
                                           "Yes",
                                           "irAE",
                                           irae_core_clinical_list)

biome_core_clinical_irae_aucprs <- grabVals(rf_biome_core_clinical_irAE_list,
                                        seed_list)

biome_core_clinical_irae_aucprs[[1]]$Predictors <- "Clinical + Core Score + Microbiome"
biome_core_clinical_irae_aucprs[[2]]$Predictors <- "Clinical + Core Score + Microbiome"
```


### Response

# Clinical model
```{r}
rf_clinical_response_list <- kTest(seed_list,
                                   in.df,
                                   "Non-Responder",
                                   "Responder",
                                   "Response",
                                   clinical_list)

clinical_response_aucprs <- grabVals(rf_clinical_response_list,
                            seed_list)

clinical_response_aucprs[[1]]$Predictors <- "Clinical"
clinical_response_aucprs[[2]]$Predictors <- "Clinical"
```


# Limited biome model
```{r}
rf_q01_response_list <- kTest(seed_list,
                              in.df,
                              "Non-Responder",
                              "Responder",
                              "Response",
                              response_list)

biome_response_aucprs <- grabVals(rf_q01_response_list,
                         seed_list)

biome_response_aucprs[[1]]$Predictors <- "Microbiome"
biome_response_aucprs[[2]]$Predictors <- "Microbiome"
```

# Just core model
```{r}
rf_core_response_list <- kTest(seed_list,
                                in.df,
                                "Non-Responder",
                                "Responder",
                                "Response",
                                core_list)

core_response_aucprs <- grabVals(rf_core_response_list,
                         seed_list)

core_response_aucprs[[1]]$Predictors <- "Core Score"
core_response_aucprs[[2]]$Predictors <- "Core Score"
```

# Biome and core model
```{r}
rf_biome_core_response_list <- kTest(seed_list,
                                      in.df,
                                      "Non-Responder",
                                      "Responder",
                                      "Response",
                                      response_core_list)

biome_and_core_response_aucprs <- grabVals(rf_biome_core_response_list,
                                   seed_list)

biome_and_core_response_aucprs[[1]]$Predictors <- "Core Score + Microbiome"
biome_and_core_response_aucprs[[2]]$Predictors <- "Core Score + Microbiome"
```

# Biome clinical model
```{r}
rf_biome_clinical_response_list <- kTest(seed_list,
                                         in.df,
                                         "Non-Responder",
                                         "Responder",
                                         "Response",
                                         response_clinical_list)

biome_clinical_response_aucprs <- grabVals(rf_biome_clinical_response_list,
                                   seed_list)

biome_clinical_response_aucprs[[1]]$Predictors <- "Clinical + Microbiome"
biome_clinical_response_aucprs[[2]]$Predictors <- "Clinical + Microbiome"
```

# core clinical model
```{r}
rf_core_clinical_response_list <- kTest(seed_list,
                                         in.df,
                                         "Non-Responder",
                                         "Responder",
                                         "Response",
                                          core_clinical_list)

core_clinical_response_aucprs <- grabVals(rf_core_clinical_response_list,
                                   seed_list)

core_clinical_response_aucprs[[1]]$Predictors <- "Clinical + Core Score"
core_clinical_response_aucprs[[2]]$Predictors <- "Clinical + Core Score"
```

# Biome core clinical model
```{r}
rf_biome_core_clinical_response_list <- kTest(seed_list,
                                               in.df,
                                               "Non-Responder",
                                               "Responder",
                                               "Response",
                                               response_core_clinical_list)

biome_core_clinical_response_aucprs <- grabVals(rf_biome_core_clinical_response_list,
                                   seed_list)

biome_core_clinical_response_aucprs[[1]]$Predictors <- "Clinical + Core Score + Microbiome"
biome_core_clinical_response_aucprs[[2]]$Predictors <- "Clinical + Core Score + Microbiome"
```


### Combine and save as one df for irAE
```{r}
# Combine ROC (AUC) for irAE
roc_irae_df <- bind_rows(
  clinical_irae_aucprs[[1]],
  biome_irae_aucprs[[1]],
  core_irae_aucprs[[1]],
  biome_and_core_irae_aucprs[[1]],
  core_and_clinical_irae_aucprs[[1]],
  biome_and_clinical_irae_aucprs[[1]],
  biome_core_clinical_irae_aucprs[[1]]
)

# Combine PR (AUCPR) for irAE
pr_irae_df <- bind_rows(
  clinical_irae_aucprs[[2]],
  biome_irae_aucprs[[2]],
  core_irae_aucprs[[2]],
  biome_and_core_irae_aucprs[[2]],
  core_and_clinical_irae_aucprs[[2]],
  biome_and_clinical_irae_aucprs[[2]],
  biome_core_clinical_irae_aucprs[[2]]
)
```

### Combine and save as one df for Response
```{r}
# Combine ROC (AUC) for Response
roc_response_df <- bind_rows(
  clinical_response_aucprs[[1]],
  biome_response_aucprs[[1]],
  core_response_aucprs[[1]],
  biome_and_core_response_aucprs[[1]],
  biome_clinical_response_aucprs[[1]],
  core_clinical_response_aucprs[[1]],
  biome_core_clinical_response_aucprs[[1]]
)

# Combine PR (AUCPR) for Response
pr_response_df <- bind_rows(
  clinical_response_aucprs[[2]],
  biome_response_aucprs[[2]],
  core_response_aucprs[[2]],
  biome_and_core_response_aucprs[[2]],
  biome_clinical_response_aucprs[[2]],
  core_clinical_response_aucprs[[2]],
  biome_core_clinical_response_aucprs[[2]]
)
```

# Let's look at the variable importance
```{r}
# =========================
# RESPONSE MODELS
# =========================

rf_response_list <- list(
  clinical = rf_clinical_response_list,
  biome = rf_q01_response_list,
  core = rf_core_response_list,
  biome_core = rf_biome_core_response_list,
  biome_clinical = rf_biome_clinical_response_list,
  core_clinical = rf_core_clinical_response_list,
  biome_core_clinical = rf_biome_core_clinical_response_list
)

imp_response_list <- lapply(rf_response_list, function(x) {
  grabImp(x, seed_list)
})


# =========================
# irAE MODELS
# =========================

rf_irae_list <- list(
  clinical = rf_clinical_irAE_list,
  biome = rf_q005_irAE_list,
  core = rf_core_irAE_list,
  biome_core = rf_biome_core_irAE_list,
  biome_clinical = rf_biome_clinical_irAE_list,
  core_clinical = rf_core_clinical_irAE_list,
  biome_core_clinical = rf_biome_core_clinical_irAE_list
)

imp_irae_list <- lapply(rf_irae_list, function(x) {
  grabImp(x, seed_list)
})


```

```{r}
# get vars from all with AUC > 0.8 from microbiome input
r.vars <- get_vars_auc_threshold(rf_q01_response_list)
i.vars <- get_vars_auc_threshold(rf_q005_irAE_list)

r.vars
i.vars

intersect(r.vars, i.vars)
```

# save objects
```{r}
save(roc_irae_df, pr_irae_df, i.vars, imp_irae_list, file = "../../data/randomforest-irae-corescore.rda")
save(roc_response_df, pr_response_df, r.vars, imp_response_list, file = "../../data/randomforest-response-corescore.rda")
```

Examine averages
```{r}
pr_irae_df %>%
  select(AUCPR, Predictors, seed) %>%
  distinct() %>%
  group_by(Predictors) %>%
  summarise(avg.aucpr = mean(AUCPR)) %>%
  arrange(avg.aucpr)

pr_response_df %>%
  select(AUCPR, Predictors, seed) %>%
  distinct() %>%
  group_by(Predictors) %>%
  summarise(avg.aucpr = mean(AUCPR)) %>%
  arrange(avg.aucpr)

roc_irae_df %>%
  select(AUCROC, Predictors, seed) %>%
  distinct() %>%
  group_by(Predictors) %>%
  summarise(avg.auc = mean(AUCROC)) %>%
  arrange(avg.auc)

roc_response_df %>%
  select(AUCROC, Predictors, seed) %>%
  distinct() %>%
  group_by(Predictors) %>%
  summarise(avg.auc = mean(AUCROC)) %>%
  arrange(avg.auc)
```




