---
title: "deseq"
author: "Caroline Wheeler"
date: "2023-10-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(DESeq2)
library(tidyverse)
```

load data
```{r}
load("../../data/counts-long.rds")
load("../../data/clinical.rds")
load("../../data/microbiome_metadata.rds")
```

Response definition??
Need to have this coded the same as the pt timeline 

create inputs 
```{r}
baseline <- mic.meta %>%
  group_by(`Patient Id`) %>%
  arrange(Days) %>%
  slice_head(n = 1) %>%
  ungroup()

clin <- left_join(baseline, demographics, by = "Patient Id") %>%
  dplyr::filter(sample %in% counts.l$sample)

irae <- clin %>% 
  dplyr::select(sample, irAE) %>%
  dplyr::filter(sample %in% counts.l$sample) %>%
  distinct()

taxa_irae <- counts.l %>%
  dplyr::filter(sample %in% irae$sample)

# response as ifelse(Category == "PD", "NR", "R")) for deseq
recist <- clin %>%
  dplyr::mutate(Response = ifelse(Category == "PD", "NR", "R")) %>%
  dplyr::select(sample, Response) %>%
  dplyr::filter(sample %in% counts.l$sample,
                !is.na(Response)) %>%
  distinct()

taxa_res <- counts.l %>%
  dplyr::filter(sample %in% recist$sample)

sub <- clin %>%
  dplyr::filter(Category != "PD",
                !is.na(Category),
                sample %in% counts.l$sample) %>%
  dplyr::select(sample, irAE)

taxa_sub <- counts.l %>%
  dplyr::filter(sample %in% sub$sample)
```

Only used if estimating counts from RA 
# ```{r}
# spec.ra <- tax.l %>% filter(grepl("s__", microbe))
# ggplot(spec.ra, aes(-log10(RelAbun*2083.33333333))) + geom_histogram(binwidth = 0.5)
# ```

## Define functions

```{r}
write_desd <- function(tax){
  
  levs <- c("phylum", "class", "order", "family", "genus", "species")
  desd.list <- list()
  
  for (i in 1:length(levs)){
    desd <- tax %>%
      filter(grepl(paste0(substr(levs[i], 0, 1), "__"), microbe)) %>% 
      pivot_wider(names_from = sample, values_from = counts) %>%
      column_to_rownames("microbe") %>%
      filter_all(any_vars(. != 0))
    
    desd.list[[i]] <- desd

  }

  names(desd.list) <- levs
  return(desd.list)  
}

# function to run deseq2 given a list of count tables
# maybe loop by index instead ? 
# that way you can get name
# add ifesle to handle irae vs recist!

run_deseq <- function(desd.list, meta, design){
  
  results.list <- list()
  
  for (i in 1:length(desd.list)) {
    desd <- desd.list[[i]]
    ord <- rownames(meta)
    desd <- desd[, ord]
    
    if(colnames(meta)[1] == "irAE"){
       dds <- DESeqDataSetFromMatrix(countData = desd,
                                  colData = meta,
                                  design = ~ irAE)
    }else{
          dds <- DESeqDataSetFromMatrix(countData = desd,
                                  colData = meta,
                                  design = ~ Response)
    }

    dds <- DESeq(dds, sfType = "poscounts")
    res <- results(dds)

    pd <- res$log2FoldChange

    resPlot <- data.frame(res) %>% 
      mutate(threshold = padj < 0.05,
             microbe = rownames(.),
             level = names(desd.list)[i])
    
    results.list[[i]] <- resPlot
  }

  resPlot <- do.call("rbind", results.list)
  return(resPlot)
}
```


ERROR for genus and species levels 
Error in estimateSizeFactorsForMatrix(counts(object), locfunc = locfunc, :
every gene contains at least one zero, cannot compute log geometric means

- only known fix is adding a pseudo count 
- looks like specfiying sfType = "poscounts" also works
- have asked on the forum about this argument 


## irAE

# get irAE data set up 

```{r}
# get list of count tables 
irae.levs <- write_desd(taxa_irae)

# make sure order of samples matches 
irae <- irae %>% arrange(factor(sample, levels = colnames(irae.levs$phylum))) %>%
  remove_rownames() %>%
  column_to_rownames(var = "sample") %>%
  mutate(irAE = as.factor(irAE))
```

```{r}
resPlot_irae <- run_deseq(irae.levs, irae)
```
```{r}
save(resPlot_irae, file = "../../data/deseq_irae_baseline.rds")
```


## RECIST 

# get recist data set up 

```{r}
# get list of count tables 
res.levs <- write_desd(taxa_res)

# make sure order of samples matches 
recist <- recist %>% arrange(factor(sample, levels = colnames(irae.levs$phylum))) %>%
  remove_rownames() %>%
  column_to_rownames(var = "sample") %>%
  mutate(Response = as.factor(Response))
```

```{r}
resPlot_recist <- run_deseq(res.levs, recist)
```

```{r}
save(resPlot_recist, file = "../../data/deseq_recist_baseline.rds")
```


## SUBSET - irAE's among responders 

# get subset data set up 

```{r}
# get list of count tables 
sub.levs <- write_desd(taxa_sub)

# make sure order of samples matches 
sub <- sub %>% arrange(factor(sample, levels = colnames(irae.levs$phylum))) %>%
  remove_rownames() %>%
  column_to_rownames(var = "sample") %>%
  mutate(irAE= as.factor(irAE))
```

```{r}
resPlot_sub <- run_deseq(sub.levs, sub)
```

```{r}
save(resPlot_sub, file = "../../data/deseq_subset_baseline.rds")
```

