---
title: "figure 1 - timeline"
author: "Caroline Wheeler"
date: "2023-10-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggplot2)
```

load data
```{r}
load("../../data/clinical.rds")
load("../../data/events.rds")
load("../../data/microbiome_metadata.rds")
```

Figure 1A - Schema

Get microbiome sample numbers and distribution of timing 
n = 88 samples total
Samples can be in four categories:
- baseline = within 10 days of C1D1
  - n = 41
  
  
- irAE = within 10 days of an irAE
  - n = 12
- 12 Weeks (EOS): 
  - n = 10
- Control (all other): 
  - n = 31

Some can be in two categories. Could be irAE and baseline or irAE and 12 wk. 

```{r}
# get baseline samples 
baseline <- mic.meta %>%
  group_by(`Patient Id`) %>%
  arrange(Days) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  filter(abs(Days) <= 10)

# get irAE samples 
irAE.time <- events %>%
  filter(Event == "irAE") %>%
  mutate(irAE.day = Days) %>%
  dplyr::select(`Patient Id`, irAE.day) 

irae.samps <- mic.meta %>%
  left_join(irAE.time) %>%
  mutate(irAE.day = ifelse(irAE.day < 0 | irAE.day > 100, NA, irAE.day),
         irAE= ifelse(is.na(irAE.day), 0,
                       ifelse(abs(Days - irAE.day) <= 10, 1, 0))) %>%
  group_by(sample) %>%
  slice_max(irAE, n = 1, with_ties = FALSE) %>%
  ungroup() 

# n = 88 total samples 
mic.samps <- irae.samps |>
  mutate(Baseline = ifelse(sample %in% baseline$sample, 1, 0),
         EOS = ifelse(Days > 73 & Days < 95, 1, 0),
         Control = ifelse(Baseline == 0 & EOS == 0 & irAE == 0, 1, 0))
```

# visualize 

Baseline samples 
```{r}
mic.samps |>
  filter(Baseline == 1) %>%
  {
    m <- mean(.$Days, na.rm = TRUE)
    s <- sd(.$Days, na.rm = TRUE)

    ggplot(., aes(x = Days)) +
      geom_histogram(binwidth = 1, fill = "darkgreen", color = "darkgreen") +
      geom_vline(xintercept = m, linetype = "dashed" , linewidth = 0.5) +
      geom_vline(xintercept = c(m - s, m + s),
                 linetype = "dashed", linewidth = 0.4) +
      annotate(
        "text",
        x = Inf, y = Inf,
        label = paste0(
          "Mean = ", round(m, 1),
          "\nSD = ", round(s, 1)
        ),
        hjust = 1.1, vjust = 1.3, size = 3
      ) +
      labs(x = "Days from C1D1", 
           y = NULL) +
      theme_void() +
      theme(
        axis.line.x = element_line(),
        axis.ticks.x = element_line(),
        axis.text.x = element_text(size = 10),
        axis.title.x = element_text(size = 11),
        plot.margin = margin(5, 5, 5, 5)
      ) +
       xlim(-10, 10) +
  ggplot2::ggtitle("Baseline (n = 41)") +
  theme(plot.title = element_text(hjust = 0.5))
      
  }

ggsave("../../figures/Fig1A-Baseline.svg", height = 2, width = 2, dpi = 400)
ggsave("../../figures/Fig1A-Baseline.png", height = 2, width = 2, dpi = 400)
```

irAE samples 
```{r}
mic.samps |>
  filter(irAE == 1) %>%
  mutate(window = Days - irAE.day) %>%
  {
    m <- mean(.$window, na.rm = TRUE)
    s <- sd(.$window, na.rm = TRUE)

    ggplot(., aes(x = window)) +
      geom_histogram(binwidth = 1, fill = "darkgreen", color = "darkgreen") +
      geom_vline(xintercept = m, linetype = "dashed", linewidth = 0.5) +
      geom_vline(xintercept = c(m - s, m + s),
                 linetype = "dashed", linewidth = 0.4) +
      annotate(
        "text",
        x = Inf, y = Inf,
        label = paste0(
          "Mean = ", round(m, 1),
          "\nSD = ", round(s, 1)
        ),
        hjust = 1.1, vjust = 1.3, size = 3
      ) +
      labs(x = "Days from irAE", 
           y = NULL) +
      theme_void() +
      theme(
        axis.line.x = element_line(),
        axis.ticks.x = element_line(),
        axis.text.x = element_text(size = 10),
        axis.title.x = element_text(size = 11),
        plot.margin = margin(5, 5, 5, 5)
      ) +
      xlim(-10, 10) +
  ggplot2::ggtitle("irAE (n = 12)") +
  theme(plot.title = element_text(hjust = 0.5))
      
  }

ggsave("../../figures/Fig1A-irAE.svg", height = 2, width = 2, dpi = 400)
ggsave("../../figures/Fig1A-irAE.png", height = 2, width = 2, dpi = 400)
```
EOS samples 
```{r}
mic.samps |>
  filter(EOS == 1) %>%
  mutate(window = Days - 84) %>%
  {
    m <- mean(.$window, na.rm = TRUE)
    s <- sd(.$window, na.rm = TRUE)

    ggplot(., aes(x = window)) +
      geom_histogram(binwidth = 1, fill = "darkgreen", color = "darkgreen") +
      geom_vline(xintercept = m, linetype = "dashed", linewidth = 0.5) +
      geom_vline(xintercept = c(m - s, m + s),
                 linetype = "dashed", linewidth = 0.4) +
      annotate(
        "text",
        x = Inf, y = Inf,
        label = paste0(
          "Mean = ", round(m, 1),
          "\nSD = ", round(s, 1)
        ),
        hjust = 1.1, vjust = 1.3, size = 3
      ) +
      labs(x = "Days from 12 Weeks", 
           y = NULL) +
      theme_void() +
      theme(
        axis.line.x = element_line(),
        axis.ticks.x = element_line(),
        axis.text.x = element_text(size = 10),
        axis.title.x = element_text(size = 11),
        plot.margin = margin(5, 5, 5, 5)
      ) +
      xlim(-10, 10) +
  ggplot2::ggtitle("12 Weeks (n = 10)") +
  theme(plot.title = element_text(hjust = 0.5))
      
  }

ggsave("../../figures/Fig1A-eos.svg", height = 2, width = 2, dpi = 400)
ggsave("../../figures/Fig1A-eos.png", height = 2, width = 2, dpi = 400)
```





Figure 1B - CONSORT

# Restrict dates 
```{r}
events <- events %>%
  filter(Days <= 113,
         Days > -21) 
```


Figure 1C - Timeline 
```{r}
events.filt <- events %>%
  left_join(demographics) %>%
  # filter to Metastatic
  filter(Cohort == "Met",
  # filter to at least one microbiome sample
        Mic.Samp == "Yes",
        !is.na(Category)) %>%
  # save irAE binary in new column
  mutate(irAE.bin = ifelse(irAE == "Yes", "Yes irAE", "No irAE"),
  # create column for each time point type
         irAE = ifelse(Event == "irAE", Days, NA),
         Microbiome = ifelse(Event == "Microbiome", Days, NA),
         CT.scan = ifelse(Event == "CT Scan", Days, NA),
         Blood = ifelse(Event == "Blood Draw", Days, NA),
         Endpoints = ifelse(Event %in% c("Start Study", "End Study"), Days, NA)
  )
```

plot
```{r}
cols <- c("Yes irAE" = "cadetblue3",
          "No irAE" = "azure2",
          "irAE" = "black",
          "CT Scan" = "orange",
          "Blood" = "red3",
          "Microbiome" = "darkgreen")

shape_override <- c(2, 15, 4, 21, NA, NA)
line_override <- c(NA, NA, NA, NA, 1, 1) 
stroke_override <- c(1 ,1 , 1 , 1 , 1, 1) 
size_override <- c(1, 1, 1, 1.8, 1.8, 1)

events.filt %>%
  ggplot() + 
  geom_line(aes(x = Endpoints, y =`Patient Id`, col=irAE.bin, group = `Patient Id`), size=1.8) +
  #scale_color_manual(values = c("cadetblue3", "azure2")) +
  geom_point(aes(x = as.numeric(irAE), y =`Patient Id`, col = "irAE"), stroke = 1, shape = 4) +
  geom_point(aes(x = as.numeric(CT.scan), y = `Patient Id`, col = "CT Scan"), stroke=1, shape=15) +
  geom_point(aes(x = as.numeric(Blood), y = `Patient Id`, col = "Blood"), stroke=1, shape=2) +
  geom_point(aes(x = as.numeric(Microbiome), y = `Patient Id`, col = "Microbiome"), stroke= 1, shape=21) +
  theme_bw() +
  facet_grid(rows = vars(Response), scales = "free_y", space = "free_y") +
  scale_color_manual(values = cols, name="Timepoint") +
  guides(color = guide_legend(
    override.aes = list(
      stroke = stroke_override,
      shape = shape_override,
      linetype = line_override,
      size = size_override)
  )) +
  theme(legend.position = "top") +
  xlab("Days") +
  ylab("Patient ID") 

# version with no limits
# ggsave("../../figures/timeline-MET.png", width = 7, height = 9)

# version with limits 
ggsave("../../figures/timeline-MET-restricted-dates.png", width = 5, height = 9)
ggsave("../../figures/timeline-MET-restricted-dates.svg", width = 5, height = 9)
```


