---
title: "figure 2 - The association of gut microbes and T-cell senescence markers with treatment response"
author: "Caroline Wheeler"
date: "2023-10-24"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(ggrepel)
library(tidyverse)
library(ggfortify)
```

# Load data 
```{r}
load("../../data/ra-long.rds")
load("../../data/ra-wide.rds")
load("../../data/clinical.rds")
load("../../data/microbiome_metadata.rds")
load("../../data/counts-wide.rds")
load("../../data/counts-long.rds")
load("../../data/ANCOM-BC2_response_baseline.RData")
load("../../data/phylum-labels.rds")
```

# Figure Features 
```{r}
colors <- c("#6699CC", "#999933", "#CC6677","#44AA99","#DDCC77","#882255")

# Response ID's
resp.id <- mic.meta %>%
  left_join(demographics) %>%
  select(sample, Days, Response)

# baseline samples 
base.samps <- mic.meta %>% 
  group_by(`Patient Id`) %>%
  arrange(Days) %>%
  slice_head(n = 1) %>%
  ungroup()
```

Figure 2A - Stacked Bar (RA)
```{r}
phyl.ra <- ra.l %>%
  filter(grepl("p__", microbe))

top.phyl <- phyl.ra %>%
  group_by(microbe) %>%
  summarise(med.abun = median(RelAbun)) %>%
  arrange(desc(med.abun)) %>%
  filter(med.abun > 0)

phyl.ra <- phyl.ra %>%
  mutate(Phylum = ifelse(microbe %in% top.phyl$microbe, gsub("p__", "", microbe), 
                         "Other")) %>%
  left_join(resp.id)

# phyl.ra %>%
#   filter(RelAbun > 0) %>%
#   ggplot(aes(x = RelAbun)) +
#   facet_wrap(vars(microbe)) +
#   geom_histogram() 

# reorder samples by the total relative abundance of Firmicutes
phyl.ra <- phyl.ra %>%
  group_by(sample) %>%
  mutate(Firmicutes_Abundance = sum(RelAbun[Phylum == "Firmicutes"], na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(sample = fct_reorder(sample, Firmicutes_Abundance))

# order of microbes with 'other' last
phyl.ra <- phyl.ra %>%
  mutate(Phylum = factor(Phylum, levels = c(
    "Firmicutes", 
    setdiff(unique(Phylum), c("Firmicutes", "Other")), 
    "Other")))

# all samples 
phyl.ra %>%
  ggplot(aes(x = sample, y = RelAbun, fill = Phylum)) +
  geom_bar(position="fill", stat = "identity") +
  scale_fill_manual(values = colors) + 
  labs(x = "", y = "") +
  theme_minimal(base_size = 7) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "bottom",
        legend.box = "horizontal",
        legend.title = element_blank()) +
  facet_grid(~Response, scales = "free_x", space = "free") +
  guides(fill = guide_legend(ncol = 6, nrow = 1, byrow = TRUE))

ggsave("../../figures/fig2a-stackedbar.png", height = 4, width = 6)

#baseline samples only
phyl.ra %>%
  filter(sample %in% base.samps$sample) %>%
  ggplot(aes(x = sample, y = RelAbun, fill = Phylum)) +
  geom_bar(position="fill", stat = "identity") +
  scale_fill_manual(values = colors) + 
  labs(x = "", y = "") +
  theme_minimal(base_size = 7) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "bottom",
        legend.box = "horizontal",
        legend.title = element_blank()) +
  facet_grid(~Response, scales = "free_x", space = "free") +
  guides(fill = guide_legend(ncol = 6, nrow = 1, byrow = TRUE))

ggsave("../../figures/fig2a-stackedbar-baseline.png", height = 4, width = 6)
```

Figure 2B - Volcano Plot R vs. NR
```{r}
phyl_labels <- phyl_labels %>%
  mutate(taxon = substr(microbe, 4, nchar(microbe)))

res.recist <- res.recist %>%
  left_join(phyl_labels) %>%
  mutate(Phylum = ifelse(q_ResponseResponder < 0.05 & abs(lfc_ResponseResponder) > 2, 
                         ifelse(Phylum %in% phyl.ra$Phylum, Phylum, "Other"), NA),
         taxon = gsub("_", " ", taxon),
        Phylum = factor(Phylum, levels = c(
    "Firmicutes", "Bacteroidota", "Actinobacteria", "Proteobacteria", "Other")))

colors <- c("#6699CC", "#999933", "#CC6677","#44AA99","#882255")
```

```{r}
g <- res.recist %>%
  filter(q_ResponseResponder < 1) %>%
  ggplot(aes(x=lfc_ResponseResponder, y=-log10(q_ResponseResponder), label = taxon)) + 
  geom_point(aes(x=lfc_ResponseResponder, y=-log10(q_ResponseResponder), color = Phylum)) + 
  geom_text_repel() +
  scale_color_manual(values = colors, na.value = "honeydew3") +
  theme_minimal() +
  xlab("log2foldchange") +
  ylab("log10(p-value)") +
  theme(legend.position =  "none") + # legend matches stacked bar 
  guides(color = guide_legend(na.translate = FALSE))

g

ggsave("../../figures/fig2a-volcano-res-baseline.png", height = 6, width = 10)
```
Figure 2C - Structural Zeroes

c) structural zeroes
Top 10 as a boxplot arranged by abs(mean counts)
Boxplot by log counts, colored by phylum again, with labels for response vs non-response 
There are lots – how to refine
Check abundances of a few, particularly those with decimal values
Enrichment of structural zeroes in responders
Mosaic plot fisher exact assoc presence with response 
consider — DON’T DO THIS YET
Bar of length nrow structural zeroes
Fill = response category
P-value by permutation test
sample (response cat) * 1000 ???? 

Get top 10 arranged by abs(mean counts)

Enrichment is in Responders
R: n = 789
NR: n = 479
```{r}
# prepare count data
counts <- counts.w %>%
  pivot_longer(!microbe, names_to = "sample", values_to = "est.reads") %>%
  filter(grepl("s__", microbe)) %>%
  mutate(microbe = str_remove(microbe, "^[a-z]__"))
  
# find top 10 based on mean counts 
r <- counts %>%
  right_join(base.samps) %>%
  left_join(demographics) %>%
  filter(Response == "Responder",
         microbe %in% (struct0.recist %>% filter(`structural_zero (Response = Non-Responder)`,
                                                 !`structural_zero (Response = Responder)`))$taxon) %>%
  select(sample, microbe, est.reads)

r.top <- r %>%
  group_by(microbe) %>%
  summarize(mean.counts = mean(est.reads)) %>%
  ungroup() %>%
  mutate(Group = "Responder")

nr <- counts %>%
  right_join(base.samps) %>%
  left_join(demographics) %>%
  filter(Response == "Non-Responder",
         microbe %in% (struct0.recist %>% filter(`structural_zero (Response = Responder)`,
                                                 !`structural_zero (Response = Non-Responder)`))$taxon) %>%
  select(sample, microbe, est.reads) 

nr.top <- nr %>%
  group_by(microbe) %>%
  summarize(mean.counts = mean(est.reads)) %>%
  ungroup() %>%
  mutate(Group = "Non-Responder")

top10.counts <- rbind(r.top, nr.top) %>%
  arrange(desc(mean.counts)) %>%
  slice_head(n =10)

top10.counts
```

check top10 using RA - all the same except megamonas and dif order
```{r}
# prepare count data
ra <- ra.w %>%
  pivot_longer(!microbe, names_to = "sample", values_to = "rel.abun") %>%
  filter(grepl("s__", microbe)) %>%
  mutate(microbe = str_remove(microbe, "^[a-z]__"))
  
# find top 10 based on mean counts 
r.c <- ra %>%
  right_join(base.samps) %>%
  left_join(demographics) %>%
  filter(Response == "Responder",
         microbe %in% (struct0.recist %>% filter(`structural_zero (Response = Non-Responder)`,
                                                 !`structural_zero (Response = Responder)`))$taxon) %>%
  select(sample, microbe, rel.abun)

r.top <- r.c %>%
  group_by(microbe) %>%
  summarize(mean.ra = mean(rel.abun),
            total.nonzero = sum(rel.abun > 0)) %>%
  ungroup() %>%
  mutate(Group = "Responder")

nr.c <- ra %>%
  right_join(base.samps) %>%
  left_join(demographics) %>%
  filter(Response == "Non-Responder",
         microbe %in% (struct0.recist %>% filter(`structural_zero (Response = Responder)`,
                                                 !`structural_zero (Response = Non-Responder)`))$taxon) %>%
  select(sample, microbe, rel.abun) 

nr.top <- nr.c %>%
  group_by(microbe) %>%
  summarize(mean.ra = mean(rel.abun),
            total.nonzero = sum(rel.abun > 0)) %>%
  ungroup() %>%
  mutate(Group = "Non-Responder")

top10.ra <- rbind(r.top, nr.top) %>%
  arrange(desc(mean.ra)) %>%
  slice_head(n =10)

top10.ra.presence <- rbind(r.top, nr.top) %>%
  arrange(desc(total.nonzero), desc(mean.ra)) %>%
  slice_head(n =10)

top10.ra
top10.ra.presence
```
could instead look at present in most samples, and then mean counts?
seems like there are just a lot of counts in a few samples
nr range 1-4 species level
r range 1-8 species level
```{r}
# prepare count data
counts <- counts.w %>%
  pivot_longer(!microbe, names_to = "sample", values_to = "est.reads") %>%
  filter(grepl("s__", microbe)) %>%
  mutate(microbe = str_remove(microbe, "^[a-z]__"))
  
# find top 10 based on mean counts 
r <- counts %>%
  right_join(base.samps) %>%
  left_join(demographics) %>%
  filter(Response == "Responder",
         microbe %in% (struct0.recist %>% filter(`structural_zero (Response = Non-Responder)`,
                                                 !`structural_zero (Response = Responder)`))$taxon) %>%
  select(sample, microbe, est.reads)

r.top <- r %>%
  group_by(microbe) %>%
  summarize(total.nonzero = sum(est.reads > 0),
            percent.nonzero = total.nonzero/n_distinct(sample),
            mean.counts = mean(est.reads)) %>%
  ungroup() %>%
  mutate(Group = "Responder")

nr <- counts %>%
  right_join(base.samps) %>%
  left_join(demographics) %>%
  filter(Response == "Non-Responder",
         microbe %in% (struct0.recist %>% filter(`structural_zero (Response = Responder)`,
                                                 !`structural_zero (Response = Non-Responder)`))$taxon) %>%
  select(sample, microbe, est.reads) 

nr.top <- nr %>%
  group_by(microbe) %>%
  summarize(total.nonzero = sum(est.reads > 0),
            percent.nonzero = total.nonzero/n_distinct(sample),
            mean.counts = mean(est.reads)) %>%
  ungroup() %>%
  mutate(Group = "Non-Responder")

top10.c.mean <- rbind(r.top, nr.top) %>%
  arrange(desc(mean.counts)) %>%
  slice_head(n =10)

top10.c.presence <- rbind(r.top, nr.top) %>%
  arrange(desc(total.nonzero), desc(mean.counts)) %>%
  slice_head(n =10)

top10.percent <- rbind(r.top, nr.top) %>%
  arrange(desc(percent.nonzero), desc(mean.counts)) %>%
  slice_head(n =10)

top10.c.mean
top10.c.presence 
top10.percent
```

prep input 
```{r}
box.in <- rbind(r, nr) %>%
  filter(microbe %in% top10.percent$microbe) %>%
  left_join(mic.meta) %>%
  left_join(demographics) %>%
  dplyr::rename(taxon = microbe) %>%
  left_join(phyl_labels) %>%
  # unnecessary line given no 'Other' Phylums present
  mutate(Phylum = ifelse(Phylum %in% phyl.ra$Phylum, Phylum, "Other"),
         taxon = gsub("_", " ", taxon))


colors <- c("#999933","#6699CC","#44AA99")
```

plot

top mean counts look like crap...
maybe need to try top abundance?
maybe need to try different visualization 
top mean counts, but still very few with counts, but be just a few samples with 
VERY high abundance of certain microbes

lets try lets look at top 10 percent present -> mean counts
```{r}
box.in %>%
  ggplot(aes(x = taxon, y=log(est.reads), fill = Phylum)) +
  geom_boxplot(width = 0.5) +
  geom_jitter() + 
  scale_fill_manual(values = colors) +
  facet_grid(~ Response, scales = "free", space = "free") +
  theme_bw() +
  theme(
    panel.spacing = unit(1, "lines"), 
    axis.text.x = element_text(angle = 45, hjust = 1),  
    strip.text = element_text(size = 10),
    legend.position = "none"
  ) +
  xlab("")

ggsave("../../figures/fig2c-struct0-baseline.png", height = 6, width = 8)
```

Figure 2D - Virus Output
```{r}

```






BONUS will be own figure? - UMAP Clustering ... assuming this should be a respondersig figure?
```{r}
# create input seperately - plot here 
umap.dat <- counts.w %>%
  pivot_longer(!microbe, names_to = "sample", values_to = "count") %>%
  filter(sample %in% base.samps$sample,
         grepl("g__", microbe)) %>% 
  pivot_wider(names_from = microbe, values_from = count) %>%
  left_join(resp.id)

pca <- prcomp(umap.dat[2:933])

autoplot(pca, data = umap.dat, colour = "Response")
```

