---
title: "figure 2 - The association of gut microbes and T-cell senescence markers with treatment response"
author: "Caroline Wheeler"
date: "2023-10-24"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(ggrepel)
library(tidyverse)
library(ggfortify)
```

# Load data 
```{r}
load("../../data/ra-long.rds")
load("../../data/clinical.rds")
load("../../data/microbiome_metadata.rds")
load("../../data/counts-wide.rds")
load("../../data/counts-long.rds")
load("../../data/ANCOM-BC2_response_baseline.RData")
load("../../data/phylum-labels.rds")
```

# Figure Features 
```{r}
colors <- c("#6699CC", "#999933", "#CC6677","#44AA99","#DDCC77","#882255")

# Response ID's
resp.id <- mic.meta %>%
  left_join(demographics) %>%
  select(sample, Days, Response)

# baseline samples 
base.samps <- mic.meta %>% 
  group_by(`Patient Id`) %>%
  arrange(Days) %>%
  slice_head(n = 1) %>%
  ungroup()
```

Figure 2A - Stacked Bar (RA)
```{r}
phyl.ra <- ra.l %>%
  filter(grepl("p__", microbe))

top.phyl <- phyl.ra %>%
  group_by(microbe) %>%
  summarise(med.abun = median(RelAbun)) %>%
  arrange(desc(med.abun)) %>%
  filter(med.abun > 0)

phyl.ra <- phyl.ra %>%
  mutate(Phylum = ifelse(microbe %in% top.phyl$microbe, gsub("p__", "", microbe), 
                         "Other")) %>%
  left_join(resp.id)

# phyl.ra %>%
#   filter(RelAbun > 0) %>%
#   ggplot(aes(x = RelAbun)) +
#   facet_wrap(vars(microbe)) +
#   geom_histogram() 

# reorder samples by the total relative abundance of Firmicutes
phyl.ra <- phyl.ra %>%
  group_by(sample) %>%
  mutate(Firmicutes_Abundance = sum(RelAbun[Phylum == "Firmicutes"], na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(sample = fct_reorder(sample, Firmicutes_Abundance))

# order of microbes with 'other' last
phyl.ra <- phyl.ra %>%
  mutate(Phylum = factor(Phylum, levels = c(
    "Firmicutes", 
    setdiff(unique(Phylum), c("Firmicutes", "Other")), 
    "Other")))

# all samples 
phyl.ra %>%
  ggplot(aes(x = sample, y = RelAbun, fill = Phylum)) +
  geom_bar(position="fill", stat = "identity") +
  scale_fill_manual(values = colors) + 
  labs(x = "", y = "") +
  theme_minimal(base_size = 7) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "bottom",
        legend.box = "horizontal",
        legend.title = element_blank()) +
  facet_grid(~Response, scales = "free_x", space = "free") +
  guides(fill = guide_legend(ncol = 6, nrow = 1, byrow = TRUE))

ggsave("../../figures/fig2a-stackedbar.png", height = 4, width = 6)

#baseline samples only
phyl.ra %>%
  filter(sample %in% base.samps$sample) %>%
  ggplot(aes(x = sample, y = RelAbun, fill = Phylum)) +
  geom_bar(position="fill", stat = "identity") +
  scale_fill_manual(values = colors) + 
  labs(x = "", y = "") +
  theme_minimal(base_size = 7) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "bottom",
        legend.box = "horizontal",
        legend.title = element_blank()) +
  facet_grid(~Response, scales = "free_x", space = "free") +
  guides(fill = guide_legend(ncol = 6, nrow = 1, byrow = TRUE))

ggsave("../../figures/fig2a-stackedbar-baseline.png", height = 4, width = 6)
```

Figure 2B - Volcano Plot R vs. NR
```{r}
phyl_labels <- phyl_labels %>%
  mutate(taxon = substr(microbe, 4, nchar(microbe)))

res.recist <- res.recist %>%
  left_join(phyl_labels) %>%
  mutate(Phylum = ifelse(q_ResponseResponder < 0.05 & abs(lfc_ResponseResponder) > 2, 
                         ifelse(Phylum %in% phyl.ra$Phylum, Phylum, "Other"), NA),
         taxon = gsub("_", " ", taxon),
        Phylum = factor(Phylum, levels = c(
    "Firmicutes", "Bacteroidota", "Actinobacteria", "Proteobacteria", "Other")))

colors <- c("#6699CC", "#999933", "#CC6677","#44AA99","#882255")
```

```{r}
g <- res.recist %>%
  filter(q_ResponseResponder < 1) %>%
  ggplot(aes(x=lfc_ResponseResponder, y=-log10(q_ResponseResponder), label = taxon)) + 
  geom_point(aes(x=lfc_ResponseResponder, y=-log10(q_ResponseResponder), color = Phylum)) + 
  geom_text_repel() +
  scale_color_manual(values = colors, na.value = "honeydew3") +
  theme_minimal() +
  xlab("log2foldchange") +
  ylab("log10(p-value)") +
  theme(legend.position =  "none") + # legend matches stacked bar 
  guides(color = guide_legend(na.translate = FALSE))

g

ggsave("../../figures/fig2a-volcano-res-baseline.png", height = 6, width = 10)
```
Figure 2C - Structural Zeroes

c) structural zeroes
Top 10 as a boxplot arranged by abs(mean counts)
Boxplot by log counts, colored by phylum again, with labels for response vs non-response 
There are lots – how to refine
Check abundances of a few, particularly those with decimal values
Enrichment of structural zeroes in responders
Mosaic plot fisher exact assoc presence with response 
consider — DON’T DO THIS YET
Bar of length nrow structural zeroes
Fill = response category
P-value by permutation test
sample (response cat) * 1000 ???? 

Get top 10 arranged by abs(mean counts)

Enrichment is actually in non-responder
NR: n = 789
R: n = 479
```{r}
# prepare count data
counts <- counts.w %>%
  pivot_longer(!microbe, names_to = "sample", values_to = "est.reads") %>%
  filter(grepl("s__", microbe)) %>%
  mutate(microbe = str_remove(microbe, "^[a-z]__"))
  
# find top 10 based on mean counts 
r <- counts %>%
  right_join(base.samps) %>%
  left_join(demographics) %>%
  filter(Response == "Responder",
         microbe %in% (struct0.recist %>% filter(`structural_zero (Response = Non-Responder)`))$taxon) %>%
  select(sample, microbe, est.reads)

r.top <- r %>%
  group_by(microbe) %>%
  summarize(mean.counts = mean(est.reads)) %>%
  ungroup() %>%
  mutate(Group = "Responder")

nr <- counts %>%
  right_join(base.samps) %>%
  left_join(demographics) %>%
  filter(Response == "Non-Responder",
         microbe %in% (struct0.recist %>% filter(`structural_zero (Response = Responder)`))$taxon) %>%
  select(sample, microbe, est.reads) 

nr.top <- nr %>%
  group_by(microbe) %>%
  summarize(mean.counts = mean(est.reads)) %>%
  ungroup() %>%
  mutate(Group = "Non-Responder")

top.10 <- rbind(r.top, nr.top) %>%
  arrange(desc(mean.counts)) %>%
  slice_head(n =10)

top.10
```
prep input
```{r}
box.in <- rbind(r, nr) %>%
  filter(microbe %in% top.10$microbe) %>%
  left_join(mic.meta) %>%
  left_join(demographics) %>%
  filter(microbe %in% top.10$microbe) %>%
  dplyr::rename(taxon = microbe) %>%
  left_join(phyl_labels) %>%
  # unnecessary line given no 'Other' Phylums present
  mutate(Phylum = ifelse(Phylum %in% phyl.ra$Phylum, Phylum, "Other")) 
```

plot

these look like crap...
maybe need to try top abundance?
maybe need to try different visualization 
top mean counts, but still very few with counts, but be just a few samples with 
VERY high abundance of certain microbes
```{r}
box.in %>%
  ggplot(aes(x = microbe, y=log2(est.reads), fill = Phylum)) +
  geom_boxplot(outlier.color = "black", outlier.shape = 16) +
  geom_jitter() +
  facet_wrap(~ Response, nrow = 1) +
  theme_bw()
```



Figure 2D - Virus Output
```{r}

```






BONUS will be own figure? - UMAP Clustering ... assuming this should be a respondersig figure?
```{r}
# create input seperately - plot here 
umap.dat <- counts.w %>%
  pivot_longer(!microbe, names_to = "sample", values_to = "count") %>%
  filter(sample %in% base.samps$sample,
         grepl("g__", microbe)) %>% 
  pivot_wider(names_from = microbe, values_from = count) %>%
  left_join(resp.id)

pca <- prcomp(umap.dat[2:933])

autoplot(pca, data = umap.dat, colour = "Response")
```

