---
title: "figure 3 - The association of gut microbes and T-cell senescence markers with immune-related adverse events"
author: "Caroline Wheeler"
date: "2023-10-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(ggrepel)
library(tidyverse)
```

# Load data
```{r}
load("../../data/ra-long.rds")
load("../../data/clinical.rds")
load("../../data/microbiome_metadata.rds")
load("../../data/counts-wide.rds")
load("../../data/ANCOM-BC2_irae_baseline.RData")
load("../../data/phylum-labels.rds")
load("../../data/taxonomy-key.rds")
load("../../data/ANCOM-BC2-fulltax-sigq_NR_irAE.rds")
load("../../data/ANCOM-BC2-fulltax-sigq_R_No_irAE.rds")
load("../../data/alpha-diversity.rds")
load("../../data/respondersig-irAE-baseline.rds")
# load("../../data/vsc.rds")
# vir.key <- read.csv("../../data/CRISPR_VSG-to-species.csv", sep = "\t")
```

# Figure Features 
```{r}
colors <- c("#6699CC", "#999933", "#CC6677","#44AA99","#DDCC77","#882255")
colors <- c("cyan4", "darkgoldenrod2", "darkgreen", "salmon2", "cornflowerblue", "deeppink4") 

# Response ID's
irae.id <- mic.meta %>%
  left_join(demographics) %>%
  mutate(irAE = ifelse(irAE == "Yes", "irAE", "No irAE")) %>%
  select(sample, Days, irAE)

# baseline samples 
base.samps <- mic.meta %>% 
  group_by(`Patient Id`) %>%
  arrange(Days) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  filter(abs(Days) <= 10)
```

Figure 3A - Stacked Bar
```{r}
phyl.ra <- ra.l %>%
  filter(grepl("p__", microbe))

top.phyl <- phyl.ra %>%
  group_by(microbe) %>%
  summarise(med.abun = median(RelAbun)) %>%
  arrange(desc(med.abun)) %>%
  filter(med.abun > 0)

phyl.ra <- phyl.ra %>%
  mutate(Phylum = ifelse(microbe %in% top.phyl$microbe, gsub("p__", "", microbe), 
                         "Other")) %>%
  left_join(irae.id)

# reorder samples by the total relative abundance of Firmicutes
phyl.ra <- phyl.ra %>%
  group_by(sample) %>%
  mutate(Firmicutes_Abundance = sum(RelAbun[Phylum == "Firmicutes"], na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(sample = fct_reorder(sample, Firmicutes_Abundance))

# order of microbes with 'other' last
phyl.ra <- phyl.ra %>%
  mutate(Phylum = factor(Phylum, levels = c(
    "Firmicutes", 
    setdiff(unique(Phylum), c("Firmicutes", "Other")), 
    "Other")))

# all samples 
# phyl.ra %>%
#   ggplot(aes(x = sample, y = RelAbun, fill = Phylum)) +
#   geom_bar(position="fill", stat = "identity") +
#   scale_fill_manual(values = colors) + 
#   labs(x = "", y = "") +
#   theme_minimal(base_size = 7) +
#   theme(axis.text.x = element_blank(),
#         axis.ticks.x = element_blank(),
#         axis.text.y = element_blank(),
#         panel.grid.major = element_blank(),
#         legend.position = "bottom",
#         legend.box = "horizontal",
#         legend.title = element_blank()) +
#   facet_grid(~irAE, scales = "free_x", space = "free") +
#   guides(fill = guide_legend(ncol = 6, nrow = 1, byrow = TRUE))


#baseline samples only
phyl.ra %>%
  filter(sample %in% base.samps$sample) %>%
  ggplot(aes(x = sample, y = RelAbun, fill = Phylum)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_fill_manual(values = colors) + 
  labs(x = "", y = "") +
  theme_minimal(base_size = 9) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_blank(),
    panel.grid.major = element_blank(),
    legend.position = "top",
    legend.box = "horizontal",
    legend.title = element_blank(),
    strip.placement = "outside", # Moves facet labels outside
    strip.background = element_blank(), # Optional: Removes the background of facet labels
    strip.text.x = element_text(size = 9) # Adjusts facet label text size
  ) +
  facet_grid(~irAE, scales = "free_x", space = "free", switch = "x") +
  guides(fill = guide_legend(ncol = 6, nrow = 1, byrow = TRUE))


ggsave("../../figures/Fig3A-irAE-baseline-stackedbar.png", height = 3.5, width = 5, dpi = 300)
```
Figure 3B - diversity boxplots
```{r}
div  %>%
  filter(data.type == "RA",
         Signature != "Inverse.Simpson") %>%
  ggplot(aes(x=irAE.status, y=log(Score))) +
  geom_boxplot(aes(fill = irAE.status)) + 
  facet_wrap(~Signature, scales = "free") + 
  theme_bw(base_size = 7) +
  scale_fill_manual(values=c("#CC6677", "#F21A00")) +
  theme(legend.position = "none") +
  stat_compare_means(
    method = "t.test",
    size = 2.5,          
    vjust = 11          
  )

ggsave("../../figures/Fig3B_boxplots_diversity-irAE.png", width = 3, height = 3, dpi = 300)
```

Figure 3C - Volcano Plot irAE vs. No irAE
initially irAE yes will be on the right, but want to swap so that R and No irAE
are on the same side
```{r}
phyl_labels <- phyl_labels %>%
  mutate(taxon = substr(microbe, 4, nchar(microbe)))

res.irae <- res.irae %>%
  left_join(phyl_labels) %>%
  mutate(Phylum = ifelse(q_irAEYes < 0.05 & abs(lfc_irAEYes) > 1, 
                         ifelse(Phylum %in% phyl.ra$Phylum, Phylum, "Other"), NA),
         taxon = gsub("_", " ", taxon),
        Phylum = factor(Phylum, levels = c(
    "Firmicutes", "Bacteroidota", "Actinobacteria", "Proteobacteria", "Other")),
    label = ifelse(taxon %in% sig.NR.irAE$taxon | taxon %in% sig.R.NoirAE$taxon, taxon, NA),
    lfc_irAEYes = -lfc_irAEYes)

colors <- c("#6699CC", "#999933", "#CC6677","#44AA99","#882255")
colors <- c("cyan4", "darkgoldenrod2", "darkgreen", "salmon2", "deeppink4") 
```

```{r}
g <- res.irae %>%
  filter(q_irAEYes < 1) %>%
  ggplot(aes(x=lfc_irAEYes, y=-log10(q_irAEYes), label = label)) + 
  geom_point(aes(x=lfc_irAEYes, y=-log10(q_irAEYes), color = Phylum)) + 
  geom_text_repel(size = 2) +
  scale_color_manual(values = colors, na.value = "honeydew3") +
  theme_bw(base_size = 9) +
  xlab("log2foldchange") +
  ylab("log10(p-value)") +
  theme(legend.position =  "none") + # legend matches stacked bar 
  guides(color = guide_legend(na.translate = FALSE)) 

g

ggsave("../../figures/Fig3C-irae-baseline-volcano.png", height = 4, width = 5, dpi = 300)
```

Figure 3D - Respondersig 

```{r}
keep_idx <- rsig.baseline$Group == "Experimental"

rsig.filt <- lapply(rsig.baseline, function(x) {
  if (is.data.frame(x)) {
    x[keep_idx, , drop = FALSE]
  } else if (length(x) == length(rsig.baseline$Group)) {
    x[keep_idx]
  } else {
    x  
  }
})

# plot
rsig.plot <- rsig_plot(rsig.filt, batch.correction = "Log.transformed")

ggsave("../../figures/Fig3D-respondersig-plot.svg", plot = rsig.plot, height = 4, width = 4, dpi = 500)
```


# ```{r}
# # create hull var
# hull_var <- rsig.baseline$Response
# hull_var[hull_var %in% c("Responder", "Non_Responder")] <- NA
# 
# # plot and save 
# rsig.plot <- rsig_plot(rsig.baseline, batch.correction = "Log.transformed", hull = hull_var)
# ggsave("../../figures/Fig3D-respondersig-plot.svg", plot = rsig.plot, height = 4, width = 4, dpi = 500)
# ```

irAE and no-irAE groups do not differ significantly 
but could show distance to centroid here 
```{r}
get_pval <- function(dist.in, group, perm = 999, seed = 528934) {
  set.seed(seed)
  keep <- !is.na(group)
  data_in <- dist.in[keep, , drop = FALSE]
  group <- droplevels(group[keep])+

  adon.results <- adonis2(data_in ~ group, 
                          method = "euclidean", 
                          permutations = perm)

  return(adon.results)
}

# for log.transformed input
pval <- get_pval(rsig.filt$Log.transformed, rsig.filt$Response)

print("Log-transformed")
pval

```


Figure 3E - Structural Zero Box plots

```{r}
counts <- counts.w %>%
  pivot_longer(!microbe, names_to = "sample", values_to = "est.reads") %>%
  filter(grepl("s__", microbe)) %>%
  mutate(microbe = str_remove(microbe, "^[a-z]__"))

irae <- counts %>%
  right_join(base.samps) %>%
  left_join(demographics) %>%
  filter(irAE == "Yes",
         microbe %in% (struct0.irae %>% filter(`structural_zero (irAE = No)` &
                                                 !`structural_zero (irAE = Yes)`))$taxon) %>%
  select(sample, microbe, est.reads)

irae.top <- irae %>%
  group_by(microbe) %>%
    summarize(total.nonzero = sum(est.reads > 0),
            percent.nonzero = total.nonzero/n_distinct(sample),
            mean.counts = mean(est.reads)) %>%
  ungroup() %>%
  mutate(iraeGroup = "irAE")

no_irae <- counts %>%
  right_join(base.samps) %>%
  left_join(demographics) %>%
  filter(irAE == "No",
         microbe %in% (struct0.irae %>% filter(`structural_zero (irAE = Yes)` & 
                                                 !`structural_zero (irAE = No)`))$taxon) %>%
  select(sample, microbe, est.reads)

no_irae.top <- no_irae %>%
  group_by(microbe) %>%
    summarize(total.nonzero = sum(est.reads > 0),
            percent.nonzero = total.nonzero/n_distinct(sample),
            mean.counts = mean(est.reads)) %>%
  ungroup() %>%
  mutate(iraeGroup = "No irAE")
```


```{r}
top10 <- rbind(irae.top, no_irae.top) %>%
  filter(percent.nonzero > 0.15) %>%
  arrange(desc(mean.counts)) %>%
  slice_head(n=10)

box.in <- rbind(irae, no_irae) %>%
  filter(microbe %in% top10$microbe) %>%
  left_join(mic.meta) %>%
  left_join(demographics) %>%
  dplyr::rename(taxon = microbe) %>%
  left_join(phyl_labels) %>%
  mutate(Phylum = ifelse(Phylum %in% phyl.ra$Phylum, Phylum, "Other"),
         taxon = gsub("_", " ", taxon),
         irAE = ifelse(irAE == "Yes", "irAE", "No irAE")) 


colors <- c("#CC6677","#6699CC","#44AA99")
colors <- c("darkgreen", "cyan4", "salmon2")
```

```{r}
box.in %>%
  ggplot(aes(x = taxon, y=log(est.reads), fill = Phylum)) +
  geom_boxplot(width = 0.5) +
  geom_jitter() + 
  scale_fill_manual(values = colors) +
  facet_grid(~ irAE, scales = "free", space = "free") +
  theme_bw(base_size = 9) +
  theme(
    panel.spacing = unit(1, "lines"), 
    axis.text.x = element_text(angle = 45, hjust = 1),  
    strip.text = element_text(size = 9),
    legend.position = "none"
  ) +
  xlab("")

ggsave("../../figures/Fig3E-irAE-s0-baseline-boxplots.png", height = 3, width = 5)
```
# Figure 3F - baseline timepoint distance to irAE centroid vs no-irAE per person
Q: are the irae pts more variable?

dist to centroid using the corrected data - no sig
betadisper on dist matrix - permutation test *
pairwise distances show a difference 

```{r}
taxa_mat <- rsig.filt$Log.transformed
group <- rsig.filt$Response
```

# first lets get the dist to centroid - in the "feature space"
```{r}
centroids <- taxa_mat %>%
  mutate(Group = group) %>%
  group_by(Group) %>%
  summarise(across(everything(), mean), .groups = "drop")

euclidean_dist <- function(x, center) sqrt(sum((x - center)^2))

distances <- mapply(function(i, g) {
  center <- centroids %>% filter(Group == g) %>% select(-Group) %>% as.numeric()
  euclidean_dist(as.numeric(taxa_mat[i, ]), center)
}, i = seq_len(nrow(taxa_mat)), g = group)

df_centroid <- data.frame(
  Sample = rownames(taxa_mat),
  Group = group,
  DistanceToCentroid = distances
)

df_centroid |>
  ggplot(aes(x = Group, y = DistanceToCentroid, fill = Group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4) +
  theme_bw() +
  labs(x = "Group", y = "Distance to group centroid") +
  stat_compare_means(method = "wilcox.test") +
  theme(legend.position = "none")
```

# betadisper defines centroids in distance matrix space, not raw feature means

# Using uncorrected values get * with permutation test 
# wilcoxon not signifianct ~ is the distribution of distances the same?
```{r}
taxa_mat <- rsig.filt$Uncorrected
dist_mat <- dist(taxa_mat, method = "euclidean")

bd <- betadisper(dist_mat, group)

df_bd <- data.frame(
  Sample = rownames(taxa_mat),
  Group = group,
  DistanceToCentroid = bd$distances
)

# Does dispersion differ between the groups? 
# permutation test: shuffles group labels and computes distances many times to
# get the null distribution
anova_bd <- permutest(bd)
anova_bd

# extract p-value from permutation test
# p_val <- anova_bd$tab["Pr(>F)"][1]  # single row

ggplot(df_bd, aes(x = Group, y = DistanceToCentroid, fill = Group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4) +
  theme_bw() +
  labs(x = "Group", y = "Distance to group centroid (betadisper)") +
  stat_compare_means(method = "wilcox.test") +
  theme(legend.position = "none")
```

# Pairwise distances 

check that not counting the same distance twice 
null model ?? 
```{r}
dist_mat <- as.matrix(dist(taxa_mat, method = "euclidean"))

# convert distance matrix to long format
dist_long <- as.data.frame(dist_mat) %>%
  mutate(Sample1 = rownames(.)) %>%
  pivot_longer(cols = -Sample1, names_to = "Sample2", values_to = "Distance") %>%
  filter(Sample1 < Sample2)  # keep unique pairs

# add group info
dist_long <- dist_long %>%
  mutate(
    Group1 = group[match(Sample1, rownames(taxa_mat))],
    Group2 = group[match(Sample2, rownames(taxa_mat))]
  ) %>%
  filter(Group1 == Group2) %>%  # only keep within-group distances
  rename(Group = Group1)

wilcox.test(Distance ~ Group, data = dist_long)

ggplot(dist_long, aes(x = Group, y = Distance, fill = Group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  theme_bw() +
  labs(x = "Group", y = "Pairwise distances within group") +
  stat_compare_means(method = "wilcox.test") +
  theme(legend.position = "none")
```

No irAE definitely more tightly clustered!






<!-- Try Volcano colored by association with R and NR -->
<!-- ```{r} -->
<!-- load("../../data/ANCOM-BC2_response_baseline.RData") -->

<!-- r.sig <- res.recist %>% -->
<!--   mutate(taxon = gsub("_", " ", taxon)) %>% -->
<!--   filter(q_ResponseResponder < 0.05, -->
<!--          lfc_ResponseResponder > 0) -->

<!-- nr.sig <- res.recist %>% -->
<!--   filter(q_ResponseResponder < 0.05, -->
<!--          lfc_ResponseResponder < 0) -->

<!-- res.irae.recist <- res.irae %>% -->
<!--   mutate(res.assoc = ifelse(taxon %in% r.sig$taxon, "Associated with R", -->
<!--                             ifelse(taxon %in% nr.sig$taxon, "Associated with NR", NA)), -->
<!--          label = ifelse((taxon %in% r.sig$taxon | taxon %in% r.sig$taxon) & q_irAEYes < 0.05, taxon, NA), -->
<!--          irAE = ifelse(lfc_irAEYes > 0, "irAE", "No-irAE")) -->

<!-- res.irae.recist %>% -->
<!--   filter(q_irAEYes < 1) %>% -->
<!--   ggplot(aes(x=lfc_irAEYes, y=-log10(q_irAEYes), label = label)) +  -->
<!--   geom_point(aes(x=lfc_irAEYes, y=-log10(q_irAEYes), color = res.assoc)) +  -->
<!--   geom_text_repel(size = 2) + -->
<!--   scale_color_manual(values = c("orange", "green4"), na.value = "honeydew3") + -->
<!--   theme_bw(base_size = 9) + -->
<!--   xlab("log2foldchange") + -->
<!--   ylab("log10(p-value)") + -->
<!--   theme(legend.position =  "bottom") + # legend matches stacked bar  -->
<!--   guides(color = guide_legend(na.translate = FALSE))  -->

<!-- ggsave("../../figures/Fig3B-irae-baseline-volcano-experimental.png", height = 4, width = 7) -->

<!-- table(res.irae.recist$res.assoc, res.irae.recist$irAE) -->
<!-- ``` -->
