---
title: "Random Forest"
author: "Noah Lepola"
author: "Edited by CD"
date: "2025-12-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggplot2)
library(randomForest)
library(ROCR)
library(caTools)
library(PRROC)
library(Boruta)
```

Loading data
```{r}
source("rf-functions.R")

load("../../data/ra-long.rds")
load("../../data/clinical.rds")
load("../../data/blood.rds")
load("../../data/microbiome_metadata.rds")
load("../../data/phylum-labels.rds")
load("../../data/ANCOM-BC2_response_baseline.Rdata")
load("../../data/ANCOM-BC2_irae_baseline.RData") 
```

```{r}
seed_list <- c(12345:12369)
```

#q value model objects
```{r, warning=FALSE}
#Object to limit microbiome sample to one ones taken within 10 days of treatment
baseline <- mic.meta %>%
  group_by(`Patient Id`) %>%
  arrange(Days) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  filter(abs(Days) <= 10) %>%
  subset(`Patient Id` %in% blood$`Patient Id`)

phyl_labels <- phyl_labels %>%
  mutate(taxon = substr(microbe, 4, nchar(microbe)))

#creating object for use in biome model
base.micro <- ra.l %>%
  pivot_wider(names_from = microbe, values_from = RelAbun) %>%
  subset(sample %in% baseline$sample)

# CD note - how do we feel about these cut offs being different??
#note it is dropping bacteria and archaea becaise it isn't in phyl_labels
#making object for irAE where q < 0.01
irAE_q <- res.irae %>%
  left_join(phyl_labels) %>%
  dplyr::select(microbe, q_irAEYes)  %>%
  filter(q_irAEYes < 0.005) %>%
  replace_na(list( microbe = "k__Archaea"))

irae_list <- irAE_q$microbe %>%
  append("Patient Id", after = 0)

irae_list[[1]] <- "Patient Id"

#making object for Response where q < 0.01
Response_q <- res.recist %>%
  left_join(phyl_labels) %>%
  dplyr::select(microbe, q_ResponseResponder)  %>%
  filter(q_ResponseResponder < 0.01) %>%
  replace_na(list( microbe = "k__Archaea"))

response_list <- Response_q$microbe %>%
  append("Patient Id", after = 0)

response_list[[1]] <- "Patient Id"
```

# blood gene objects
```{r}
#limiting blood samples to ones also with microbiome sample and creating object
base.gene <- blood %>%
  subset(`Patient Id` %in% baseline$`Patient Id`) %>%
  dplyr::select(`Patient Id`:YWHAZ)

blood_list <- colnames(base.gene)
irae_blood_list <- c(irae_list, blood_list)
response_blood_list <- c(response_list, blood_list)
```

# clinical objects
```{r}
base.clinical <- demographics %>%
  dplyr::select(`Patient Id`, 
         Sex,
         Age,
         Treatment.bin,
         Stage,
         `Total.WBC.(K/µL)`,
         `Relative.lymphocyte.count.(%)`, 
         `Relative.eosinophil.count.(%)`, 
         `LDH.(U/L)`, 
         `BMI.(kg/m2)`)

base.clinical <- base.clinical %>% 
  mutate_at(c("Sex",
              "Treatment.bin", 
              "Stage"), as.factor)

clinical_list <- colnames(base.clinical)

irae_clinical_list <- c(irae_list, clinical_list)
blood_clinical_list <- c(blood_list, clinical_list)

response_clinical_list <- c(irae_list, clinical_list)
blood_clinical_list <- c(blood_list, clinical_list)

irae_blood_clinical_list <- c(irae_blood_list, clinical_list)
response_blood_clinical_list <- c(response_blood_list, clinical_list)
```

# Object with all factors and outcomes to pass to rf func
```{r}
outcomes <- demographics %>%
  select(`Patient Id`, Response, irAE)

in.df <- baseline %>%
  dplyr::select(sample, `Patient Id`) %>%
  left_join(base.micro) %>%
  left_join(base.clinical) %>%
  left_join(base.gene) %>%
  left_join(outcomes) %>%
  dplyr::select(-sample)
```

### irAE 

# Clinical model
```{r}
rf_clinical_irAE_list <- kTest(seed_list,
                               in.df,
                               "No",
                               "Yes",
                               "irAE",
                               clinical_list)

clinical_irae_aucprs <- grabVals(rf_clinical_irAE_list,
                            seed_list)

clinical_irae_aucprs[[1]]$Predictors <- "Clinical"
clinical_irae_aucprs[[2]]$Predictors <- "Clinical"
```

# Limited biome model
```{r}
rf_q005_irAE_list <- kTest(seed_list,
                           in.df,
                          "No",
                          "Yes",
                          "irAE",
                          irae_list)

biome_irae_aucprs <- grabVals(rf_q005_irAE_list,
                         seed_list)

biome_irae_aucprs[[1]]$Predictors <- "Microbiome"
biome_irae_aucprs[[2]]$Predictors <- "Microbiome"
```

# Just blood model
```{r}
rf_blood_irAE_list <- kTest(seed_list,
                            in.df,
                            "No",
                            "Yes",
                            "irAE",
                            blood_list)

blood_irae_aucprs <- grabVals(rf_blood_irAE_list,
                         seed_list)

blood_irae_aucprs[[1]]$Predictors <- "Blood"
blood_irae_aucprs[[2]]$Predictors <- "Blood"
```

# Biome and blood model
```{r}
rf_biome_blood_irAE_list <- kTest(seed_list,
                                  in.df,
                                  "No",
                                  "Yes",
                                  "irAE",
                                  irae_blood_list)

biome_and_blood_irae_aucprs <- grabVals(rf_biome_blood_irAE_list,
                                   seed_list)

biome_and_blood_irae_aucprs[[1]]$Predictors <- "Blood + Microbiome"
biome_and_blood_irae_aucprs[[2]]$Predictors <- "Blood + Microbiome"
```

# Blood and clinical model
```{r}
rf_blood_clinical_irAE_list <- kTest(seed_list,
                                     in.df,
                                     "No",
                                     "Yes",
                                     "irAE",
                                     blood_clinical_list)

blood_and_clinical_irae_aucprs <- grabVals(rf_blood_clinical_irAE_list,
                                   seed_list)

blood_and_clinical_irae_aucprs[[1]]$Predictors <- "Clinical + Blood"
blood_and_clinical_irae_aucprs[[2]]$Predictors <- "Clinical + Blood"
```

# Biome and clinical model
```{r}
rf_biome_clinical_irAE_list <- kTest(seed_list,
                                     in.df,
                                     "No",
                                     "Yes",
                                     "irAE",
                                     irae_clinical_list)

biome_and_clinical_irae_aucprs <- grabVals(rf_biome_clinical_irAE_list,
                                   seed_list)

biome_and_clinical_irae_aucprs[[1]]$Predictors <- "Clinical + Microbiome"
biome_and_clinical_irae_aucprs[[2]]$Predictors <- "Clinical + Microbiome"
```

# Biome blood clinical model
```{r}
rf_biome_blood_clinical_irAE_list <- kTest(seed_list,
                                           in.df,
                                           "No",
                                           "Yes",
                                           "irAE",
                                           irae_blood_clinical_list)

biome_blood_clinical_irae_aucprs <- grabVals(rf_biome_blood_clinical_irAE_list,
                                        seed_list)

biome_blood_clinical_irae_aucprs[[1]]$Predictors <- "Clinical + Blood + Microbiome"
biome_blood_clinical_irae_aucprs[[2]]$Predictors <- "Clinical + Blood + Microbiome"
```


### Response

# Clinical model
```{r}
rf_clinical_response_list <- kTest(seed_list,
                                   in.df,
                                   "Non-Responder",
                                   "Responder",
                                   "Response",
                                   clinical_list)

clinical_response_aucprs <- grabVals(rf_clinical_response_list,
                            seed_list)

clinical_response_aucprs[[1]]$Predictors <- "Clinical"
clinical_response_aucprs[[2]]$Predictors <- "Clinical"
```


# Limited biome model
```{r}
rf_q01_response_list <- kTest(seed_list,
                              in.df,
                              "Non-Responder",
                              "Responder",
                              "Response",
                              response_list)

biome_response_aucprs <- grabVals(rf_q01_response_list,
                         seed_list)

biome_response_aucprs[[1]]$Predictors <- "Microbiome"
biome_response_aucprs[[2]]$Predictors <- "Microbiome"
```

# Just blood model
```{r}
rf_blood_response_list <- kTest(seed_list,
                                in.df,
                                "Non-Responder",
                                "Responder",
                                "Response",
                                blood_list)

blood_response_aucprs <- grabVals(rf_blood_response_list,
                         seed_list)

blood_response_aucprs[[1]]$Predictors <- "Blood"
blood_response_aucprs[[2]]$Predictors <- "Blood"
```

# Biome and blood model
```{r}
rf_biome_blood_response_list <- kTest(seed_list,
                                      in.df,
                                      "Non-Responder",
                                      "Responder",
                                      "Response",
                                      response_blood_list)

biome_and_blood_response_aucprs <- grabVals(rf_biome_blood_response_list,
                                   seed_list)

biome_and_blood_response_aucprs[[1]]$Predictors <- "Blood + Microbiome"
biome_and_blood_response_aucprs[[2]]$Predictors <- "Blood + Microbiome"
```

# Biome clinical model
```{r}
rf_biome_clinical_response_list <- kTest(seed_list,
                                         in.df,
                                         "Non-Responder",
                                         "Responder",
                                         "Response",
                                         response_clinical_list)

biome_clinical_response_aucprs <- grabVals(rf_biome_clinical_response_list,
                                   seed_list)

biome_clinical_response_aucprs[[1]]$Predictors <- "Clinical + Microbiome"
biome_clinical_response_aucprs[[2]]$Predictors <- "Clinical + Microbiome"
```

# Blood clinical model
```{r}
rf_blood_clinical_response_list <- kTest(seed_list,
                                         in.df,
                                         "Non-Responder",
                                         "Responder",
                                         "Response",
                                          blood_clinical_list)

blood_clinical_response_aucprs <- grabVals(rf_blood_clinical_response_list,
                                   seed_list)

blood_clinical_response_aucprs[[1]]$Predictors <- "Clinical + Blood"
blood_clinical_response_aucprs[[2]]$Predictors <- "Clinical + Blood"
```

# Biome blood clinical model
```{r}
rf_biome_blood_clinical_response_list <- kTest(seed_list,
                                               in.df,
                                               "Non-Responder",
                                               "Responder",
                                               "Response",
                                               response_blood_clinical_list)

biome_blood_clinical_response_aucprs <- grabVals(rf_biome_blood_clinical_response_list,
                                   seed_list)

biome_blood_clinical_response_aucprs[[1]]$Predictors <- "Clinical + Blood + Microbiome"
biome_blood_clinical_response_aucprs[[2]]$Predictors <- "Clinical + Blood + Microbiome"
```


### Combine and save as one df for irAE
```{r}
# Combine ROC (AUC) for irAE
roc_irae_df <- bind_rows(
  clinical_irae_aucprs[[1]],
  biome_irae_aucprs[[1]],
  blood_irae_aucprs[[1]],
  biome_and_blood_irae_aucprs[[1]],
  blood_and_clinical_irae_aucprs[[1]],
  biome_and_clinical_irae_aucprs[[1]],
  biome_blood_clinical_irae_aucprs[[1]]
)

# Combine PR (AUCPR) for irAE
pr_irae_df <- bind_rows(
  clinical_irae_aucprs[[2]],
  biome_irae_aucprs[[2]],
  blood_irae_aucprs[[2]],
  biome_and_blood_irae_aucprs[[2]],
  blood_and_clinical_irae_aucprs[[2]],
  biome_and_clinical_irae_aucprs[[2]],
  biome_blood_clinical_irae_aucprs[[2]]
)
```

### Combine and save as one df for Response
```{r}
# Combine ROC (AUC) for Response
roc_response_df <- bind_rows(
  clinical_response_aucprs[[1]],
  biome_response_aucprs[[1]],
  blood_response_aucprs[[1]],
  biome_and_blood_response_aucprs[[1]],
  biome_clinical_response_aucprs[[1]],
  blood_clinical_response_aucprs[[1]],
  biome_blood_clinical_response_aucprs[[1]]
)

# Combine PR (AUCPR) for Response
pr_response_df <- bind_rows(
  clinical_response_aucprs[[2]],
  biome_response_aucprs[[2]],
  blood_response_aucprs[[2]],
  biome_and_blood_response_aucprs[[2]],
  biome_clinical_response_aucprs[[2]],
  blood_clinical_response_aucprs[[2]],
  biome_blood_clinical_response_aucprs[[2]]
)
```

# Get top vars 
```{r}
# get vars from all with AUC > 0.8 from microbiome input
r.vars <- get_vars_auc_threshold(rf_q01_response_list)
i.vars <- get_vars_auc_threshold(rf_q005_irAE_list)

r.vars
i.vars

intersect(r.vars, i.vars)
```

# save objects
```{r}
save(roc_irae_df, pr_irae_df, i.vars, file = "../../data/randomforest-irae.rda")
save(roc_response_df, pr_response_df, r.vars, file = "../../data/randomforest-response.rda")
```

Examine averages
```{r}
pr_irae_df %>%
  select(AUCPR, Predictors, seed) %>%
  distinct() %>%
  group_by(Predictors) %>%
  summarise(avg.aucpr = mean(AUCPR)) %>%
  arrange(avg.aucpr)

pr_response_df %>%
  select(AUCPR, Predictors, seed) %>%
  distinct() %>%
  group_by(Predictors) %>%
  summarise(avg.aucpr = mean(AUCPR)) %>%
  arrange(avg.aucpr)

roc_irae_df %>%
  select(AUCROC, Predictors, seed) %>%
  distinct() %>%
  group_by(Predictors) %>%
  summarise(avg.auc = mean(AUCROC)) %>%
  arrange(avg.auc)

roc_response_df %>%
  select(AUCROC, Predictors, seed) %>%
  distinct() %>%
  group_by(Predictors) %>%
  summarise(avg.auc = mean(AUCROC)) %>%
  arrange(avg.auc)
```
