---
title: "Random Forest"
author: "Noah Lepola"
author: "Edited by CD"
date: "2025-12-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggplot2)
library(randomForest)
library(ROCR)
library(caTools)
library(PRROC)
library(Boruta)
```

Loading data
```{r}
source("rf-functions.R")

load("../../data/ra-long.rds")
load("../../data/clinical.rds")
load("../../data/blood.rds")
load("../../data/microbiome_metadata.rds")
load("../../data/phylum-labels.rds")
load("../../data/ANCOM-BC2_response_baseline.Rdata")
load("../../data/ANCOM-BC2_irae_baseline.RData") 
```

```{r}
seed_list <- c(12345:12369)
```

#q value model objects
```{r, warning=FALSE}
#Object to limit microbiome sample to one ones taken within 10 days of treatment
baseline <- mic.meta %>%
  group_by(`Patient Id`) %>%
  arrange(Days) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  filter(abs(Days) <= 10) %>%
  subset(`Patient Id` %in% blood$`Patient Id`)

phyl_labels <- phyl_labels %>%
  mutate(taxon = substr(microbe, 4, nchar(microbe)))

#creating object for use in biome model
base.micro <- ra.l %>%
  pivot_wider(names_from = microbe, values_from = RelAbun) %>%
  subset(sample %in% baseline$sample)

full_micro_list <- colnames(base.micro)
full_micro_list[[1]] <- "Patient Id"

# CD note - how do we feel about these cut offs being different??
#note it is dropping bacteria and archaea becaise it isn't in phyl_labels
#making object for irAE where q < 0.01
irAE_q <- res.irae %>%
  left_join(phyl_labels) %>%
  dplyr::select(microbe, q_irAEYes)  %>%
  filter(q_irAEYes < 0.005) %>%
  replace_na(list( microbe = "k__Archaea"))

irae_list <- irAE_q$microbe %>%
  append("Patient Id", after = 0)

irae_list[[1]] <- "Patient Id"

#making object for Response where q < 0.01
Response_q <- res.recist %>%
  left_join(phyl_labels) %>%
  dplyr::select(microbe, q_ResponseResponder)  %>%
  filter(q_ResponseResponder < 0.01) %>%
  replace_na(list( microbe = "k__Archaea"))

response_list <- Response_q$microbe %>%
  append("Patient Id", after = 0)

response_list[[1]] <- "Patient Id"
```

# blood gene objects
```{r}
#limiting blood samples to ones also with microbiome sample and creating object
base.gene <- blood %>%
  subset(`Patient Id` %in% baseline$`Patient Id`) %>%
  dplyr::select(`Patient Id`:YWHAZ)

blood_list <- colnames(base.gene)
irae_blood_list <- c(irae_list, blood_list)
response_blood_list <- c(response_list, blood_list)
```

# clinical objects
```{r}
base.clinical <- demographics %>%
  dplyr::select(`Patient Id`, 
         Sex,
         Age,
         Treatment.bin,
         Stage,
         `Total.WBC.(K/µL)`,
         `Relative.lymphocyte.count.(%)`, 
         `Relative.eosinophil.count.(%)`, 
         `LDH.(U/L)`, 
         `BMI.(kg/m2)`)

base.clinical <- base.clinical %>% 
  mutate_at(c("Sex",
              "Treatment.bin", 
              "Stage"), as.factor)

clinical_list <- colnames(base.clinical)

irae_clinical_list <- c(irae_list, clinical_list)
blood_clinical_list <- c(blood_list, clinical_list)

response_clinical_list <- c(irae_list, clinical_list)
blood_clinical_list <- c(blood_list, clinical_list)

irae_blood_clinical_list <- c(irae_blood_list, clinical_list)
response_blood_clinical_list <- c(response_blood_list, clinical_list)
```

# Object with all factors
```{r}
all_factors <- baseline %>%
  dplyr::select(sample, `Patient Id`) %>%
  full_join(base.micro, baseline, by = "sample") %>%
  full_join(base.gene, baseline, by = "Patient Id") %>%
  left_join(base.clinical, baseline, by = "Patient Id") %>%
  dplyr::select(-sample)
```


### irAE 

# Clinical model
```{r}
rf_clinical_irAE_list <- kTest(seed_list,
                               "No",
                               "Yes",
                               "irAE",
                               clinical_list,
                               FALSE,
                               1000)

clinical_irae_aucprs <- grabVals(rf_clinical_irAE_list,
                            seed_list)
clinical_irae_aucprs$Predictors <- "Clinical"
```


# Limited biome model
```{r}
rf_q005_irAE_list <- kTest(seed_list,
                          "No",
                          "Yes",
                          "irAE",
                          irae_list,
                          FALSE,
                          1000)

biome_irae_aucprs <- grabVals(rf_q005_irAE_list,
                         seed_list)
biome_irae_aucprs$Predictors <- "Microbiome"
```

# Just blood model
```{r}
rf_blood_irAE_list <- kTest(seed_list,
                            "No",
                            "Yes",
                            "irAE",
                            blood_list,
                            FALSE,
                            1000)

blood_irae_aucprs <- grabVals(rf_blood_irAE_list,
                         seed_list)
blood_irae_aucprs$Predictors <- "Blood"
```

# Biome and blood model
```{r}
rf_biome_blood_irAE_list <- kTest(seed_list,
                                  "No",
                                  "Yes",
                                  "irAE",
                                  irae_blood_list,
                                  FALSE,
                                  1000)

biome_and_blood_irae_aucprs <- grabVals(rf_biome_blood_irAE_list,
                                   seed_list)
biome_and_blood_irae_aucprs$Predictors <- "Blood + Microbiome"
```

# Blood and clinical model
```{r}
rf_blood_clinical_irAE_list <- kTest(seed_list,
                                  "No",
                                  "Yes",
                                  "irAE",
                                  blood_clinical_list,
                                  FALSE,
                                  1000)

blood_and_clinical_irae_aucprs <- grabVals(rf_blood_clinical_irAE_list,
                                   seed_list)
blood_and_clinical_irae_aucprs$Predictors <- "Clinical + Blood"
```

# Biome and clinical model
```{r}
rf_biome_clinical_irAE_list <- kTest(seed_list,
                                  "No",
                                  "Yes",
                                  "irAE",
                                  irae_clinical_list,
                                  FALSE,
                                  1000)

biome_and_clinical_irae_aucprs <- grabVals(rf_biome_clinical_irAE_list,
                                   seed_list)
biome_and_clinical_irae_aucprs$Predictors <- "Clinical + Microbiome"
```

# Biome blood clinical model
```{r}
rf_biome_blood_clinical_irAE_list <- kTest(seed_list,
                                           "No",
                                           "Yes",
                                           "irAE",
                                           irae_blood_clinical_list,
                                           FALSE,
                                           1000)

biome_blood_clinical_irae_aucprs <- grabVals(rf_biome_blood_clinical_irAE_list,
                                        seed_list)
biome_blood_clinical_irae_aucprs$Predictors <- "Clinical + Blood + Microbiome"
```


### Response

# Clinical model
```{r}
rf_clinical_response_list <- kTest(seed_list,
                                "Non-Responder",
                                "Responder",
                                "Response",
                                clinical_list,
                                FALSE,
                                1000)

clinical_response_aucprs <- grabVals(rf_clinical_response_list,
                            seed_list)
clinical_response_aucprs$Predictors <- "Clinical"
```


# Limited biome model
```{r}
rf_q01_response_list <- kTest(seed_list,
                              "Non-Responder",
                              "Responder",
                              "Response",
                              response_list,
                              FALSE,
                              1000)

biome_response_aucprs <- grabVals(rf_q01_response_list,
                         seed_list)
biome_response_aucprs$Predictors <- "Microbiome"
```

# Just blood model
```{r}
rf_blood_response_list <- kTest(seed_list,
                                "Non-Responder",
                                "Responder",
                                "Response",
                                blood_list,
                                FALSE,
                                1000)

blood_response_aucprs <- grabVals(rf_blood_response_list,
                         seed_list)
blood_response_aucprs$Predictors <- "Blood"
```

# Biome and blood model
```{r}
rf_biome_blood_response_list <- kTest(seed_list,
                                  "Non-Responder",
                                  "Responder",
                                  "Response",
                                  response_blood_list,
                                  FALSE,
                                  1000)

biome_and_blood_response_aucprs <- grabVals(rf_biome_blood_response_list,
                                   seed_list)
biome_and_blood_response_aucprs$Predictors <- "Blood + Microbiome"
```

# Biome clinical model
```{r}
rf_biome_clinical_response_list <- kTest(seed_list,
                                          "Non-Responder",
                                          "Responder",
                                          "Response",
                                          response_clinical_list,
                                          FALSE,
                                          1000)

biome_clinical_response_aucprs <- grabVals(rf_biome_clinical_response_list,
                                   seed_list)
biome_clinical_response_aucprs$Predictors <- "Clinical + Microbiome"
```

# Blood clinical model
```{r}
rf_blood_clinical_response_list <- kTest(seed_list,
                                          "Non-Responder",
                                          "Responder",
                                          "Response",
                                          blood_clinical_list,
                                          FALSE,
                                          1000)

blood_clinical_response_aucprs <- grabVals(rf_blood_clinical_response_list,
                                   seed_list)
blood_clinical_response_aucprs$Predictors <- "Clinical + Blood"
```

# Biome blood clinical model
```{r}
rf_biome_blood_clinical_response_list <- kTest(seed_list,
                                          "Non-Responder",
                                          "Responder",
                                          "Response",
                                          response_blood_clinical_list,
                                          FALSE,
                                          1000)

biome_blood_clinical_response_aucprs <- grabVals(rf_biome_blood_clinical_response_list,
                                   seed_list)
biome_blood_clinical_response_aucprs$Predictors <- "Clinical + Blood + Microbiome"
```


### Combine and save as one df for irAE
```{r}
irae.rf <- rbind(clinical_irae_aucprs, blood_irae_aucprs, biome_irae_aucprs, biome_and_blood_irae_aucprs, blood_and_clinical_irae_aucprs, biome_and_clinical_irae_aucprs, biome_blood_clinical_irae_aucprs)

save(irae.rf, file = "../data/randomforest-irae-boruta.rds")
```

### Combine and save as one df for Response
```{r}
response.rf <- rbind(clinical_response_aucprs, blood_response_aucprs, biome_response_aucprs, biome_clinical_response_aucprs, blood_clinical_response_aucprs,  biome_and_blood_response_aucprs, biome_blood_clinical_response_aucprs)

save(response.rf, file = "../data/randomforest-response-boruta.rds")
```

## Get variables used 
```{r}
# get vars from all with AUC > 0.8 from microbiome input
r.vars <- get_vars_auc_threshold(rf_q01_response_list)
i.vars <- get_vars_auc_threshold(rf_q005_irAE_list)

r.vars
i.vars

intersect(r.vars, i.vars)

saveRDS(r.vars, "../data/r-vars.rds")
saveRDS(i.vars, "../data/i-vars.rds")
```
