---
title: "table 1"
author: "Caroline Wheeler"
date: "2023-10-20"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(table1)
library(tableone)
library(gridExtra)
library(htmltools)
```

```{r}
load("../../data/clinical.rds")
load("../../data/irae.rds")
load("../../data/microbiome_metadata.rds")
```

Remove participant without baseline sample 
```{r}
baseline <- mic.meta %>%
  group_by(`Patient Id`) %>%
  arrange(Days) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  filter(abs(Days) <= 10)

demo.filt <- demographics %>%
  filter(`Patient Id` %in% baseline$`Patient Id`) %>%
  mutate(Treatment = Treatment.bin)
```


function to get p-value 
```{r}
pvalue <- function(x, ...) {
  x <- x[-length(x)]  # Remove "overall" group
  # Construct vectors of data y, and groups (strata) g
  y <- unlist(x)
  g <- factor(rep(1:length(x), times=sapply(x, length)))
  if (is.numeric(y)) {
    # For numeric variables, perform an ANOVA
    p <- summary(aov(y ~ g))[[1]][["Pr(>F)"]][1]
  } else {
    # For categorical variables, perform a chi-squared test of independence
    p <- chisq.test(table(y, g))$p.value
  }
  # Format the p-value, using an HTML entity for the less-than sign.
  # The initial empty string places the output on the line below the variable 
  #label.
  c("", sub("<", "&lt;", format.pval(p, digits=3, eps=0.001)))
}
```

Table 1 - strat by Response
```{r}
table1 <- table1(~ Age + Sex + Race + Treatment + Stage + irAE  | Response,
       data=demo.filt, overall="Total", extra.col=list(`P-value`=pvalue), 
       render.missing=NULL, render.continuous=c(.="Mean (SD)"))

print(table1, quote = FALSE, noSpaces = TRUE, printToggle = TRUE)

html_file <- "../../figures/table1.html"
save_html(table1, file = html_file)
```



If we decide to make our own version 
```{r}
tab1 <- CreateTableOne(
  vars = c("Age", "Sex", "Race", "Treatment", "Stage", "irAE"),  # Variables
  strata = "Response",  # Stratification variable
  data = demo.filt,
  test = TRUE,  # Enables default p-value calculation
  addOverall = TRUE  # Add an "Overall" column
)

print(tab1)
```

Table 2 - strat by irAE
```{r}
demo.filt <- demo.filt %>%
    mutate(irAE = ifelse(irAE == "Yes", "irAE", "No irAE"))

table2 <- table1(~ Age + Sex + Race + Treatment + Stage + Response  | irAE,
       data=demo.filt, overall="Total", extra.col=list(`P-value`=pvalue), 
       render.missing=NULL, render.continuous=c(.="Mean (SD)"))

print(table2, quote = FALSE, noSpaces = TRUE, printToggle = TRUE)

html_file <- "../../figures/table2.html"
save_html(table2, file = html_file)
```


```{r}
tab2 <- CreateTableOne(
  vars = c("Age", "Sex", "Race", "Treatment", "Stage", "Response"),  # Variables
  strata = "irAE",  # Stratification variable
  data = demo.filt,
  test = TRUE,  # Enables default p-value calculation
  addOverall = TRUE  # Add an "Overall" column
)

print(tab2)
```

Table 3 - Summary of the immune-related adverse event types and grades

"Common irAEs include skin toxicity, colitis, hepatitis, pneumonia, nephritis, and endocrine disorders. Rare but severe irAEs include encephalitis, myocarditis, and haematological toxicity"

Adrenal insufficiency  Autoimmune hepatitis              Delirium              Diarrhea 
                    3                     2                     1                     1 
      Hyperthyroidism        Hypothyroidism    Maculo-paular rash       Psoriasis flare 
                    1                     2                     1                     1 
                 Rash          Skin itching           Thyroiditis 
                    5                     1                     1 
                    
```{r}
irae.in <- irae %>%
  filter(`Patient Id` %in% (demo.filt %>% filter(irAE == "Yes"))$`Patient Id`) %>%
  mutate(Grade = as.factor(`CTCAE grade (final)`),
         Type = ifelse(Symptoms %in% c("Rash", "Skin itching", "Maculo-paular rash", "Psoriasis flare"), 
                       "Skin Toxicity", 
                       ifelse(Symptoms %in% c("Adrenal insufficiency", "Hypothyroidism", "Hyperthyroidism", "Thyroiditis"), "Endocrine Disorder",
                       ifelse(Symptoms == "Autoimmune hepatitis", "Autoimmune Hepatitis", Symptoms))))
```

```{r}
tab3 <- CreateTableOne(
  vars = c("Type", "Grade"),  # Variables
  data = irae.in,
)

print(tab3)

# Convert TableOne to a data frame
tab3_df <- as.data.frame(print(tab3, quote = FALSE, noSpaces = TRUE))

# Save as CSV
write.csv(tab3_df, "../../figures/supplemental-table1.csv", row.names = TRUE)
```


```{r}
table3 <- table1(~ Type + Grade, data = irae.in)
  
  table1(Type + Grade,
       data=irae.in,  extra.col=list(`P-value`=pvalue), 
       render.missing=NULL, render.continuous=c(.="Mean (SD)"))

table3

print(table3, quote = FALSE, noSpaces = TRUE, printToggle = TRUE)

html_file <- "../../figures/table3.html"
save_html(table3, file = html_file)
```


