---
title: "check metaphlan ouput"
author: "Caroline Wheeler"
date: "9/9/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

```{r}
# fastqs
fastq.dir1 <- "/fs/ess/PAS1695/projects/mitox/data/2022-01-31_fastqs"
fastq.dir2 <- "/fs/ess/PAS1695/projects/mitox/data/2022-08-08_fastqs"

all.files1 <- list.files(fastq.dir1, full.names = T)
all.files2 <- list.files(fastq.dir2, full.names = T)
all.files <- c(all.files1, all.files2)
fnames <- gsub(".fastq.gz", "", c(list.files(fastq.dir1), list.files(fastq.dir2)))

# metaphlan output 
met.dir <- "/fs/ess/PAS1695/projects/mitox/data/2024-08-30-all-fastqs_metaphlan"
met.files <- list.files(met.dir, full.names = T)
mnames <- gsub(".txt", "", list.files(met.dir, full.names = F))
```

```{r}
all.equal(fnames, mnames)
```

## combine metaphlan ouput
```{r}
mettab.ls <- lapply(met.files, function(x) read.table(paste0(x),
                                                            header = F, sep = "\t", stringsAsFactors = F))

names(mettab.ls) <- mnames

mettab.ls <- lapply(mettab.ls, function(x) mutate(x,
           V2 = as.character(V2)))


metab.df <- bind_rows(lapply(mnames, function(x) mettab.ls[[x]] %>% mutate(sample = x)))

metab.df.form <- metab.df %>%
  rename("Taxonomy" = "V1",
         "TaxNum" = "V2",
         "RelAbun" = "V3",
         "Alternative.Tax" = "V4") %>%
  filter(Taxonomy != "UNCLASSIFIED")
```

save combined taxonomic profiling 
```{r}
write.csv(metab.df.form, "/fs/ess/PAS1695/projects/mitox/data/2024-08-30_mpa_aggregated_long.csv", row.names = F)
```

## combine virus output
```{r}
# vsc output 
vir.dir <- "/fs/ess/PAS1695/projects/mitox/data/2024-08-30-all-fastqs_vsc"
vir.files <- list.files(vir.dir, full.names = T)
vnames <- gsub(".vsc.txt", "", list.files(vir.dir, full.names = F))

all.equal(vnames, fnames)
```

```{r}
virtab.ls <- lapply(vir.files, function(x) read.table(paste0(x),
                                                            header = T, sep = "\t", stringsAsFactors = F))

names(virtab.ls) <- vnames

virtab.ls <- lapply(virtab.ls, function(x) mutate(x, M.Group.Cluster = as.character(M.Group.Cluster)))
virtab.ls <- lapply(virtab.ls, function(x) mutate(x, genomeName = as.character(genomeName)))
virtab.ls <- lapply(virtab.ls, function(x) mutate(x, M.Group.Type..k.u. = as.character(M.Group.Type..k.u.)))
virtab.ls <- lapply(virtab.ls, function(x) mutate(x, First.Genome.in.Cluster = as.character(First.Genome.in.Cluster)))
virtab.ls <- lapply(virtab.ls, function(x) mutate(x, Other.Genomes = as.character(Other.Genomes)))

virtab.df <- bind_rows(lapply(vnames, function(x) virtab.ls[[x]] %>% mutate(sample = x)))
```

save combined virus output
```{r}
write.csv(virtab.df, "/fs/ess/PAS1695/projects/mitox/data/2024-08-30_vsc_aggregated_long.csv", row.names = F)
```

## combine estimated count output
```{r}
# count output 
c.dir <- "/fs/ess/PAS1695/projects/mitox/data/2024-08-30-all-fastqs_metaphlan-counts"
c.files <- list.files(c.dir, full.names = T)
cnames <- gsub(".tsv", "", list.files(c.dir, full.names = F))

all.equal(cnames, fnames)
```

```{r}
ctab.ls <- lapply(c.files, function(x) read.table(paste0(x),
                                                            header = F, sep = "\t", stringsAsFactors = F))

names(ctab.ls) <- cnames

ctab.ls <- lapply(ctab.ls, function(x) mutate(x,
           V2 = as.character(V2)))


ctab.df <- bind_rows(lapply(cnames, function(x) ctab.ls[[x]] %>% mutate(sample = x)))

ctab.df.form <- cab.df %>%
  rename("Taxonomy" = "V1",
         "TaxNum" = "V2",
         "RelAbun" = "V3",
         "coverage" = "V4",
         "est.reads" = "V5") %>%
  filter(Taxonomy != "unclassified")
```

est.read dervied via...
In brief, this is estimated by normalizing the observed average coverage of the marker genes with respect to the genome size. So if we have 100 reads mapping against the markers of a species A and if the markers of species A are covering 10% of the genome, the estimated number of reads coming from the target genome is 1000. 

check equalities
```{r}
all.equal(ctab.df.form$Taxonomy, metab.df.form$Taxonomy)
all.equal(ctab.df.form$TaxNum, metab.df.form$TaxNum)
all.equal(ctab.df.form$RelAbun, metab.df.form$RelAbun)
```

so we can just use this one table for RA and counts. Note: these numbers will not directly correspond 
due to metaphlan methods of estimatation. RA centric..

save 
```{r}
write.csv(ctab.df.form, "/fs/ess/PAS1695/projects/mitox/data/2024-08-30_mpa-counts_aggregated_long.csv", row.names = F)
```

