---
title: "generate data objects for publication"
author: "Caroline Wheeler"
date: "2023-10-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readxl)

# source("00-paths.R")
```

Publication objects = cleaned R objects in the repo, without PHI, that are the inputs to ALL of the figures

file path 
```{r}
paths.tdrive <- "/Volumes/spakowicz/"
#paths.tdrive <- "T:/Labs/Spakowicz/"
```


read in necessary data
```{r}
# pt_log <- readxl::read_xlsx(file.path(paths.tdrive,"mitox/Patient Logs/OSU 19125 Patient Log.xlsx"), skip = 3) %>%
#   dplyr::select(-`Pt Name`, -MRN)
# 
irAE <- readxl::read_xlsx(file.path(paths.tdrive,"mitox/Clinical info/Mitoxstudy_patient_symptoms_AM_2.xlsx")) %>%
  dplyr::select(-`Patient Name`, - MRN)

# recist, CT scans
recist.scan <- readxl::read_xlsx(file.path(paths.tdrive, "mitox/Clinical info/OSU-19125_patient_progression_RECIST_KK_v4.xlsx")) %>%
  dplyr::select(-Name, -MRN)

micSamp <- read.table(file.path(paths.tdrive, 'mitox/data/labii/mitox-samples-export_2022-10-21.tsv'), sep = '\t', header = TRUE, fill = NA)
# 
# blood.adj <- openxlsx::readWorkbook(xlsxFile =  file.path(paths.tdrive, "mitox/Clinical info/Patient Data 7-1-22.xlsm"), sheet = "Adj Datasheet", startRow = 2)
# blood.met <- openxlsx::readWorkbook(xlsxFile =  file.path(paths.tdrive, "mitox/Clinical info/Patient Data 7-1-22.xlsm"), sheet = "Met Datasheet", startRow = 2)
# blood.metj <- openxlsx::readWorkbook(xlsxFile =  file.path(paths.tdrive, "mitox/Clinical info/Patient Data 7-1-22.xlsm"), sheet = "Jason Met Cohort", startRow = 2)
# 
demo.dat <- read_excel(file.path(paths.tdrive, "mitox/Patient Logs/Demographics/Patient_demographics.xlsx")) %>%
   select(-MRN, -Name)
# 
# ae.stat <- read.csv(file.path(paths.tdrive, "mitox/data/table_patients-with-AE_baseline-samps.csv" )) %>%
#   select(-mrn)


# PFS
pfs <- readxl::read_xlsx(file.path(paths.tdrive, "mitox/Clinical info/OSU-19125_patient_progression_RECIST_KK_v4.xlsx"),
                               sheet = "Progression dates")

# microbiome taxonomic data merged metaphlan output 
taxa <- read.csv("../../../data/mpa_aggregated_long_10-5-22.csv")

# microbiome sample matching 
samps <- read_xlsx( file.path(paths$oneDrive, "/data/onecodex/2021-10-27_mitox_One Codex Sample Registration.xlsx"))
samps2 <- read_xlsx( file.path(paths$oneDrive, "/data/onecodex/2022-07-27_mitox_remaining_One Codex Sample Registration.xlsx")) %>%
   select(-`Specimen Type`)
samps <- rbind(samps, samps2) 
```

Demographics (Input to Tables 1, S1)
Rows are pts
```{r}
# clean up demographic data
demographics <- demo.dat %>% 
  filter(!is.na(`Patient Id`)) %>%
  select(-c(No., Notes))

# get irAE, symptoms, response, pfs, microbiome sample presence
recist.mic <- recist.scan %>%
  select(`Patient ID`, RECIST, Category, Samples) %>%
  rename("Patient Id" = `Patient ID`,
         "Mic.Samp" = Samples) %>%
  filter(!is.na(`Patient Id`))

pfs <- pfs %>%
  select(`Patient Id`, `12-month PFS`)

irae <- irAE %>%
  dplyr::select(`Patient Id`, `Attribution (Final)`) %>%
  rename(irAE = `Attribution (Final)`) %>%
  filter(irAE == "Yes") %>%
  distinct()

# join 
demographics <- merge(demographics, recist.mic, all=T)
demographics <- merge(demographics, pfs, all=T)
demographics <- merge(demographics, irae, all=T)

# slight mutation to irae
demographics <- demographics %>%
  mutate(irAE = case_when(is.na(irAE)~ "No", .default = irAE))

save(demographics, file = "../data/clinical.rds")
```


Events object (input to Figure 1C (timeline figure))
Rows are events
ptID, event, time from C1D1 
```{r}
# get average number of days between date collected and date processed 
x <- micSamp %>%
  filter(sample_id != "",
         Type.of.Sample != "") %>%
  mutate(test = as.integer(as.Date(date_processed) - as.Date(Date.Sample.Collected))) %>%
  filter(!is.na(test))

x <- mean(x$test)


```


irAE table
Rows are irAEs
ptID, grade, type (input to Table S2)
```{r}
irae <- irAE %>%
  dplyr::select(`Patient Id`, Symptoms, `CTCAE grade (final)`, Onset, `Symptom Treatment`, `Attribution (Final)`) %>%
  rename(irAE = `Attribution (Final)`) %>%
  filter(!is.na(`Patient Id`),
         irAE == "Yes") 

save(irae, file = "../../data/irae.rds")
```


Microbiome sample metadata
Rows are microbiome samples
ptID, category label [baseline, 12wk, irae, control], link to 
```{r}
temp.mic <- micSamp %>%
  filter(sample_id != "",
         Type.of.Sample != "") %>%
  select(sample_id, Type.of.Sample, record_id) %>%
  # use sample_id substr instead of record_id, which is missing values
  mutate(`Patient Id` = ifelse(record_id != "", record_id, substr(sample_id, 0, 4)),
         Type.of.Sample = ifelse(Type.of.Sample == "12Wk", "12wk", Type.of.Sample),
         #sample_id = substr(sample_id, 0, nchar(sample_id)-2),
         sample_id = gsub(" ", ".", sample_id)) %>%
  select(-record_id) %>%
  rename("Category" = Type.of.Sample) %>% 
  distinct() 

samps <- samps %>%
  select("Barcode ID", "Name") %>%
  mutate(sample_id = gsub("_", ".", Name)) %>%
  mutate(sample_id = gsub(" ", ".", sample_id)) %>%
  select(-Name) %>%
  rename("sample" = "Barcode ID")

test <- merge(samps, temp.mic, all=T)
mic.meta <- left_join(samps, temp.mic)

save(mic.meta, file = "../../data/microbiome_metadata.rds")
```
ISSUE - sid = ruxA0a40xe3f4tyDINSX3  has record_id = M020, but the sample_id = M011.200911.2
RESOLVED it is M020

ISSUE M048 has a 12wk microbiome sample but no collection date 
RESOLVED average days between collection and processing and add collection dates based on processing date 

Microbiome data
Abundances 
Rows are microbes
```{r}
mic <- taxa %>%
  select(sample, Taxonomy, RelAbun) %>%
  pivot_wider(names_from = sample, values_from = RelAbun, values_fill = 0) 

save(mic, file = "../../data/microbiome.rds")
```

Taxonomy table
```{r}
tax.tab <- taxa %>%
   dplyr::select(sample, Taxonomy, RelAbun) %>%
    mutate(kingdom = gsub("(k__\\w+).*", "\\1", Taxonomy),
           phylum = ifelse(grepl("p__", Taxonomy),gsub(".*(p__\\w+).*", "\\1", Taxonomy), NA),
           order = ifelse(grepl("o__", Taxonomy),gsub(".*(o__\\w+).*", "\\1", Taxonomy), NA),
           class = ifelse(grepl("c__", Taxonomy),gsub(".*(c__\\w+).*", "\\1", Taxonomy), NA),
           family = ifelse(grepl("f__", Taxonomy),gsub(".*(f__\\w+).*", "\\1", Taxonomy), NA),
           genus = ifelse(grepl("g__", Taxonomy),gsub(".*(g__\\w+).*", "\\1", Taxonomy), NA),
           species = ifelse(grepl("s__", Taxonomy), gsub(".*(s__.*)", "\\1", Taxonomy), NA)) %>%
  filter(!is.na(species)) 

save(tax.tab, file = "../../data/taxonomy.rds")
```

Blood data
Rows are blood samples
ptID, genes
```{r}

```

