---
title: "generate data objects for publication"
author: "Caroline Wheeler"
date: "2023-10-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readxl)

source("../../../00-paths.R")
```

Publication objects = cleaned R objects in the repo, without PHI, that are the 
inputs to ALL  figures

need to remove problem sample - FF08034522
it is in the adj cohort, so no problem here.

Should we remove M032 since we don't have response information and no irAE?
  Will leave in at this stage. 

<!-- file path  -->
<!-- ```{r} -->
<!-- paths.tdrive <- "/Volumes/spakowicz/" -->
<!-- #paths.tdrive <- "T:/Labs/Spakowicz/" -->
<!-- ``` -->

read in necessary data
```{r warning=FALSE}
# irAEs
irAE <- readxl::read_xlsx(file.path(paths$tdrive,"mitox/Clinical info/Mitoxstudy_patient_symptoms_AM_2.xlsx")) %>%
  dplyr::select(-`Patient Name`, - MRN)

# recist, CT scans
recist.scan <- readxl::read_xlsx(file.path(paths$tdrive, "mitox/Clinical info/OSU-19125_patient_progression_RECIST_KK_v4.xlsx")) %>%
  dplyr::select(-Name, -MRN)

# microbiome metadata
micSamp <- read.table(file.path(paths$tdrive, 'mitox/data/labii/mitox-samples-export_2022-10-21.tsv'), sep = '\t', header = TRUE, fill = NA)

# demographics
demo.dat <- read_excel(file.path(paths$tdrive, "mitox/Patient Logs/Demographics/Patient_demographics.xlsx")) %>%
   select(-MRN, -Name)

# PFS
# Nye maybe changed the sheet name to 12-Month PFS ?
pfs <- readxl::read_xlsx(file.path(paths$tdrive, "mitox/Clinical info/OSU-19125_patient_progression_RECIST_KK_v4.xlsx"),
                               sheet = "12-month PFS") %>%
  select(-MRN)

# microbiome taxonomic data merged metaphlan output 
# old processing 
# taxa.o <- read.csv("../../../data/mpa_aggregated_long_2023-10-26.csv")
# new processing
taxa <- read.csv("../../../data/2024-08-30_mpa_aggregated_long.csv")
# counts
counts <- read.csv("../../../data/2024-08-30_mpa-counts_aggregated_long.csv")
# virus output
vir <- read.csv("../../../data/2024-08-30_vsc_aggregated_long.csv")

# microbiome sample matching 
samps <- read_xlsx( file.path(paths$onedrive, "/data/onecodex/2021-10-27_mitox_One Codex Sample Registration.xlsx"))
samps2 <- read_xlsx( file.path(paths$onedrive, "/data/onecodex/2022-07-27_mitox_remaining_One Codex Sample Registration.xlsx")) %>%
   select(-`Specimen Type`)
samps <- rbind(samps, samps2) 

# blood data
blood.adj <- openxlsx::readWorkbook(xlsxFile =  file.path(paths$tdrive, "mitox/Clinical info/Patient Data 7-1-22.xlsm"), sheet = "Adj Datasheet", startRow = 2)
blood.met <- openxlsx::readWorkbook(xlsxFile =  file.path(paths$tdrive, "mitox/Clinical info/Patient Data 7-1-22.xlsm"), sheet = "Met Datasheet", startRow = 2)
blood.metj <- openxlsx::readWorkbook(xlsxFile =  file.path(paths$tdrive, "mitox/Clinical info/Patient Data 7-1-22.xlsm"), sheet = "Jason Met Cohort", startRow = 2, cols = 3:53) 

# Patient log
pt_log <- readxl::read_xlsx(file.path(paths$tdrive,"mitox/Patient Logs/OSU 19125 Patient Log.xlsx"), skip = 3) %>%
  dplyr::select(-`Pt Name`, -MRN)

# Nanostring summary 
nano.adj <- readxl::read_xlsx(file.path(paths$tdrive, "mitox/data/Burd_clinicdata/Burd_NanostringSummary_June2022.xlsx"), sheet = 2)
nano.metdis <- readxl::read_xlsx(file.path(paths$tdrive, "mitox/data/Burd_clinicdata/Burd_NanostringSummary_June2022.xlsx"), sheet = 4)
nano.metval <- readxl::read_xlsx(file.path(paths$tdrive, "mitox/data/Burd_clinicdata/Burd_NanostringSummary_June2022.xlsx"), sheet = 5)
nano.metnivo <- readxl::read_xlsx(file.path(paths$tdrive, "mitox/data/Burd_clinicdata/Burd_NanostringSummary_June2022.xlsx"), sheet = 6)
```

Demographics (Input to Tables 1, S1)
Rows are pts
```{r}
# clean up demographic data
# MDSC-152 should be MDSC-152B
# manually change
demographics <- demo.dat %>% 
  filter(!is.na(`Patient Id`)) %>%
  select(-c(No., Notes)) %>%
  mutate(`MDSC #` = ifelse(`MDSC #` == "MDSC-152", "MDSC-152B", `MDSC #`))

# get irAE, symptoms, response, pfs, microbiome sample presence
recist.mic <- recist.scan %>%
  select(`Patient ID`, RECIST, Category, Samples) %>%
  rename("Patient Id" = `Patient ID`,
         "Mic.Samp" = Samples) %>%
  filter(!is.na(`Patient Id`))

pfs.filt <- pfs %>%
  select(-Notes, -Cohort) %>%
  mutate(days = `Date of progression/death/toxicity` - `C1D1 Date`,
         `6-month PFS` = ifelse(!is.na(`6-month PFS`), `6-month PFS`,
                                ifelse(days < 183, "No", "Yes"))) %>%
  select(`Patient Id`, `6-month PFS`, `12-month PFS`)
  
  
irae <- irAE %>%
  dplyr::select(`Patient Id`, `Attribution (Final)`, Onset) %>%
  left_join(pt_log %>% select(`Patient Id`, `C1D1 date`)) %>%
  filter(`Attribution (Final)` == "Yes") %>%
  mutate(days = as.Date(Onset) - as.Date(`C1D1 date`),
         irAE = `Attribution (Final)`) %>%
  filter(0 <= days,
        days < 113) %>%
  select(`Patient Id`, irAE) %>%
  distinct()

# join 
demographics <- merge(demographics, recist.mic, all=T)
demographics <- merge(demographics, pfs.filt, all=T)
demographics <- merge(demographics, irae, all=T)

# slight mutation to irae and MDSC
# create response value 
demographics <- demographics %>%
  mutate(irAE = case_when(is.na(irAE) ~ "No", .default = irAE),
         `MDSC #` = case_when(`MDSC #` == "N/A"~NA, .default = `MDSC #`),
          Response = case_when(
            Category == "PD" ~ "Non-Responder",
            Category %in% c("PR", "CR") ~ "Responder",
            Category == "SD" & `6-month PFS`== "Yes" ~ "Responder",  # Logical column `6-month PFS`
            Category == "SD" & `6-month PFS`== "No" ~ "Non-Responder",
            TRUE ~ NA ),
         # MET patients are only stage 3 and 4
          Stage = case_when(
            Stage == 2 ~ "Stage II",
            Stage == 3 ~ "Stage III", 
            Stage == 4 ~ "Stage IV",
            TRUE ~ NA)
         )

# filter to MET only
# has microbiome sample
# has response info
# remove M005
demographics <- demographics %>%
  filter(Cohort == "Met",
         Mic.Samp == "Yes",
         !is.na(Category),
         `Patient Id` != "M005") %>%
  # remove duplicate entry?
  distinct()

save(demographics, file = "../../data/clinical.rds")
```

Blood data
Rows are blood samples
ptID, genes
```{r}
nano.adj <- nano.adj %>%
  select(Sample, ADORA2A:POS_F)
nano.metdis <- nano.metdis %>%
  select(Sample, ADORA2A:POS_F)
nano.metnivo <- nano.metnivo %>%
  select(Sample, ADORA2A:POS_F)
nano.metval <- nano.metval %>%
  select(Sample, ADORA2A:POS_F)

key <- demographics %>%
  select(`Patient Id`, `MDSC #`)

blood <- rbind(nano.adj, nano.metdis, nano.metnivo, nano.metval) %>%
  filter(!is.na(Sample)) %>%
  mutate(`MDSC #` = gsub(" ", "-", Sample)) %>%
  # manually remove 203 due to early timepoint
  # pt was adj -> met
  filter(`MDSC #` != "MDSC-203") %>%
  left_join(key) %>%
  filter(!is.na(`Patient Id`)) %>%
  select(`Patient Id`, 2:94) %>%
  filter(!is.na(ADORA2A))

save(blood, file = "../../data/blood.rds")
```


Start Events object (input to Figure 1C (timeline figure))
Rows are events
ptID, event, time from C1D1 
```{r}
## clean up start and end dates
eos.d <-  pt_log %>% 
  select(`Patient Id`, `C1D1 date`, `EOS Date`) %>%
  filter(!is.na(`C1D1 date`),
         !is.na(`EOS Date`)) %>%
  mutate(`C1D1 date` = as.Date(`C1D1 date`),
         `EOS Date` = as.Date(`EOS Date`),
          Days = `EOS Date` - `C1D1 date`,
          Event = "End Study") %>%
  select(`Patient Id`, Days, Event)

start.key <- pt_log %>%
  select(`Patient Id`, `C1D1 date`) %>%
  filter(!is.na(`Patient Id`),
         !is.na(`C1D1 date`))

start.d <- start.key %>%
  mutate(Days = 0,
         Event = "Start Study") %>%
  select(-`C1D1 date`)
  
## clean up irAE dates
irae.d <- irAE %>%
  select(`Patient Id`, Onset, `Attribution (Final)`) %>%
  rename(irAE = `Attribution (Final)`) %>%
  fill(`Patient Id`) %>%
  #mutate(irAE_date = as.Date(as.character(Onset))) %>%
  mutate(irAE_date = Onset) %>%
  filter(irAE == "Yes",
          !is.na(irAE_date)) %>%
  select(`Patient Id`, irAE_date) %>%
  left_join(start.key) %>%
  #mutate(Days = irAE_date - `C1D1 date` ,
  mutate(Days = difftime(irAE_date, `C1D1 date`, units = "days"),
         Event = "irAE") %>%
  select(`Patient Id`, Days, Event)

## clean up microbiome sample dates

# get average number of days between date collected and date processed 
x <- micSamp %>%
  filter(sample_id != "",
         Type.of.Sample != "") %>%
  mutate(test = as.integer(as.Date(date_processed) - as.Date(Date.Sample.Collected))) %>%
  filter(!is.na(test))

x <- round(mean(x$test))
```

Microbiome sample metadata
Rows are microbiome samples
ptID, category label [baseline, 12wk, irae, control], link to 
```{r}
temp.mic <- micSamp %>%
  filter(sample_id != "",
         Type.of.Sample != "") %>%
  mutate(`Patient Id` = ifelse(record_id != "", record_id, substr(sample_id, 0, 4)),
         sample_id = gsub(" ", ".", sample_id),
         #date = Date.Sample.Collected) %>%
         date = as.Date(Date.Sample.Collected),
         date_processed = as.Date(date_processed),
         date = case_when(is.na(date)~ date_processed-x, .default = date)) %>%
  select(`Patient Id`, sample_id, date) %>%
  filter(!is.na(`Patient Id`)) %>%
  distinct() %>%
  left_join(start.key) %>%
  mutate(Days = difftime(date, `C1D1 date`, units = "days")) %>%
  select(`Patient Id`, sample_id, Days)

samps <- samps %>%
  select("Barcode ID", "Name") %>%
  mutate(sample_id = gsub("_", ".", Name)) %>%
  mutate(sample_id = gsub(" ", ".", sample_id)) %>%
  select(-Name) %>%
  rename("sample" = "Barcode ID")

mic.meta <- left_join(samps, temp.mic) %>%
  distinct()

# filter to met
mic.meta <- mic.meta %>%
  filter(`Patient Id` %in% demographics$`Patient Id`)

save(mic.meta, file = "../../data/microbiome_metadata.rds")
```
ISSUE - sid = ruxA0a40xe3f4tyDINSX3  has record_id = M020, but the sample_id = M011.200911.2
RESOLVED it is M020

ISSUE M048 has a 12wk microbiome sample but no collection date 
RESOLVED average days between collection and processing and add collection dates based on processing date 

# Complete Events Object 
```{r}
mic.d <- micSamp %>%
  filter(sample_id != "",
         Type.of.Sample != "") %>%
  # use sample_id substr instead of record_id, which is missing values
  mutate(`Patient Id` = ifelse(record_id != "", record_id, substr(sample_id, 0, 4)),
         sample_id = gsub(" ", ".", sample_id),
         #date = Date.Sample.Collected) %>%
         date = as.Date(Date.Sample.Collected),
         date_processed = as.Date(date_processed)) %>%
  filter(sample_id %in% mic.meta$sample_id) %>%
  select(`Patient Id`, date, date_processed) %>%
  distinct() %>%
  mutate(date = case_when(is.na(date)~ date_processed-x, .default = date)) %>%
  select(`Patient Id`, date) %>%
  left_join(start.key) %>%
  #mutate(Days = date - `C1D1 date`
  mutate(Days = difftime(date, `C1D1 date`, units = "days"),
         Event = "Microbiome") %>%
  select(`Patient Id`, Days, Event)

## clean up scan dates
scan.d <- recist.scan %>%
  select(`Patient ID`, `Baseline Scan`, `12th Week Scan`, `Other Scan`) %>%
  filter(!is.na(`Patient ID`)) %>%
  rename(`Patient Id` = `Patient ID`) %>%
  mutate(CT.baseline = openxlsx::convertToDate(`Baseline Scan`),
         CT.12wk = openxlsx::convertToDate(`12th Week Scan`),
         CT.other = as.Date(`Other Scan`)) %>%
  select(`Patient Id`, CT.baseline, CT.12wk, CT.other) %>%
  # WIDE TO LONG
  pivot_longer(!`Patient Id`, names_to = "Scan", values_to = "date") %>%
  filter(!is.na(date)) %>%
  left_join(start.key) %>%
  mutate(Days = difftime(date, `C1D1 date`, units = "days"),
         Event = "CT Scan") %>%
  select(`Patient Id`, Days, Event)

## clean up blood draw dates
blood.adj <- blood.adj %>%
  subset(., select = which(!duplicated(names(.)))) %>%
  select(`Coded.Sample.#`, Date.of.Draw)

blood.met <- blood.met %>%
  subset(., select = which(!duplicated(names(.)))) %>%
  select(`Coded.Sample.#`, Date.of.Draw)

blood.metj <- blood.metj %>%
  subset(., select = which(!duplicated(names(.)))) %>%
  select(`Coded.Sample.#`, Date.of.Draw)

blood.d <- rbind(blood.adj, blood.met, blood.metj) %>%
  mutate(`MDSC #` = substr(`Coded.Sample.#`, 0, 9)) %>%
  left_join(key) %>%
  filter(!is.na(`Patient Id`)) %>%
  mutate(`Blood Draw` = openxlsx::convertToDate(Date.of.Draw)) %>%
  select(`Patient Id`, `Blood Draw`) %>%
  filter(!is.na(`Blood Draw`)) %>%
  left_join(start.key) %>%
  mutate(Days = difftime(`Blood Draw`, `C1D1 date`, units = "days"),
         Event = "Blood Draw") %>%
  select(`Patient Id`, Days, Event)
  
## COMBINE AND SAVE EVENTS OBJECT
events <- rbind(start.d, eos.d, mic.d, irae.d, blood.d, scan.d) %>%
  filter(`Patient Id` %in% demographics$`Patient Id`)

save(events, file = "../../data/events.rds")
```


Microbiome data RA 
Rows are microbes, note sum for each level will by 100, but sum of all will not
be meaningful
```{r}
ra.w <- taxa %>%
  select(sample, Taxonomy, RelAbun) %>%
  filter(Taxonomy != "UNCLASSIFIED") %>%
  pivot_wider(names_from = sample, values_from = RelAbun, values_fill = 0) %>%
  mutate(microbe = sub(".*\\|", "", Taxonomy)) %>%
  # filter to MET only
  select(microbe, mic.meta$sample)

ra.l <- ra.w %>%
  pivot_longer(!microbe, names_to = "sample", values_to = "RelAbun")

save(ra.w, file = "../../data/ra-wide.rds")
save(ra.l, file = "../../data/ra-long.rds")
```

estimated counts
```{r}
counts.w <- counts %>%
  select(sample, Taxonomy, est.reads) %>%
  pivot_wider(names_from = sample, values_from = est.reads, values_fill = 0) %>%
  mutate(microbe = sub(".*\\|", "", Taxonomy)) %>%
  # filter to MET only
  select(microbe, mic.meta$sample)

counts.l <- counts %>%
  select(-TaxNum, -RelAbun, -coverage)

save(counts.w, file = "../../data/counts-wide.rds")
save(counts.l, file = "../../data/counts-long.rds")
```


Virus output 
don't know how we are using yet so will leave in current format 
```{r}
vir <- vir %>%
  # filter to MET only
  filter(sample %in% mic.meta$sample)

save(vir, file = "../../data/vsc.rds")
```


irAE table
Rows are irAEs
ptID, grade, type (input to Table S2)
```{r}
irae <- irAE %>%
  dplyr::select(`Patient Id`, Symptoms, `CTCAE grade (final)`, `Attribution (Final)`, Onset) %>%
  rename(irAE = `Attribution (Final)`) %>%
  fill(`Patient Id`) %>%
  filter(!is.na(`Patient Id`),
         irAE == "Yes") %>%
  left_join(start.key) %>%
  #mutate(Days = irAE_date - `C1D1 date` ,
  mutate(Days = difftime(Onset, `C1D1 date`, units = "days")) %>%
  # filt to met
  filter(`Patient Id` %in% demographics$`Patient Id`) %>%
  select(-irAE, -Onset, -`C1D1 date`)

save(irae, file = "../../data/irae.rds")
```

make phylum labels
```{r}
phyl_labels <- taxa %>%
  select(Taxonomy) %>%
  separate(Taxonomy, c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species", "t"), "\\|") %>%
  select(-Kingdom) %>%
  pivot_longer(cols = `Class`:`t`, 
               names_to = "level",
               values_to = "microbe") %>%
  select(-level) %>%
  mutate(Phylum = substr(Phylum, 4, nchar(Phylum))) %>%
  distinct() %>%
  filter(!is.na(Phylum)) %>%
  mutate(microbe = ifelse(is.na(microbe), paste0("p__", Phylum), microbe))

save(phyl_labels, file = "../../data/phylum-labels.rds")
```

make taxonomy ID table
```{r}
tax.key <- taxa %>%
  select(Taxonomy) %>%
  separate(Taxonomy, c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species", "t"), "\\|") %>%
  select(-t) %>%
  mutate(microbe = ifelse(!is.na(Species), Species,
                          ifelse(!is.na(Genus), Genus,
                                 ifelse(!is.na(Family), Family,
                                        ifelse(!is.na(Order), Order,
                                               ifelse(!is.na(Class), Class,
                                                      ifelse(!is.na(Phylum), Phylum,
                                                             Kingdom)))))),
         taxon = substr(microbe, 4, nchar(microbe))) 

save(tax.key, file = "../../data/taxonomy-key.rds")
```


