---
title: "processing script"
author: "Caroline Wheeler"
date: "4/3/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

find ./PRJNA541981 -mindepth 2 -type f -exec mv {} ./PRJNA541981/ \;
find ./PRJNA541981 -mindepth 1 -type d -empty -delete
^ this step should/could be added to the download batch job to get the files where we want them

## get sample names
```{r cars}
generate_humann3 <- function(proj, samps){
  fileOut <- file(paste0("/fs/ess/PAS1695/projects/meta-analysis/batch/", proj, "_humann3.pbs"))
  writeLines(c(paste0("#PBS -N humann3_", proj),
               "#PBS -A PAS1695",
               "#PBS -l walltime=24:00:00",
               "#PBS -l nodes=1:ppn=28",
               "#PBS -j oe",
               "",
               
               #List sample naames in a way that can be indexed
               paste0("snames=('", paste(samps, collapse = "' '"), "')"),
               "",
               paste0("cd /fs/ess/PAS1695/projects/meta-analysis/data/", proj, "/concat_fastqs"),
               "module load python/3.9-2022.05",
               "source activate humann3",
               "source activate metaphlan3",
               "",
               paste0("humann -i ${snames[$PBS_ARRAY_INDEX]}.fastq.gz -o ../humann3/${snames[$PBS_ARRAY_INDEX]}",
                      " --threads 28 --input-format fastq.gz --nucleotide-database",
                      " /fs/ess/PAS1695/db/chocophlan/chocophlan --protein-database", 
                      " /fs/ess/PAS1695/db/chocophlan/uniref"),
               ""
    ),
    fileOut
  )
}

generate_batch <- function(project){
  fastq.dir <- file.path("/fs", "ess", "PAS1695","projects", "meta-analysis", "data", project, "fastqs")
  proj.dir <- file.path("/fs", "ess", "PAS1695","projects", "meta-analysis", "data", project)
  all.files <- list.files(fastq.dir, pattern = ".fastq")
  snames <- gsub("_.*$", "", all.files)
  samples <- unique(snames)
  
  # commented out b/c ran separately for first 4 projects
  # this is to get the ENA files in the folder structure we want, in future can add this to the download batch job
  # cd.command <- paste0("cd ", proj.dir)
  # mv.command <- paste0("find ./", project, " -mindepth 2 -type f -exec mv {} ./", project, "/ \;")
  # rm.command <- paste0("find ./", project, " -mindepth 1 -type d -empty -delete")
  # system(cd.command)
  # system(mv.command)
  # system(rm.command)
  
  # for(s in samples){
  #   concat.command <- paste0("cat ", fastq.dir, "/", s, "_1.fastq.gz ", fastq.dir, "/", s, "_2.fastq.gz > ",
  #                            proj.dir, "/concat_fastqs/", s, ".fastq.gz")
  #   system(concat.command)
  # }
  generate_humann3(project, samples)
}
```
 fileOut<-file(paste0("/fs/scratch/PDE0023/georgia/batch/STAR_align.pbs"))
  
  # these two are getting the fastq files
  fastq1 <- paste0("/fs/ess/PDE0023/georgia/fastqs/${snames[$PBS_ARRAY_INDEX]}_R1_001.fastq.gz")
  fastq2 <- paste0("/fs/ess/PDE0023/georgia/fastqs/${snames[$PBS_ARRAY_INDEX]}_R2_001.fastq.gz")
  
  # The actual content in job scripts
  writeLines(c(paste0("#PBS -N georgia_STAR_align"),
               "#PBS -l walltime=20:00:00",
               "#PBS -l nodes=1:ppn=28",
               "#PBS -j oe",
               "#PBS -m ae",
               "#PBS -A PAS1695",
               "",
```{r}
generate_batch("PRJEB22893")
# generate_batch("PRJNA397906")
# generate_batch("PRJNA399742")
# generate_batch("PRJNA541981")
# generate_batch("PRJNA762360")
```
