---
title: "sanity check"
output: html_document
date: "2024-10-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

load data
```{r}
load("../manuscript/data/counts-long.rds")
load("../manuscript/data/ra-long.rds")
load("../manuscript/data/clinical.rds")
load("../manuscript/data/microbiome_metadata.rds")
```

Some of the top microbes:
Segatella
Bifidobacterium dentium
Enterobacter_hormaechei
Anaerotignum
```{r}
top <- c("g__Segatella", "s__Bifidobacterium_dentium", "s__Enterobacter_hormaechei", "g__Anaerotignum")
```


create inputs for counts
```{r}
baseline <- mic.meta %>%
  group_by(`Patient Id`) %>%
  arrange(Days) %>%
  slice_head(n = 1) %>%
  ungroup()

clin <- dplyr::left_join(baseline, demographics, by = "Patient Id") %>%
  dplyr::filter(sample %in% counts.l$sample)

irae <- clin %>% 
  dplyr::select(sample, irAE) %>%
  dplyr::filter(sample %in% counts.l$sample) %>%
  distinct()

top_irae <- counts.l %>%
  dplyr::filter(sample %in% irae$sample) %>%
  dplyr::left_join(irae) %>%
  filter(microbe %in% top)
```

```{r}
top_irae %>%
  group_by(microbe, irAE) %>%
  summarize(total = sum(counts))
```

further check those with 0 count in either group 
```{r}
counts.irae <- counts.l %>%
  dplyr::filter(sample %in% irae$sample) %>%
  dplyr::left_join(irae) %>%
  group_by(microbe, irAE) %>%
  summarize(total = sum(counts)) %>%
  ungroup() %>%
  filter(total == 0) %>%
  mutate(taxon = substr(microbe, 4, nchar(microbe)))
```


```{r}
load("../data/deseq_counts-pseudo.rds")
load("../data/deseq_counts-poscounts.rds")
counts.poscounts <- resPlot_irae
rm(resPlot_irae)
load("../data/ancombc2_counts.rds")
```

is ANCOM identifying the taxa with 0 in one group? Nope
```{r}
ancom.sig <- res.tab.c %>%
  filter(p_irAEYes < 0.05)

intersect(ancom.sig$taxon, counts.irae$taxon)
```

What happens when we take these mics out of deseq results?
```{r}
deseq.c.pos <- counts.poscounts %>%
  mutate(taxon = substr(microbe, 4, nchar(microbe))) %>%
  filter(!taxon %in% counts.irae$taxon,
         padj < 0.05) 
deseq.c.pseudo <- counts.pseudo %>%
  mutate(taxon = substr(microbe, 4, nchar(microbe))) %>%
  filter(!taxon %in% counts.irae$taxon,
         padj < 0.05)

x = list(ancom.sig$taxon, deseq.c.pos$taxon, deseq.c.pseudo$taxon)
names(x) <- c("c.ANCOM","c.poscounts","c.pseudo")

ggVennDiagram(x, label_alpha = 0) + scale_fill_gradient(low = "#F4FAFE", high = "#4981BF") + ggtitle("all significant")


c.pos.top <- (deseq.c.pos %>% arrange(padj) %>% slice_head(n = 50))$taxon
c.pseudo.top <- (deseq.c.pseudo %>% arrange(padj) %>% slice_head(n = 50))$taxon
ancom.top.c <- (ancom.sig %>% arrange(p_irAEYes)%>% slice_head(n = 50))$taxon

x = list(ancom.top.c, c.pseudo.top , c.pos.top)
names(x) <- c("c.ANCOM","c.pseudo","c.poscounts")

ggVennDiagram(x, label_alpha = 0) + scale_fill_gradient(low = "#F4FAFE", high = "#4981BF") + ggtitle("top 50")
```

get group not sig in deseq results
```{r}
x <- ancom.sig %>%
  filter(!taxon %in% deseq.c.pos$taxon & !taxon %in% deseq.c.pseudo$taxon) %>%
  arrange(p_irAEYes)
```

Lets check - ANCOM says significant, but not deseq
"Intestinimonas_massiliensis"
"Lacrimispora_amygdalina"
"Bifidobacterium_adolescentis"
"Akkermansia_muciniphila"
```{r}
mics <- c("s__Intestinimonas_massiliensis",
"s__Lacrimispora_amygdalina",
"s__Bifidobacterium_adolescentis",
"s__Akkermansia_muciniphila")

counts.l %>%
  dplyr::filter(sample %in% irae$sample) %>%
  dplyr::left_join(irae) %>%
  filter(microbe %in% mics) %>%
  group_by(microbe, irAE) %>%
  summarize(total = sum(counts)) %>%
  ungroup()
```

kw test
```{r}
y <- counts.l %>%
  dplyr::filter(sample %in% irae$sample) %>%
  filter(microbe == "s__Akkermansia_muciniphila") %>%
  dplyr::left_join(irae) 

kruskal.test(counts ~ irAE, data = y)
```
loop through KW test 
```{r}

```

