---
title: "trial ANCOM-BC2"
output: html_document
date: "2024-10-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ANCOMBC)
library(phyloseq)
library(tidyverse)
library(ggplot2)
library(ggrepel)
```

load data
```{r}
taxa <- read.csv("../data/2024-08-30_mpa_aggregated_long.csv")
# counts <- read.csv("../data/2024-08-30_mpa-counts_aggregated_long.csv")

load("../manuscript/data/counts-long.rds")
load("../manuscript/data/clinical.rds")
load("../manuscript/data/microbiome_metadata.rds")
```

prepare data
```{r}
baseline <- mic.meta %>%
  group_by(`Patient Id`) %>%
  arrange(Days) %>%
  slice_head(n = 1) %>%
  ungroup()

clin <- left_join(baseline, demographics, by = "Patient Id") %>%
  dplyr::filter(sample %in% counts.l$sample)

irae <- clin %>% 
  dplyr::select(sample, irAE) %>%
  dplyr::filter(sample %in% counts.l$sample) %>%
  distinct()

taxa <- taxa %>%
  filter(sample %in% irae$sample,
         grepl("s__", Taxonomy),
         !grepl("t__", Taxonomy)) %>%
  select(-TaxNum, -Alternative.Tax) %>%
  pivot_wider(names_from = "sample", values_from = "RelAbun", values_fill = 0) %>%
  column_to_rownames("Taxonomy")

irae <- irae %>%
  arrange(factor(sample, levels = colnames(taxa))) %>%
  column_to_rownames("sample")

#check
all.equal(rownames(irae), colnames(taxa))
```

<!-- set up function run -->

<!-- If a matrix or data.frame is provided, ensure that the row names of the metadata match the sample names (column names if taxa_are_rows is TRUE, and row names otherwise) in data -->

<!-- ... ^ outdated documentation, must be a phyloseq or TreeSummarizedExperiment object  -->
<!-- ```{r} -->
<!-- # taxa_are_rows = TRUE is default  -->
<!-- ANCOMBC::ancombc2(data = tax.levs[1]) -->
<!-- ``` -->

Metaphlan to Phyloseq to TreeSummarizeExperiment
```{r}
## tax is a matrix or data.frame with the table of taxonomic abundances, rows are taxa, columns are samples
## metadat is an optional data.frame of specimen metadata, rows are samples, columns are variables
## if simplenames=TRUE, use only the most detailed level of taxa names in the final object
## if roundtointeger=TRUE, values will be rounded to the nearest integer
metaphlanToPhyloseq <- function(tax, metadat, simplenames=TRUE, 
                                roundtointeger=FALSE, split="|"){
  
    xnames = rownames(tax)
    shortnames = gsub(paste0(".+\\", split), "", xnames)
    
    if(simplenames){
        rownames(tax) = shortnames
    }
    
    if(roundtointeger){
        tax = round(tax * 1e4)
    }
    
    x2 = strsplit(xnames, split=split, fixed=TRUE)
    taxmat = matrix(NA, ncol=max(sapply(x2, length)), nrow=length(x2))
    colnames(taxmat) = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species", "Strain")[1:ncol(taxmat)]
    rownames(taxmat) = rownames(tax)
    
    for (i in 1:nrow(taxmat)){
        taxmat[i, 1:length(x2[[i]])] <- x2[[i]]
    }
    
    taxmat = gsub("[a-z]__", "", taxmat)
    taxmat = phyloseq::tax_table(taxmat)
    otutab = phyloseq::otu_table(tax, taxa_are_rows=TRUE)
    
    if(is.null(metadat)){
        res = phyloseq::phyloseq(taxmat, otutab)
    }else{
        res = phyloseq::phyloseq(taxmat, otutab, phyloseq::sample_data(metadat))
    }
    
    return(res)
}

# RA - not converted to integer
phylo <- metaphlanToPhyloseq(taxa, irae)
phylo

tree_ob = mia::makeTreeSummarizedExperimentFromPhyloseq(phylo)
tree_ob
```


# check that phyloseq object is set up properly
```{r}
library(phyloseq)
library(microViz)

phylo %>%
  tax_agg("Phylum") %>%
  ps_seriate(dist = "bray", method = "OLO_ward") %>% # these are the defaults
  comp_barplot(tax_level = "Phylum", sample_order = "asis", n_taxa = 5)
```
Try running ANCOM-BC2

Warning: The group variable has < 3 categories 
The multi-group comparisons (global/pairwise/dunnet/trend) will be deactivated

This is expected behavior. 



Warning: NaNs producedWarning: NaNs producedWarning: NaNs producedWarning: NaNs produced - repeatedly

Warning: longer object length is not a multiple of shorter object length - repeatedly 

Tried installing package from Pbrad branch version with fix for this error, 
still not resolved, just going to run and compare to deseq2 results 

```{r, warning = FALSE}
run_ancombc2 <- function(level) {
  # try with phyloseq object
out <- ANCOMBC::ancombc2(data = phylo, 
                         tax_level = level, 
                         fix_formula = "irAE", 
                         group = "irAE", 
                         struc_zero = TRUE, 
                         s0_perc = 0.05,
                         pseudo_sens = TRUE)

res <- out$res
return(res)

}

res_spec <- run_ancombc2("Species")
res_gen <- run_ancombc2("Genus")
res_fam <- run_ancombc2("Family")
res_ord <- run_ancombc2("Order")
res_class <- run_ancombc2("Class")
res_phyl <- run_ancombc2("Phylum")
res_king <- run_ancombc2("Kingdom")

res.tab <- rbind(res_spec, res_gen, res_fam, res_ord, res_class, res_phyl, res_king)
```

save output
```{r}
save(res.tab, file = "../data/ancombc2_RA.rds")
```

plot
```{r}
res.tab %>%
  ggplot(aes(x=lfc_irAEYes, y=-log10(p_irAEYes), label = taxon)) + 
  geom_point() + 
  geom_text_repel() + 
  theme_minimal()
```




Example 
```{r}
data(atlas1006, package = "microbiome")
tse = mia::makeTreeSummarizedExperimentFromPhyloseq(atlas1006)

# subset to baseline
tse = tse[, tse$time == 0]

# Re-code the bmi group
tse$bmi = recode(tse$bmi_group,
                 obese = "obese",
                 severeobese = "obese",
                 morbidobese = "obese")
# Subset to lean, overweight, and obese subjects
tse = tse[, tse$bmi %in% c("lean", "overweight", "obese")]

# Note that by default, levels of a categorical variable in R are sorted 
# alphabetically. In this case, the reference level for `bmi` will be 
# `lean`. To manually change the reference level, for instance, setting `obese`
# as the reference level, use:
tse$bmi = factor(tse$bmi, levels = c("obese", "overweight", "lean"))
# You can verify the change by checking:
# levels(sample_data(tse)$bmi)

# Create the region variable
tse$region = recode(as.character(tse$nationality),
                    Scandinavia = "NE", UKIE = "NE", SouthEurope = "SE", 
                    CentralEurope = "CE", EasternEurope = "EE",
                    .missing = "unknown")

# Discard "EE" as it contains only 1 subject
# Discard subjects with missing values of region
tse = tse[, ! tse$region %in% c("EE", "unknown")]

print(tse)
```
```{r}
set.seed(123)
# It should be noted that we have set the number of bootstrap samples (B) equal 
# to 10 in the 'trend_control' function for computational expediency. 
# However, it is recommended that users utilize the default value of B, 
# which is 100, or larger values for optimal performance.
output = ancombc2(data = tse, assay_name = "counts", tax_level = "Family",
                  fix_formula = "age + region + bmi", rand_formula = NULL,
                  p_adj_method = "holm", pseudo_sens = TRUE,
                  prv_cut = 0.10, lib_cut = 1000, s0_perc = 0.05,
                  group = "bmi", struc_zero = TRUE, neg_lb = TRUE,
                  alpha = 0.05, n_cl = 2, verbose = TRUE,
                  global = TRUE, pairwise = TRUE, dunnet = TRUE, trend = TRUE,
                  iter_control = list(tol = 1e-2, max_iter = 20, 
                                      verbose = TRUE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = lme4::lmerControl(),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100),
                  trend_control = list(contrast = list(matrix(c(1, 0, -1, 1),
                                                              nrow = 2, 
                                                              byrow = TRUE),
                                                       matrix(c(-1, 0, 1, -1),
                                                              nrow = 2, 
                                                              byrow = TRUE),
                                                       matrix(c(1, 0, 1, -1),
                                                              nrow = 2, 
                                                              byrow = TRUE)),
                                       node = list(2, 2, 1),
                                       solver = "ECOS",
                                       B = 10))
```






