---
title: "ancom_v_deseq-boxplots"
output: html_document
date: "2024-10-25"
---

Generate boxplots to check DESeq2 vs ANCOM-BC2 results 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggplot2)
library(ggpubr)
library(gridExtra)
```

b/c RA results are very similar to count results in both deseq and ancom, let's
just use count results here to identify taxa of interest

load results and og count data 
```{r}
load("../data/deseq_counts-pseudo.rds")
load("../data/deseq_counts-poscounts.rds")
counts.poscounts <- resPlot_irae
rm(resPlot_irae)

deseq.c.pos <- counts.poscounts %>%
  filter(level == "genus",
         padj < 0.05) %>%
  arrange(padj) %>%
  slice_head(n = 25) %>%
  mutate(taxon = substr(microbe, 4, nchar(microbe)),
         method = "deseq counts poscounts")
deseq.c.pseudo <- counts.pseudo %>%
  filter(level == "genus",
         padj < 0.05) %>%
   arrange(padj) %>%
  slice_head(n = 25) %>%
  mutate(taxon = substr(microbe, 4, nchar(microbe)),
         method = "deseq counts pseudo")
  
load("../data/ancombc2_counts-genus.rds")

ancom.c <- res_gen %>%
  filter(p_irAEYes < 0.05) %>%
  arrange(p_irAEYes) %>%
  slice_head(n = 25) %>%
  mutate(method = "ancom counts")

rm(counts.poscounts, counts.pseudo, res_gen)

load("../manuscript/data/counts-long.rds")
load("../manuscript/data/clinical.rds")
load("../manuscript/data/microbiome_metadata.rds")

baseline <- mic.meta %>%
  group_by(`Patient Id`) %>%
  arrange(Days) %>%
  slice_head(n = 1) %>%
  ungroup()

clin <- dplyr::left_join(baseline, demographics, by = "Patient Id") %>%
  dplyr::filter(sample %in% counts.l$sample)

irae <- clin %>% 
  dplyr::select(sample, irAE) %>%
  dplyr::filter(sample %in% counts.l$sample) %>%
  distinct()

counts.irae <- counts.l %>%
  dplyr::filter(sample %in% irae$sample) %>%
  dplyr::left_join(irae) 

rm(counts.poscounts, counts.pseudo, res_gen, mic.meta, irae, demographics, counts.l, clin, baseline)
```

Pick set of microbes to examine

Let's start with genus level - top 25

21 from ancom.c alone
9 from deseq poscounts alone
7 from deseq pseudo alone
2 shared by ancom and pseudo
14 shared by pos and pseudo
2 shared by all 3
```{r}
ancom.alone <- (ancom.c %>%
  filter(!taxon %in% deseq.c.pos$taxon & !taxon %in% deseq.c.pseudo$taxon))$taxon
pos.alone <- (deseq.c.pos %>%
  filter(!taxon %in% ancom.c$taxon & !taxon %in% deseq.c.pseudo$taxon))$taxon
pseudo.alone <- (deseq.c.pseudo %>%
  filter(!taxon %in% deseq.c.pos$taxon & !taxon %in% ancom.c$taxon))$taxon
ancom.pseudo <- (ancom.c %>%
  filter(!taxon %in% deseq.c.pos$taxon & taxon %in% deseq.c.pseudo$taxon))$taxon
deseq <- (deseq.c.pos %>%
  filter(!taxon %in% ancom.c$taxon & taxon %in% deseq.c.pseudo$taxon))$taxon
all <- (ancom.c %>%
  filter(taxon %in% deseq.c.pos$taxon & taxon %in% deseq.c.pseudo$taxon))$taxon

mics <- c(ancom.alone, pos.alone, pseudo.alone, ancom.pseudo, deseq, all)
```

generate boxplots
```{r}
p <- lapply(ancom.alone, function(mic) {
  counts.irae %>%
    mutate(microbe = substr(microbe, 4, nchar(microbe))) %>%
    filter(microbe == mic) %>%
    ggplot(aes(y = log(counts), x = irAE)) +
    geom_boxplot() +
    theme_bw() +
    stat_compare_means(method = "kruskal.test", vjust = 1) +
    ggtitle(mic)
})

ggsave(
   filename = "../figures/ancom-only-sig.pdf", 
   plot = marrangeGrob(p, nrow=3, ncol=3),
   height = 11, width = 8.5
)

p <- lapply(pos.alone, function(mic) {
  counts.irae %>%
    mutate(microbe = substr(microbe, 4, nchar(microbe))) %>%
    filter(microbe == mic) %>%
    ggplot(aes(y = log(counts), x = irAE)) +
    geom_boxplot() +
    theme_bw() +
    stat_compare_means(method = "kruskal.test", vjust = 1) +
    ggtitle(mic)
})

ggsave(
   filename = "../figures/poscounts-only-sig.pdf", 
   plot = marrangeGrob(p, nrow=3, ncol=3),
   height = 11, width = 8.5
)

p <- lapply(pseudo.alone, function(mic) {
  counts.irae %>%
    mutate(microbe = substr(microbe, 4, nchar(microbe))) %>%
    filter(microbe == mic) %>%
    ggplot(aes(y = log(counts), x = irAE)) +
    geom_boxplot() +
    theme_bw() +
    stat_compare_means(method = "kruskal.test", vjust = 1) +
    ggtitle(mic)
})

ggsave(
   filename = "../figures/pseudo-only-sig.pdf", 
   plot = marrangeGrob(p, nrow=3, ncol=3),
   height = 11, width = 8.5
)

p <- lapply(deseq, function(mic) {
  counts.irae %>%
    mutate(microbe = substr(microbe, 4, nchar(microbe))) %>%
    filter(microbe == mic) %>%
    ggplot(aes(y = log(counts), x = irAE)) +
    geom_boxplot() +
    theme_bw() +
    stat_compare_means(method = "kruskal.test", vjust = 1) +
    ggtitle(mic)
})

ggsave(
   filename = "../figures/deseq-only-sig.pdf", 
   plot = marrangeGrob(p, nrow=3, ncol=3),
   height = 11, width = 8.5
)

p <- lapply(ancom.pseudo, function(mic) {
  counts.irae %>%
    mutate(microbe = substr(microbe, 4, nchar(microbe))) %>%
    filter(microbe == mic) %>%
    ggplot(aes(y = log(counts), x = irAE)) +
    geom_boxplot() +
    theme_bw() +
    stat_compare_means(method = "kruskal.test", vjust = 1) +
    ggtitle(mic)
})

ggsave(
   filename = "../figures/ancom-pseudo-sig.pdf", 
   plot = marrangeGrob(p, nrow=3, ncol=3),
   height = 11, width = 8.5
)

p <- lapply(all, function(mic) {
  counts.irae %>%
    mutate(microbe = substr(microbe, 4, nchar(microbe))) %>%
    filter(microbe == mic) %>%
    ggplot(aes(y = log(counts), x = irAE)) +
    geom_boxplot() +
    theme_bw() +
    stat_compare_means(method = "kruskal.test", vjust = 1) +
    ggtitle(mic)
})

ggsave(
   filename = "../figures/all-sig.pdf", 
   plot = marrangeGrob(p, nrow=3, ncol=3),
   height = 11, width = 8.5
)
```



