---
title: "ancom vs deseq"
output: html_document
date: "2024-10-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(ggrepel)
library(ggVennDiagram)
library(VennDiagram)
```

```{r}
load("../manuscript/data/counts-long.rds")

load("../data/deseq_ra-pseudo.rds")
load("../data/deseq_ra-poscounts.rds")
load("../data/deseq_counts-pseudo.rds")
load("../data/deseq_counts-poscounts.rds")
counts.poscounts <- resPlot_irae
rm(resPlot_irae)
load("../data/deseq_ra-poscounts-sample-scale.rds")
load("../data/deseq_ra-pseudo-sample-scale.rds")

load("../data/ancombc2_RA-estcounts.rds")
load("../data/ancombc2_counts.rds")
```

get taxa present in >1 sample
```{r}
counts.l <- counts.l %>%
  pivot_wider(names_from = "sample", values_from = "counts") %>%
  column_to_rownames("microbe")

# 890 microbes dropped when filter those only present in 1 sample
pres <- rownames(counts.l[rowSums(counts.l == 0) <= 88,])
drop <- rownames(counts.l[rowSums(counts.l == 0) > 88,])
```

```{r}
colnames(res.tab.c)
```

Top Sets .... does not bode well, overlap of 3 with ra poscounts
```{r}
counts.poscounts.top <- substr((counts.poscounts %>% arrange(padj) %>% slice_head(n = 50))$microbe, 4, nchar((counts.poscounts %>% arrange(padj) %>% slice_head(n = 50))$microbe))

counts.pseudo.top <- substr((counts.pseudo %>% arrange(padj) %>% slice_head(n = 50))$microbe, 4, nchar((counts.pseudo %>% arrange(padj) %>% slice_head(n = 50))$microbe))

ancom.top.c <- (res.tab.c %>% arrange(p_irAEYes)%>% slice_head(n = 50))$taxon

x = list(ancom.top.c, counts.pseudo.top , counts.poscounts.top)
names(x) <- c("c.ANCOM","c.pseudo","c.poscounts")

ggVennDiagram(x, label_alpha = 0) + scale_fill_gradient(low = "#F4FAFE", high = "#4981BF")
```

```{r}
ra.poscounts.top <- substr((ra.poscounts %>% arrange(padj) %>% slice_head(n = 50))$microbe, 4 , nchar((ra.poscounts %>% arrange(padj) %>% slice_head(n = 50))$microbe))

ra.pseudo.top <- substr((ra.pseudo %>% arrange(padj) %>% slice_head(n = 50))$microbe, 4, nchar((ra.pseudo %>% arrange(padj) %>% slice_head(n = 50))$microbe))

ancom.top.ra <- (res.tab.ra %>% arrange(p_irAEYes)%>% slice_head(n = 50))$taxon

x = list(ancom.top.ra, ra.poscounts.top,  ra.pseudo.top)
names(x) <- c("ra.ANCOM", "ra.poscounts", "ra.pseudo")

ggVennDiagram(x, label_alpha = 0) + scale_fill_gradient(low = "#F4FAFE", high = "#4981BF")
```

lets only look at microbes present in ANCOM results, since it seems filtered more strictly
should also check what the default filters are
```{r}
ancom.sig.c <- res.tab.c %>%
  filter(p_irAEYes < 0.05)

ancom.sig.ra <- res.tab.ra %>%
  filter(p_irAEYes < 0.05)

counts.pseudo$method <- "pseudo counts"
counts.poscounts$method <- "poscounts counts"
ra.pseudo.ss$method <- "pseudo RA - sample scale"
ra.poscounts.ss$method <- "poscounts RA- sample scale"
ra.poscounts$method <- "poscounts RA - universal"
ra.pseudo$method <- "pseudo RA- universal"

resplot.c <- rbind(counts.pseudo, counts.poscounts) %>%
  mutate(microbe = substr(microbe, 4, nchar(microbe))) %>%
  filter(padj < 0.05,
         microbe %in% ancom.sig.c$taxon)

resplot.ra <- rbind(ra.poscounts, ra.pseudo, ra.pseudo.ss, ra.poscounts.ss) %>%
  mutate(microbe = substr(microbe, 4, nchar(microbe))) %>%
  filter(padj < 0.05,
         microbe %in% ancom.sig.ra$taxon)

table(resplot.c$method)
table(resplot.ra$method)
```

Looking solely at overlap in taxa identified as sig (p < 0.05), 
counts + pseudo count had the most overlap with ANCOM-BC2

lets look at this another way
```{r}
c.filt <- counts.pseudo %>%
  mutate(microbe = substr(microbe, 4, nchar(microbe))) %>%
  filter(padj < 0.05)
  
x = list(ancom.sig.c$taxon, c.filt$microbe)
names(x) <- c("c.ANCOM","c.deseq")

ggVennDiagram(x, label_alpha = 0) + scale_fill_gradient(low = "#F4FAFE", high = "#4981BF")
```


