---
title: "deseq2"
author: "Caroline Wheeler"
date: '2023-01-27'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(DESeq2)
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(dplyr)
```

read in data
```{r}
clin <- read.csv("/Volumes/spakowicz/mitox/data/mitox_table_Recist_irae.csv") %>%
  dplyr::rename("record_id" = Patient.ID,
         "irAE" = Attribution..Final.)
taxa <- read.csv("../data/mpa_aggregated_long_10-5-22.csv")
samp.match <- read.csv("../data/mitox_sample_matching.csv") %>%
  select(sample, record_id)

clin <- merge(clin, samp.match)
```

get irAE df
```{r}
irAE <- clin %>%
  select(sample, irAE) %>%
  filter(!is.na(irAE)) %>%
  distinct() %>%
  filter(sample %in% taxa$sample) 
```

split 'Taxonomy' column
```{r}
taxa <- taxa %>%
   dplyr::select(sample, Taxonomy, RelAbun) %>%
    mutate(kingdom = gsub("(k__\\w+).*", "\\1", Taxonomy),
           phylum = ifelse(grepl("p__", Taxonomy),gsub(".*(p__\\w+).*", "\\1", Taxonomy), NA),
           order = ifelse(grepl("o__", Taxonomy),gsub(".*(o__\\w+).*", "\\1", Taxonomy), NA),
           class = ifelse(grepl("c__", Taxonomy),gsub(".*(c__\\w+).*", "\\1", Taxonomy), NA),
           family = ifelse(grepl("f__", Taxonomy),gsub(".*(f__\\w+).*", "\\1", Taxonomy), NA),
           genus = ifelse(grepl("g__", Taxonomy),gsub(".*(g__\\w+).*", "\\1", Taxonomy), NA),
           species = ifelse(grepl("s__", Taxonomy), gsub(".*(s__.*)", "\\1", Taxonomy), NA)) %>%
  filter(!is.na(species))
```

```{r}
ggplot(taxa, aes(-log10(RelAbun))) + geom_histogram(binwidth = 0.5)
ggplot(taxa, aes(-log10(RelAbun*1e1))) + geom_histogram(binwidth = 0.5)
```

function to get count tables
```{r}
write_desd <- function(taxa, lev){
  desd <- taxa %>%
    dplyr::select("sample", all_of(lev), "RelAbun") %>%
    mutate(count = round(RelAbun*1e1))%>%
    dplyr::select(-"RelAbun") 
  
  return(desd)
}
```

get counts
```{r}
taxa_irAE <- taxa %>%
  filter(sample %in% irAE$sample)

phyl <- write_desd(taxa_irAE, "phylum") %>%
  add_count(sample, phylum, wt=count) %>%
  dplyr::select(-"count") %>%
  distinct() %>%
  spread(sample, n, fill = 0) %>%
  remove_rownames %>%
  column_to_rownames(var="phylum")

class <- write_desd(taxa_irAE, "class") %>%
  add_count(sample, class, wt=count) %>%
  dplyr::select(-"count") %>%
  distinct() %>%
  spread(sample, n, fill = 0) %>%
  remove_rownames %>%
  column_to_rownames(var="class")

ord <- write_desd(taxa_irAE, "order") %>%
  add_count(sample, order, wt=count) %>%
  dplyr::select(-"count") %>%
  distinct() %>%
  spread(sample, n, fill = 0) %>%
  remove_rownames %>%
  column_to_rownames(var="order")

fam <- write_desd(taxa_irAE, "family") %>%
  add_count(sample, family, wt=count) %>%
  dplyr::select(-"count") %>%
  distinct() %>%
  spread(sample, n, fill = 0) %>%
  remove_rownames %>%
  column_to_rownames(var="family")

gen <- write_desd(taxa_irAE, "genus") %>%
  add_count(sample, genus, wt=count) %>%
  dplyr::select(-"count") %>%
  distinct() %>%
  spread(sample, n, fill = 0) %>%
  remove_rownames %>%
  column_to_rownames(var="genus")

spec <- write_desd(taxa_irAE, "species") %>%
  add_count(sample, species, wt=count) %>%
  dplyr::select(-"count") %>%
  distinct() %>%
  spread(sample, n, fill = 0) %>%
  remove_rownames %>%
  column_to_rownames(var="species")
```

function to run DESeq
```{r}
run_deseq <- function(desd, meta, lev){
  ord <- rownames(meta)
  desd <- desd[, ord]
  dds <- DESeqDataSetFromMatrix(countData = desd,
                                colData = meta,
                                design = ~ irAE)

  dds <- DESeq(dds)
  res <- results(dds)
  
  pd <- res$log2FoldChange
  
  resPlot <- data.frame(res) %>% mutate(threshold = padj < 0.05)
  
  resPlot$level <- lev
  return(resPlot)
}
```

make sure cols and rows match and make samples rownames
```{r}
irAE <- irAE %>% arrange(factor(sample, levels = colnames(fam))) %>%
  remove_rownames() %>%
  column_to_rownames(var = "sample")
```


# deseq2 for irAE
```{r}
phyl_rp <- run_deseq(phyl, irAE, "phylum")
clas_rp <- run_deseq(class, irAE, "class")
ord_rp <- run_deseq(ord, irAE, "order")
fam_rp <- run_deseq(fam, irAE, "family")
gen_rp <- run_deseq(gen, irAE, "genus")

# add constant value to all cells in species matrix
# previously getting error
# Every gene contains at least one zero, cannot compute log geometric means
spec_rp <- run_deseq(spec + 1, irAE, "species")
```

```{r}
resPlot_irAE <- rbind(phyl_rp, clas_rp, ord_rp, fam_rp, gen_rp, spec_rp)

# write.csv(resPlot_irAE, "../data/irAE_deseq2_results.csv")
```


## deseq2 for RECIST

get recist df
```{r}
recist <- clin %>%
  filter(!is.na(Category)) %>%
  mutate(response = ifelse(Category == "PD", "NR", "R")) %>%
  select(sample, response) %>%
  distinct() %>%
  filter(sample %in% taxa$sample) 
```

get counts
```{r}
taxa_res <- taxa %>%
  filter(sample %in% recist$sample)

phyl <- write_desd(taxa_res, "phylum") %>%
  add_count(sample, phylum, wt=count) %>%
  dplyr::select(-"count") %>%
  distinct() %>%
  spread(sample, n, fill = 0) %>%
  remove_rownames %>%
  column_to_rownames(var="phylum")

class <- write_desd(taxa_res, "class") %>%
  add_count(sample, class, wt=count) %>%
  dplyr::select(-"count") %>%
  distinct() %>%
  spread(sample, n, fill = 0) %>%
  remove_rownames %>%
  column_to_rownames(var="class")

ord <- write_desd(taxa_res, "order") %>%
  add_count(sample, order, wt=count) %>%
  dplyr::select(-"count") %>%
  distinct() %>%
  spread(sample, n, fill = 0) %>%
  remove_rownames %>%
  column_to_rownames(var="order")

fam <- write_desd(taxa_res, "family") %>%
  add_count(sample, family, wt=count) %>%
  dplyr::select(-"count") %>%
  distinct() %>%
  spread(sample, n, fill = 0) %>%
  remove_rownames %>%
  column_to_rownames(var="family")

gen <- write_desd(taxa_res, "genus") %>%
  add_count(sample, genus, wt=count) %>%
  dplyr::select(-"count") %>%
  distinct() %>%
  spread(sample, n, fill = 0) %>%
  remove_rownames %>%
  column_to_rownames(var="genus")

spec <- write_desd(taxa_res, "species") %>%
  add_count(sample, species, wt=count) %>%
  dplyr::select(-"count") %>%
  distinct() %>%
  spread(sample, n, fill = 0) %>%
  remove_rownames %>%
  column_to_rownames(var="species")
```

make sure cols and rows match and make samples rownames
```{r}
recist <- recist %>% arrange(factor(sample, levels = colnames(fam))) %>%
  remove_rownames() %>%
  column_to_rownames(var = "sample")
```

# deseq2 for recist
```{r}
phyl_rp <- run_deseq(phyl, recist, "phylum")
clas_rp <- run_deseq(class, recist, "class")
ord_rp <- run_deseq(ord, recist, "order")
fam_rp <- run_deseq(fam, recist, "family")
gen_rp <- run_deseq(gen, recist, "genus")

# add constant value to all cells in species matrix
# previously getting error
# Every gene contains at least one zero, cannot compute log geometric means
spec_rp <- run_deseq(spec + 1, recist, "species")
```
```{r}
resPlot_recist <- rbind(phyl_rp, clas_rp, ord_rp, fam_rp, gen_rp, spec_rp)

# write.csv(resPlot_recist, "../data/recist_deseq2_results.csv")
```

## deseq2 for subset

get subset df - response includes PR, CR, SD
```{r}
sub <- clin %>%
  filter(Category != "PD") %>%
  select(sample, irAE) %>%
  distinct() %>%
  filter(sample %in% taxa$sample,
         !is.na(irAE))
```

get counts
```{r}
taxa_sub <- taxa %>%
  filter(sample %in% sub$sample)

phyl <- write_desd(taxa_sub, "phylum") %>%
  add_count(sample, phylum, wt=count) %>%
  dplyr::select(-"count") %>%
  distinct() %>%
  spread(sample, n, fill = 0) %>%
  remove_rownames %>%
  column_to_rownames(var="phylum")

class <- write_desd(taxa_sub, "class") %>%
  add_count(sample, class, wt=count) %>%
  dplyr::select(-"count") %>%
  distinct() %>%
  spread(sample, n, fill = 0) %>%
  remove_rownames %>%
  column_to_rownames(var="class")

ord <- write_desd(taxa_sub, "order") %>%
  add_count(sample, order, wt=count) %>%
  dplyr::select(-"count") %>%
  distinct() %>%
  spread(sample, n, fill = 0) %>%
  remove_rownames %>%
  column_to_rownames(var="order")

fam <- write_desd(taxa_sub, "family") %>%
  add_count(sample, family, wt=count) %>%
  dplyr::select(-"count") %>%
  distinct() %>%
  spread(sample, n, fill = 0) %>%
  remove_rownames %>%
  column_to_rownames(var="family")

gen <- write_desd(taxa_sub, "genus") %>%
  add_count(sample, genus, wt=count) %>%
  dplyr::select(-"count") %>%
  distinct() %>%
  spread(sample, n, fill = 0) %>%
  remove_rownames %>%
  column_to_rownames(var="genus")

spec <- write_desd(taxa_sub, "species") %>%
  add_count(sample, species, wt=count) %>%
  dplyr::select(-"count") %>%
  distinct() %>%
  spread(sample, n, fill = 0) %>%
  remove_rownames %>%
  column_to_rownames(var="species")
```

make sure cols and rows match and make samples rownames
```{r}
sub <- sub %>% arrange(factor(sample, levels = colnames(fam))) %>%
  remove_rownames() %>%
  column_to_rownames(var = "sample")
```

# deseq2 for recist
```{r}
phyl_rp <- run_deseq(phyl, sub, "phylum")
clas_rp <- run_deseq(class, sub, "class")
ord_rp <- run_deseq(ord, sub, "order")
fam_rp <- run_deseq(fam, sub, "family")
gen_rp <- run_deseq(gen, sub, "genus")

# add constant value to all cells in species matrix
# previously getting error
# Every gene contains at least one zero, cannot compute log geometric means
spec_rp <- run_deseq(spec + 1, sub, "species")
```

```{r}
resPlot_sub <- rbind(phyl_rp, clas_rp, ord_rp, fam_rp, gen_rp, spec_rp)

write.csv(resPlot_sub, "../data/irAEsubset_deseq2_results.csv")
```

