---
title: "response clustering"
author: "C. Dravillas"
date: "2025-04-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(compositions)
library(ggplot2)
library(pheatmap)
library(vegan)
library(dendextend)
library(umap)
library(patchwork)
```

## Load data
```{r}
load("../manuscript/data/clinical.rds")
load("../manuscript/data/microbiome_metadata.rds")
load("../manuscript/data/counts-wide.rds")
load("../manuscript/data/counts-long.rds")
load("../manuscript/data/events.rds")
```

## Prepare data... 
- CLR transformation? or use RA
- Need to define time points 

do people clr transform counts?? thought usually RA? 

```{r}
# counts
taxa <- counts.w %>%
  # remove major outlier 
  # dplyr::select(-`FF08034405`) %>%
  column_to_rownames("microbe") 

# add pseudo count for CLR transformation 
taxa[taxa == 0] <- 1 
clr_taxa <- t(apply(taxa, 1, function(x) clr(x)))

# Remove 0 variance taxa - n=891
non_constant_taxa <- apply(clr_taxa, 1, function(x) var(x, na.rm = TRUE)) > 0
clr_taxa_filt <- clr_taxa[non_constant_taxa, ]
clr_taxa_filt_t <- t(clr_taxa_filt)

# check order and add demographic info
meta <- mic.meta %>%
  # remove major outlier 
  # filter(!sample == "FF08034405") %>%
  left_join(demographics)

all.equal(rownames(clr_taxa_filt_t), meta$sample)

# define time points 
irAE.time <- events %>%
  filter(Event == "irAE") %>%
  mutate(irAE.day = Days) %>%
  dplyr::select(`Patient Id`, irAE.day) 

# need to handle multiple irAE dates
# M028, none of them are irAE samples per 10 day rule
# M042, one irAE date makes it an irAE sample 
# should have n=33
meta.irae <- meta %>%
  left_join(irAE.time) %>%
  mutate(irAE.day = ifelse(irAE.day < 0 | irAE.day > 100, NA, irAE.day),
         irAE.sample = ifelse(is.na(irAE.day), 0,
                       ifelse(abs(Days - irAE.day) <= 10, "irAE", "Other"))) %>%
  filter(irAE == "Yes") %>%
  group_by(sample) %>%
  slice_max(irAE.sample, n = 1, with_ties = FALSE) %>%
  ungroup() 

# all samples from all people n=88
meta.all <- meta %>%
  left_join(irAE.time) %>%
  mutate(irAE.day = ifelse(irAE.day < 0 | irAE.day > 100, NA, irAE.day),
         irAE.sample = ifelse(is.na(irAE.day), "Other",
                       ifelse(abs(Days - irAE.day) <= 10, "irAE", "Other"))) %>%
  group_by(sample) %>%
  slice_max(irAE.sample, n = 1, with_ties = FALSE) %>%
  ungroup() 

taxa.irae <- as.matrix(as.data.frame(clr_taxa_filt_t) %>%
  rownames_to_column("sample") %>%
  slice(match(meta.irae$sample, sample)) %>%
  filter(sample %in% meta.irae$sample) %>%
  column_to_rownames("sample"))

taxa.all <- as.matrix(as.data.frame(clr_taxa_filt_t) %>%
  rownames_to_column("sample") %>%
  slice(match(meta.all$sample, sample)) %>%
  filter(sample %in% meta.all$sample) %>%
  column_to_rownames("sample"))

# Remove 0 variance taxa - remove 856 more taxa for irAE pts only 
taxa.irae_filt <- taxa.irae[, apply(taxa.irae, 2, var) > 0]
taxa.all_filt <- taxa.all[, apply(taxa.all, 2, var) > 0]

all.equal(rownames(taxa.irae_filt), meta.irae$sample)
all.equal(rownames(taxa.all_filt), meta.all$sample)
```


## Clustering

# PCA irAE samples only

```{r}
# Run PCA
pca_vegan <- rda(taxa.all_filt , scale = TRUE)
scores_pca <- scores(pca_vegan, display = "sites")

# Combine with metadata
pca_df <- as.data.frame(scores_pca[, 1:2])
colnames(pca_df) <- c("PC1", "PC2")
pca_df$PatientID <- meta.all$`Patient Id`
pca_df$Timepoint <- meta.all$irAE.sample

# Distances in PCA space
pca_dist <- dist(scores_pca)

# By PatientID
bd_patient <- betadisper(pca_dist, group = meta.all$`Patient Id`)
dist_to_patient <- as.data.frame(bd_patient$distances)
dist_to_patient$Group <- meta.all$`Patient Id`
colnames(dist_to_patient) <- c("Distance", "PatientID")

# By Timepoint
bd_timepoint <- betadisper(pca_dist, group = meta.all$irAE.sample)
dist_to_timepoint <- as.data.frame(bd_timepoint$distances)
dist_to_timepoint$Group <- meta.all$irAE.sample
colnames(dist_to_timepoint) <- c("Distance", "Timepoint")

# Requires: dist_to_patient and dist_to_timepoint from earlier

# Base PCA plot colored by PatientID
p1 <- ggplot(pca_df, aes(x = PC1, y = PC2, color = PatientID)) +
  geom_point(size = 3) +
  stat_ellipse(aes(group = PatientID), linetype = "dashed") +
  theme_minimal() +
  theme(legend.position = "none")

# PCA plot colored by Timepoint
p2 <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Timepoint)) +
  geom_point(size = 3) +
  stat_ellipse(aes(group = Timepoint), linetype = "dashed") +
  theme_minimal() 

p_box_pat <- ggplot(dist_to_patient, aes(x = PatientID, y = Distance)) +
  geom_boxplot(fill = "cornflowerblue") +
  theme_minimal() +
  labs(y = "Distance to Centroid") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_box_time <- ggplot(dist_to_timepoint, aes(x = Timepoint, y = Distance)) +
  geom_boxplot(fill = "deeppink3") +
  theme_minimal() +
  labs(y = "Distance to Centroid")

(p1 + p2) / (p_box_pat + p_box_time)
```

```{r}
# Run PCA
pca_vegan <- rda(taxa.irae_filt, scale = TRUE)
scores_pca <- scores(pca_vegan, display = "sites")

# Combine with metadata
pca_df <- as.data.frame(scores_pca[, 1:2])
colnames(pca_df) <- c("PC1", "PC2")
pca_df$PatientID <- meta.irae$`Patient Id`
pca_df$Timepoint <- meta.irae$irAE.sample

# Distances in PCA space
pca_dist <- dist(scores_pca)

# By PatientID
bd_patient <- betadisper(pca_dist, group = meta.irae$`Patient Id`)
dist_to_patient <- as.data.frame(bd_patient$distances)
dist_to_patient$Group <- meta.irae$`Patient Id`
colnames(dist_to_patient) <- c("Distance", "PatientID")

# By Timepoint
bd_timepoint <- betadisper(pca_dist, group = meta.irae$irAE.sample)
dist_to_timepoint <- as.data.frame(bd_timepoint$distances)
dist_to_timepoint$Group <- meta.irae$irAE.sample
colnames(dist_to_timepoint) <- c("Distance", "Timepoint")

# Requires: dist_to_patient and dist_to_timepoint from earlier

# Base PCA plot colored by PatientID
p1 <- ggplot(pca_df, aes(x = PC1, y = PC2, color = PatientID)) +
  geom_point(size = 3) +
  stat_ellipse(aes(group = PatientID), linetype = "dashed") +
  theme_minimal() +
  theme(legend.position = "none")

# PCA plot colored by Timepoint
p2 <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Timepoint)) +
  geom_point(size = 3) +
  stat_ellipse(aes(group = Timepoint), linetype = "dashed") +
  theme_minimal() 

p_box_pat <- ggplot(dist_to_patient, aes(x = PatientID, y = Distance)) +
  geom_boxplot(fill = "cornflowerblue") +
  theme_minimal() +
  labs(y = "Distance to Centroid") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_box_time <- ggplot(dist_to_timepoint, aes(x = Timepoint, y = Distance)) +
  geom_boxplot(fill = "deeppink3") +
  theme_minimal() +
  labs(y = "Distance to Centroid")

(p1 + p2) / (p_box_pat + p_box_time)
```



```{r}
pca_result <- prcomp(taxa.irae_filt, scale. = TRUE)

pca_df <- as.data.frame(pca_result$x[, 1:2])
colnames(pca_df) <- c("PC1", "PC2")

pca_df$PatientID <- meta.irae$`Patient Id`
pca_df$Timepoint <- meta.irae$irAE.sample

# Plot colored by PatientID
p1.pca <- ggplot(pca_df, aes(x = PC1, y = PC2, color = PatientID)) +
  geom_point(size = 3) +
  stat_ellipse(aes(group = PatientID), linetype = "dashed") +
  theme_minimal() +
  theme(legend.position = "right")

# Plot colored by Timepoint (e.g., "irAE", "Other")
p2.pca <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Timepoint)) +
  geom_point(size = 3) +
  stat_ellipse(aes(group = Timepoint), linetype = "dashed") +
  theme_minimal() +
  theme(legend.position = "right")

# Combine plots side by side
pca.plot <- p1.pca + p2.pca
pca.plot

ggsave("../figures/exploratory-outlier-pca-irAE.png", pca.plot, width = 12, height = 6, dpi = 300)
```

# UMAP irAE samples only

```{r}
umap_result <- umap(taxa.irae_filt)

# Convert UMAP output to a dataframe
umap_df <- as.data.frame(umap_result$layout)
umap_df$PatientID <- meta.irae$`Patient Id`
umap_df$Timepoint <- meta.irae$irAE.sample

umap_long <- umap_df %>%
  pivot_longer(cols = c(PatientID, Timepoint), names_to = "Grouping", values_to = "Label")

p1 <- ggplot(umap_df, aes(x = V1, y = V2, color = PatientID)) +
  geom_point(size = 3) +
  stat_ellipse(aes(group = PatientID), linetype = "dashed") +
  theme_minimal() +
  theme(legend.position = "right")

p2 <- ggplot(umap_df, aes(x = V1, y = V2, color = Timepoint)) +
  geom_point(size = 3) +
  stat_ellipse(aes(group = Timepoint), linetype = "dashed") +
  theme_minimal() +
  theme(legend.position = "right")

# Combine plots side by side
p <- p1 + p2 
p

ggsave("../figures/exploratory-umap-irAE.png", p, width = 12, height = 6, dpi = 300)
```

# UMAP all samples from all people 

```{r}
umap_result <- umap(taxa.all_filt)

# Convert UMAP output to a dataframe
umap_df <- as.data.frame(umap_result$layout)
umap_df$PatientID <- meta.all$`Patient Id`
umap_df$Timepoint <- meta.all$irAE.sample

umap_long <- umap_df %>%
  pivot_longer(cols = c(PatientID, Timepoint), names_to = "Grouping", values_to = "Label")

p1 <- ggplot(umap_df, aes(x = V1, y = V2, color = PatientID)) +
  geom_point(size = 3) +
  stat_ellipse(aes(group = PatientID), linetype = "dashed") +
  theme_minimal() +
  theme(legend.position = "right")

p2 <- ggplot(umap_df, aes(x = V1, y = V2, color = Timepoint)) +
  geom_point(size = 3) +
  stat_ellipse(aes(group = Timepoint), linetype = "dashed") +
  theme_minimal() +
  theme(legend.position = "right") 

# Combine plots side by side
p.all <- p1 + p2 
p.all

ggsave("../figures/exploratory-umap-all.png", p.all, width = 12, height = 6, dpi = 300)
```


# PCA irAE samples only

```{r}
pca_result <- prcomp(taxa.irae_filt, scale. = TRUE)

pca_df <- as.data.frame(pca_result$x[, 1:2])
colnames(pca_df) <- c("PC1", "PC2")

pca_df$PatientID <- meta.irae$`Patient Id`
pca_df$Timepoint <- meta.irae$irAE.sample

# Plot colored by PatientID
p1.pca <- ggplot(pca_df, aes(x = PC1, y = PC2, color = PatientID)) +
  geom_point(size = 3) +
  stat_ellipse(aes(group = PatientID), linetype = "dashed") +
  theme_minimal() +
  theme(legend.position = "right")

# Plot colored by Timepoint (e.g., "irAE", "Other")
p2.pca <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Timepoint)) +
  geom_point(size = 3) +
  stat_ellipse(aes(group = Timepoint), linetype = "dashed") +
  theme_minimal() +
  theme(legend.position = "right")

# Combine plots side by side
pca.plot <- p1.pca + p2.pca
pca.plot

ggsave("../figures/exploratory-outlier-pca-irAE.png", pca.plot, width = 12, height = 6, dpi = 300)
```


PCA outlier ? 
Try removing? Whats going on with this sample



```{r}
pca_result <- prcomp(taxa.irae_filt, center = TRUE, scale. = TRUE)
pca_df <- as.data.frame(pca_result$x)
pca_df$sample <- rownames(pca_df)

pca_df <- merge(pca_df, meta, by = "sample")
```

# Plot 

```{r, warning=FALSE, message=FALSE}
ggplot(pca_df, aes(x = PC1, y = PC2, color = `Patient Id`)) + 
  geom_point(size = 3) + 
  theme_minimal() +
  # remove outlier from visual field 
  #xlim(-10, 5) +
  stat_ellipse(aes(group = `Patient Id`), level = 0.95, linetype = "dashed")
```

# UMAP
```{r}
umap_result <- umap(taxa.irae_filt)

# Convert UMAP output to a dataframe
umap_df <- as.data.frame(umap_result$layout)
umap_df$PersonID <- meta.irae$`Patient Id`
umap_df$Timepoint <- meta.irae$irAE.sample

umap_long <- umap_df %>%
  pivot_longer(cols = c(PersonID, Timepoint), names_to = "Grouping", values_to = "Label")

# ggplot(umap_long, aes(x = V1, y = V2, color = Label)) +
#   geom_point(size = 3) +
#   theme_minimal() +
#   facet_wrap(~Grouping) +
#   theme(legend.position = "bottom")

p1 <- ggplot(umap_df, aes(x = V1, y = V2, color = PersonID)) +
  geom_point(size = 3) +
  theme_minimal() +
  theme(legend.position = "right")

p2 <- ggplot(umap_df, aes(x = V1, y = V2, color = Timepoint)) +
  geom_point(size = 3) +
  theme_minimal() +
  theme(legend.position = "right")

# Combine plots side by side
p1 + p2 
```

One major outlier M018, Sample FF08034405. This sample seems to be missing a lot
of taxa present in others. 


# Heirarchical
```{r}
# Compute Aitchison distance (Euclidean on CLR data)
dist_matrix <- dist(clr_taxa_filt_t, method = "euclidean")
hc <- hclust(dist_matrix, method = "ward.D2")
plot(hc, main = "Hierarchical Clustering")

# Convert hclust object to a dendrogram
dend <- as.dendrogram(hc)

# Assuming you have a vector of patient IDs (e.g., "patient_id") corresponding to samples
# Example: patient_id <- c("Patient1", "Patient2", "Patient1", "Patient3", "Patient2")
patient_id <- meta$`Patient Id` # Replace with your actual patient ID data

patient_colors <- as.factor(patient_id)
palette_colors <- rainbow(length(unique(patient_colors)))

# Color the labels with distinct colors for each patient
dend <- dend %>%
  set("labels_col", palette_colors[patient_colors]) %>%
  set("labels_cex", 0.8)  # Adjust label size if needed

# Plot the dendrogram with colored labels
plot(dend, main = "Hierarchical Clustering by Patient")
```

# K means 
```{r}
set.seed(123)
k <- 3 # Set number of clusters
kmeans_result <- kmeans(clr_taxa_filt_t, centers = k)

# Add cluster labels to metadata
meta$Cluster <- as.factor(kmeans_result$cluster)

# PCA Plot with clusters
ggplot(pca_df, aes(x = PC1, y = PC2, color = meta$Cluster)) + 
  geom_point(size = 3) + 
  theme_minimal() + 
  ggtitle("K-means Clustering of Microbiome Samples")
```
M070 ends up in their own cluster with k=3 and k=2...


```{r}
umap_result <- umap(clr_taxa_filt_t)

# Convert UMAP result to a data frame for easy plotting
umap_df <- as.data.frame(umap_result$layout)
umap_df$Cluster <- as.factor(meta$Cluster)  # Add your cluster or grouping variable here

# Plot UMAP with ellipses for each cluster
ggplot(umap_df, aes(x = V1, y = V2, color = Cluster)) + 
  geom_point(size = 3) + 
  stat_ellipse(type = "norm", size = 1) +  # Add ellipses based on normal distribution
  theme_minimal() + 
  ggtitle("UMAP of CLR-Transformed Microbiome Data with Clusters and Ellipses")
```


# Permanova
```{r}
library(vegan)

# Compute Aitchison distance
dist_matrix <- dist(clr_data, method = "euclidean")

# Run PERMANOVA (testing clustering by PersonID)
permanova_result <- adonis2(dist_matrix ~ metadata$PersonID, data = metadata)

# Print results
print(permanova_result)

```


