---
title: "longitudinal vis"
author: "C. Dravillas"
date: "2025-03-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggplot2)
```

# Load data
```{r}
load("../manuscript/data/clinical.rds")
load("../manuscript/data/microbiome_metadata.rds")
load("../manuscript/data/ra-long.rds")
load("../manuscript/data/counts-long.rds")
```

```{r}
vis.in <- ra.l %>%
  left_join(mic.meta) %>%
  left_join(demographics) %>%
  # create binary outcome var
  mutate(response.bin = ifelse(Response == "Responder", 1, 0),
  # make days numeric
         Days = as.integer(Days),
  # create quadratic days 
         Days2 = Days^2) %>%
  rename(Patient_Id = `Patient Id`) 

vis.in.c <- counts.l %>%
  mutate(microbe = str_extract(Taxonomy, "[^|]+$")) %>%
  dplyr::select(-Taxonomy) %>%
  pivot_wider(names_from = microbe, values_from = est.reads, values_fill = 0) %>%
  pivot_longer(!sample, names_to = "microbe", values_to = "est.reads") %>%
  left_join(mic.meta) %>%
  left_join(demographics) %>%
  # create binary outcome var
  mutate(response.bin = ifelse(Response == "Responder", 1, 0),
  # make days numeric
         Days = as.integer(Days),
  # create quadratic days 
         Days2 = Days^2) %>%
  rename(Patient_Id = `Patient Id`) 
```

Let's look at some taxa of note from the model output:
response ~ microbe*Days + (1 | Patient_Id)
s__Sellimonas_intestinalis - pval 0.000000e+00
s__Bacteroides_ovatus - pval 1.316880e-188

response ~ microbe*(Days + Days2) + (1 | Patient_Id)
s__Bifidobacterium_dentium - pval 0.000000e+00
s__Blautia_glucerasea - pval 7.128973e-94

microbe ~ response*Days + (1 | Patient_Id)
s__Sutterella_SGB9281
s__Clostridium_perfringens

w/o padj
s__Sutterella
s__Lachnospira_sp_NSJ_43 - pval 0.03
g__Roseburia - pval 0.0453133

microbe ~ response*(Days+Days2) + (1 | Patient_Id)
g__Fusobacterium - pval 1.007218e-10
s__Clostridium_perfringens - pval 0.018

```{r}
mics <- c("s__Sellimonas_intestinalis",
"s__Bacteroides_ovatus",
"s__Bifidobacterium_dentium",
"s__Bifidobacterium_pseudolongum",
"s__Blautia_glucerasea",
"s__Sutterella_SGB9281",
"s__Lachnospira_sp_NSJ_43" ,
"g__Roseburia",
"g__Fusobacterium",
"s__Clostridium_perfringens",
"g__Dialister") 

mics <- c("s__Klebsiella_pneumoniae",
          "f__FGB1380",
          "s__Veillonella_atypica",
          "s__Mediterraneibacter_butyricigenes",
          "s__Blautia_stercoris",
          "s__Bifidobacterium_dentium")
```

```{r}
vis.in %>% 
  filter(microbe %in% mics) %>%
  ggplot(aes(x=Days, y=RelAbun, group = Patient_Id, color=Response, alpha = 0.5)) +
  geom_line() +
  facet_wrap(~microbe, scales = "free") +
  theme_minimal() 
```

```{r}
vis.in.c %>% 
  filter(microbe %in% mics) %>%
  ggplot(aes(x=Days, y=est.reads, group = Patient_Id, color=Response, alpha = 0.5)) +
  geom_line() +
  facet_wrap(~microbe, scales = "free") +
  theme_minimal() 
```

```{r}
mics <- c(
  "s__Oscillospiraceae_unclassified_SGB15256",
  "s__Segatella_copri",
  "s__Blautia_producta",
  "s__Clostridium_perfringens",
  "s__Ruminococcus_gnavus",
  "s__Blautia_luti"
)

vis.in.c %>% 
  filter(microbe %in% mics) %>%
  ggplot(aes(x=Days, y=est.reads, group = Patient_Id, color=Response, alpha = 0.5)) +
  geom_line() +
  facet_wrap(~microbe, scales = "free") +
  theme_minimal() 
```



```{r}
vis.in %>%
  filter(grepl("Blautia", microbe)) %>%
  ggplot(aes(x=Days, y=RelAbun, group = Patient_Id, color=Response, alpha = 0.5)) +
  geom_line() +
  facet_wrap(~microbe, scales = "free") +
  theme_minimal() 
```


