---
title: "Frequentist vs. Bayesian"
author: "C. Dravillas"
date: "2025-03-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lme4)
library(broom.mixed)
library(rlist)
library(NBZIMM)
library(nlme)
library(BhGLM)
library(Metrics)
library(glmmTMB)
```

# Load data
```{r}
load("../manuscript/data/clinical.rds")
load("../manuscript/data/microbiome_metadata.rds")
load("../manuscript/data/counts-long.rds")
load("../manuscript/data/ra-long.rds")
```

# Prepare data

```{r}
# Let's create a pre-filter for microbes present in less than 4 samples 
pres <- counts.l %>%
  filter(sample %in% mic.meta$sample) %>%
  group_by(Taxonomy) %>%
  summarize(pres = sum(est.reads > 0)) %>%
  mutate(microbe = str_extract(Taxonomy, "[^|]+$")) %>%
  dplyr::select(-Taxonomy) 

model.in <- counts.l %>%
  mutate(microbe = str_extract(Taxonomy, "[^|]+$")) %>%
  dplyr::select(-Taxonomy) %>%
  right_join(mic.meta) %>%
  left_join(demographics) %>%
  # create binary outcome var
  mutate(response.bin = ifelse(Response == "Responder", 1, 0),
  # make days numeric
         Days = as.integer(Days)) %>%
  rename(Patient_Id = `Patient Id`) %>%
  filter(!grepl("t__", microbe)) %>%
  pivot_wider(names_from = microbe, values_from = est.reads, values_fill = 0) %>%
  filter(sample %in% mic.meta$sample)

model.in.ra <- ra.l %>%
  right_join(mic.meta) %>%
  left_join(demographics) %>%
  # create binary outcome var
  mutate(response.bin = ifelse(Response == "Responder", 1, 0),
  # make days numeric
         Days = as.integer(Days)) %>%
  rename(Patient_Id = `Patient Id`) %>%
  filter(!grepl("t__", microbe)) %>%
  pivot_wider(names_from = microbe, values_from = RelAbun, values_fill = 0) %>%
  filter(sample %in% mic.meta$sample)

lib.size <- counts.l %>%
  filter(Taxonomy %in% c("k__Bacteria","k__Archaea")) %>%
  group_by(sample) %>%
  summarize(library.size = sum(est.reads))

model.in <- model.in %>% 
  left_join(lib.size) 

microbes <- colnames(model.in)
microbes <- microbes[grepl("__", microbes)]
microbes.ra <- colnames(model.in.ra)
microbes.ra <- microbes.ra[grepl("__", microbes.ra)]
```

Compare models on small test set 

```{r}
test.mics <- c("s__Alistipes_finegoldii",  "s__Lachnospiraceae_bacterium_AM48_27BH", "s__Intestinimonas_massiliensis","f__Barnesiellaceae", "s__Phascolarctobacterium_faecium", "g__Dialister", "s__Bifidobacterium_longum")
```

```{r}
# Function to fit models
fit_models <- function(x, model.in, model.in.ra) {
  model_formula_off <- as.formula(paste0(x, " ~ Response*Days + offset(log(library.size))"))
  model_formula <- as.formula(paste0(x, " ~ Response*Days"))

  models <- list(
    # FZINBMM (Counts)
    "NB" = glmm.zinb(model_formula_off, data = model.in, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL),
    # FZINBMM w/ AR1
    "AR_NB" = glmm.zinb(model_formula_off, data = model.in, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL, correlation = corAR1()),
    # FZINBMM w/ no offset
    "noOff_NB" = glmm.zinb(model_formula, data = model.in, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL),
    # ZIGMM (RA)
    "Gaussian" = lme.zig(model_formula, data = model.in.ra, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL),
    # ZIGMM w/ AR1
    "AR_Gaussian" = lme.zig(model_formula, data = model.in.ra, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL, correlation = corAR1()),
    # ZIGMM w/ CAR1
    "CAR_Gaussian" = lme.zig(model_formula, data = model.in.ra, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL, correlation = corCAR1()),
    # ZIGMM w/ ARMA
    "ARMA_Gaussian" = lme.zig(model_formula, data = model.in.ra, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL, correlation = corARMA(p = 1, q = 1))
  )

  return(models)
}

# Compute model metrics
compute_metrics <- function(model, observed) {
  # mse = mean of (error)^2
  predictions <- predict(model, type = "response")
  residuals_sq <- (predictions - observed)^2
  mse <- mean(residuals_sq, na.rm = TRUE)
  rmse <- sqrt(mse)

  # define n and p 
  n <- length(observed)
  p <- length(coef(model))
  
  # variance of residuals
  sigma2 <- sum(residuals(model)^2) / (n - p)
  
  # Loglik 
  logLik <- -(n/2) * (log(2*pi*sigma2) + 1) - sum(residuals(model)^2) / (2 * sigma2)

  # number of estimate params..i think k should be 6 ?
  # + random intercept and zi intercept 
  k <- length(coef(model)) + 2
  aic <- (2 * k) - (2 * logLik)
  bic <- (log(n) * k) - (2 * logLik)

  return(c(MSE = mse, RMSE = rmse, AIC = aic, BIC = bic, LogLik = logLik))
}

# Loop over microbes
results <- list()
for (x in test.mics) {
  models <- fit_models(x, model.in, model.in.ra)

  model_metrics <- lapply(names(models), function(model_name) {
    # Use the correct observed dataset
    observed <- if (grepl("Gaussian", model_name)) model.in.ra[[x]] else model.in[[x]]
    compute_metrics(models[[model_name]], observed)
  })

  model_metrics <- do.call(rbind, model_metrics)
  rownames(model_metrics) <- names(models)

  results[[x]] <- model_metrics
  print(paste("Microbe:", x))
  print(model_metrics)
}

# Combine all results into a single data frame
final_results <- do.call(rbind, lapply(names(results), function(m) {
  cbind(Microbe = m, results[[m]])
}))

final_results <- as.data.frame(final_results)
final_results[, -1] <- lapply(final_results[, -1], as.double)

print(final_results)
```
save model comparison output
```{r}
write.csv(final_results, file = "../data/modeling-comparison-metrics.csv")
```

Examine Results
- based on MSE we can rule out NB models
  - we do expect a large MSE due to offset and using counts, but these MSE's are
    larger than the range of counts 
- Next, look at AIC, BIC and LogLik
- Gaussian (RA) with no specified corr struct is the clear winner
- NB has better LogLik fo two mics but by very little and still has crazy MSE

Implementation of ZIGMM with RA and no corr struct 

```{r}
capture.ZIGMM <- function(modin.dat, microbes, zero_threshold = 0.1) {
  mods.list <- list()
  zero_prop <- sapply(microbes, function(x) mean(modin.dat[[x]] == 0, na.rm = TRUE))
  
  for (x in microbes) {
    tryCatch({
      model_formula <- as.formula(paste0(x, " ~ Response*Days"))
      
        if (zero_prop[x] >= zero_threshold) {
        # Use Zero-Inflated Gaussian Model (ZIGMM)
        model <- lme.zig(model_formula, 
                           data = modin.dat, 
                           random = ~ 1|Patient_Id, 
                           zi_fixed = ~1, 
                           zi_random = NULL
                           )
        model_type <- "ZIGMM"
      } else {
        # Use Gaussian Model (GMM)
        model <- lme(model_formula, 
                         data = modin.dat, 
                         random = ~ 1|Patient_Id
                         )
        model_type <- "GMM"
      }
      
      model_results <- broom.mixed::tidy(model)
      model_results <- model_results %>% mutate(microbe = x, model_type = model_type)
      
      mods.list[[x]] <- model_results
      
      
    }, error = function(e) {
      message("Model failed for ", x, ": ", e$message)
      mods.list[[x]] <- NULL  # Assign NULL to failed models
    })
  }
  
  mods.list.clean <- mods.list[!sapply(mods.list, is.null)]
  
  if (length(mods.list.clean) > 0) {
    mods.df <- bind_rows(mods.list.clean)
  } else {
    mods.df <- data.frame() 
  }
  
  return(mods.df)
}

gmm.out <- capture.ZIGMM(model.in.ra, microbes.ra)
gmm.out <- gmm.out %>%
  filter(term == "ResponseResponder:Days") %>%
  mutate(padj = p.adjust(p.value, method = "bonferroni")) %>%
  left_join(pres)
```

save model output
```{r}
write.csv(gmm.out, file = "../data/modeling-gaussian.csv", row.names = F)
```


Errors include 
Model failed for s__Streptococcus_agalactiae: system is computationally singular: reciprocal condition number = 8.59737e-19
Model failed for g__GGB3817: Overfitted model!

Warning: Variance function structure of classvarFixedwith no parameters, or uninitialized

Much less significant and all unknown taxa 



Previous work... 

Zero-inflated Negative Binomial mixed models (ZINBMMs) annd NBMMs
using counts

<!-- Using ZINBMM for taxa with >= 10% zeroes  -->
<!-- Using NBMM for taxa with <10% zeroes  -->
<!-- ```{r} -->
<!-- capture.ZINBMM <- function(modin.dat, microbes, zero_threshold = 0.1) { -->
<!--   mods.list <- list() -->
<!--   zero_prop <- sapply(microbes, function(x) mean(modin.dat[[x]] == 0, na.rm = TRUE)) -->

<!--   for (x in microbes) { -->
<!--     tryCatch({ -->
<!--       model_formula <- as.formula(paste0(x, " ~ Response*Days + offset(log(library.size))")) -->

<!--         if (zero_prop[x] >= zero_threshold) { -->
<!--         # Use Zero-Inflated Negative Binomial Model (ZINBMM) -->
<!--         model <- glmm.zinb(model_formula,  -->
<!--                            data = modin.dat,  -->
<!--                            random = ~ 1|Patient_Id,  -->
<!--                            zi_fixed = ~1,  -->
<!--                            zi_random = NULL, -->
<!--                            correlation = corAR1() -->
<!--                            ) -->
<!--         model_type <- "ZINBMM" -->
<!--       } else { -->
<!--         # Use Negative Binomial Model (NBMM) -->
<!--         model <- glmm.nb(model_formula,  -->
<!--                          data = modin.dat,  -->
<!--                          random = ~ 1|Patient_Id, -->
<!--                          correlation = corAR1() -->
<!--                          ) -->
<!--         model_type <- "NBMM" -->
<!--       } -->

<!--       model_results <- broom.mixed::tidy(model) -->
<!--       model_results <- model_results %>% mutate(microbe = x, model_type = model_type) -->

<!--       mods.list[[x]] <- model_results -->


<!--     }, error = function(e) { -->
<!--       message("Model failed for ", x, ": ", e$message) -->
<!--       mods.list[[x]] <- NULL  # Assign NULL to failed models -->
<!--     }) -->
<!--   } -->

<!--   mods.list.clean <- mods.list[!sapply(mods.list, is.null)] -->

<!--   if (length(mods.list.clean) > 0) { -->
<!--     mods.df <- bind_rows(mods.list.clean) -->
<!--   } else { -->
<!--     mods.df <- data.frame()  -->
<!--   } -->

<!--   return(mods.df) -->
<!-- } -->

<!-- nbmm.out <- capture.ZINBMM(model.in, microbes) -->
<!-- nbmm.out <- nbmm.out %>% -->
<!--   filter(term == "ResponseResponder:Days") %>% -->
<!--   mutate(padj = p.adjust(p.value, method = "bonferroni")) %>% -->
<!--   left_join(pres) -->

<!-- nbmm.ar1.out <- capture.ZINBMM(model.in, microbes) -->
<!-- nbmm.ar1.out <- nbmm.ar1.out %>% -->
<!--   filter(term == "ResponseResponder:Days") %>% -->
<!--   mutate(padj = p.adjust(p.value, method = "bonferroni")) %>% -->
<!--   left_join(pres) -->

<!-- # # test with one taxa -->
<!-- # mod = glmm.zinb(`s__Blautia_wexlerae` ~ Response + Days + offset(log(library.size)), data = model.in,  -->
<!-- #                 random = ~ 1|Patient_Id, zi_fixed = ~1, zi_random =NULL) -->
<!-- #  -->
<!-- # mod.out <- broom.mixed::tidy(mod) -->
<!-- ``` -->
