---
title: "longitudinal vis"
author: "C. Dravillas"
date: "2025-03-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggplot2)
```

# Load data
```{r}
load("../manuscript/data/clinical.rds")
load("../manuscript/data/microbiome_metadata.rds")
load("../manuscript/data/ra-long.rds")
```

```{r}
vis.in <- ra.l %>%
  left_join(mic.meta) %>%
  left_join(demographics) %>%
  # create binary outcome var
  mutate(response.bin = ifelse(Response == "Responder", 1, 0),
  # make days numeric
         Days = as.integer(Days),
  # create quadratic days 
         Days2 = Days^2) %>%
  rename(Patient_Id = `Patient Id`) 
```

Let's look at some taxa of note from the model output:
response ~ microbe*Days + (1 | Patient_Id)
s__Sellimonas_intestinalis - pval 0.000000e+00
s__Bacteroides_ovatus - pval 1.316880e-188

response ~ microbe*(Days + Days2) + (1 | Patient_Id)
s__Bifidobacterium_dentium - pval 0.000000e+00
s__Blautia_glucerasea - pval 7.128973e-94

microbe ~ response*Days + (1 | Patient_Id)
s__Sutterella
s__Lachnospira_sp_NSJ_43 - pval 0.03
g__Roseburia - pval 0.0453133

microbe ~ response*(Days+Days2) + (1 | Patient_Id)
g__Fusobacterium - pval 1.007218e-10
s__Clostridium_perfringens - pval 0.018

