---
title: "venn diagram"
author: "C. Dravillas"
date: "2025-05-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(VennDiagram)
```

Check overlap between the various model results 

4 groups total 
ancom-bc vs longitudinal
responders and non-responders 

read in necessary results 
```{r}
load("../manuscript/data/ANCOM-BC2_response_baseline.RData")
load("../manuscript/data/ZIGMM-model-output.rds")
```

get microbe sets 
- should also check/include structural zeroes 
- what term do we want to look at ?
  - if interested in how the two models perform, we could just look at response
  - this term is dif between groups (at baseline)
  - but first we will look at the interaction term since that's what i have saved
  - baseline, linear, quadratic terms 

```{r}
ancom.r <- (res.recist %>%
  filter(q_ResponseResponder < 0.05,
         lfc_ResponseResponder > 0))$taxon
ancom.nr <- (res.recist %>%
  filter(q_ResponseResponder < 0.05,
         lfc_ResponseResponder < 0))$taxon

# almost no point in this, only one value...this is telling 
# base.r <- (gmm.out %>%
#   filter(term == "ResponseResponder",
#          padj < 0.05,
#          estimate > 0))$microbe 
# base.nr <- (gmm.out %>%
#   filter(term == "ResponseResponder",
#          padj < 0.05,
#          estimate < 0))$microbe 
# what if we look at unadjusted p-value instead
base.r <- (gmm.out %>%
  filter(term == "ResponseResponder",
         p.value < 0.05,
         estimate > 0))$microbe 
base.nr <- (gmm.out %>%
  filter(term == "ResponseResponder",
         p.value < 0.05,
         estimate < 0))$microbe 

lin.r <- (gmm.out %>%
  filter(term == "ResponseResponder:Days",
         padj < 0.05,
         estimate > 0))$microbe 
lin.nr <- (gmm.out %>%
  filter(term == "ResponseResponder:Days",
         padj < 0.05,
         estimate < 0))$microbe 

quad.r <- (gmm.out %>%
  filter(term == "ResponseResponder:I(Days^2)",
         padj < 0.05,
         estimate > 0))$microbe 
quad.nr <- (gmm.out %>%
  filter(term == "ResponseResponder:I(Days^2)",
         padj < 0.05,
         estimate < 0))$microbe 
```


Examine overlap
```{r}

```

