---
title: "irAE longitudinal models"
author: "C. Dravillas"
date: "2025-04-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(broom.mixed)
library(rlist)
library(NBZIMM)
```

```{r}
load("../manuscript/data/clinical.rds")
load("../manuscript/data/microbiome_metadata.rds")
load("../manuscript/data/counts-long.rds")
load("../manuscript/data/ra-long.rds")
load("../manuscript/data/events.rds")
```

From last meeting 
Irae model
- NO time
- Use the same gaussian model 
- Also try irae ~ microbe + (1|ptid) 


some other potential ideas

irAE as response var, with irAE changing over time, so pt becomes irAE yes after 
having an irAE? Forever or for some amount of time?

Could also look for known signatures? 



# Prepare data 
- lets say irAE yes if within 10 days so same as clustering def

```{r}
# define time points 
irAE.time <- events %>%
  filter(Event == "irAE") %>%
  mutate(irAE.day = Days) %>%
  dplyr::select(`Patient Id`, irAE.day) 

meta.irae <- mic.meta %>%
  left_join(demographics) %>%
  left_join(irAE.time) %>%
  mutate(irAE.day = ifelse(irAE.day < 0 | irAE.day > 100, NA, irAE.day),
         irAE.sample = ifelse(is.na(irAE.day), 0,
                       ifelse(abs(Days - irAE.day) <= 10, 1, 0))) %>%
  group_by(sample) %>%
  slice_max(irAE.sample, n = 1, with_ties = FALSE) %>%
  ungroup() 

pres <- counts.l %>%
  filter(sample %in% mic.meta$sample) %>%
  group_by(Taxonomy) %>%
  summarize(pres = sum(est.reads > 0)) %>%
  mutate(microbe = str_extract(Taxonomy, "[^|]+$")) %>%
  dplyr::select(-Taxonomy)

model.in <- counts.l %>%
  mutate(microbe = str_extract(Taxonomy, "[^|]+$")) %>%
  dplyr::select(-Taxonomy) %>%
  full_join(meta.irae) %>%
  # create binary outcome var
  mutate(Days = as.integer(Days)) %>%
  rename(Patient_Id = `Patient Id`) %>%
  filter(!grepl("t__", microbe)) %>%
  pivot_wider(names_from = microbe, values_from = est.reads, values_fill = 0) %>%
  filter(sample %in% mic.meta$sample)

lib.size <- counts.l %>%
  filter(Taxonomy %in% c("k__Bacteria","k__Archaea")) %>%
  group_by(sample) %>%
  summarize(library.size = sum(est.reads))

model.in <- model.in %>%
  left_join(lib.size)

model.in.ra <- ra.l %>%
  full_join(meta.irae) %>% 
  # create binary outcome var
  mutate(Days = as.integer(Days)) %>%
  rename(Patient_Id = `Patient Id`) %>%
  filter(!grepl("t__", microbe)) %>%
  pivot_wider(names_from = microbe, values_from = RelAbun, values_fill = 0) %>%
  filter(sample %in% mic.meta$sample)

microbe_mat <- ra.l %>%
  filter(!grepl("t__", microbe)) %>%
  pivot_wider(names_from = microbe, values_from = RelAbun, values_fill = 0) %>%
  filter(sample %in% mic.meta$sample) %>%
  column_to_rownames("sample") %>%
  as.matrix()

# add pseudo count
microbe_mat_pseudo <- microbe_mat + 1e-6
# CLR transform 
clr_microbe_mat <- compositions::clr(microbe_mat_pseudo)

# combine with metadata
model.in.ra.clr <- as.data.frame(clr_microbe_mat) %>%
  rownames_to_column("sample") %>%
  full_join(meta.irae) %>% 
  # create binary outcome var
  mutate(Days = as.integer(Days)) %>%
  rename(Patient_Id = `Patient Id`)

microbes <- colnames(model.in)
microbes <- microbes[grepl("__", microbes)]
microbes.ra <- colnames(model.in.ra)
microbes.ra <- microbes.ra[grepl("__", microbes.ra)]
```

# Select dif test microbes for irAE modeling 

```{r}
test.mics <- c("f__Muribaculaceae",  "s__Bacteroides_pectinophilus", "s__Enterobacter_roggenkampii", "s__Pseudoflavonifractor_gallinarum", "g__Fournierella", "s__Blautia_sp_MCC283")
```


microbe ~ irAE + (1|id)
Need to add try catch for model convergence 
```{r}
# Fit a list of models for each microbe
fit_models <- function(x, model.in, model.in.ra) {
  # Make sure x is a column in the datasets
  model.in[[x]] <- as.numeric(model.in[[x]])
  model.in.ra[[x]] <- as.numeric(model.in.ra[[x]])

  # Formulas where microbe counts are outcome
  model_formula_off <- as.formula(paste0(x, " ~ irAE.sample + offset(log(library.size))"))
  model_formula <- as.formula(paste0(x, " ~ irAE.sample"))

  models <- list(
    "ZINBMM_Offset" = tryCatch(
      glmm.zinb(model_formula_off, data = model.in, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL),
      error = function(e) NULL
    ),
    "ZINBMM_AR1" = tryCatch(
      glmm.zinb(model_formula_off, data = model.in, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL, correlation = corAR1()),
      error = function(e) NULL
    ),
    "ZINBMM" = tryCatch(
      glmm.zinb(model_formula, data = model.in, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL),
      error = function(e) NULL
    ),
    "ZIGMM" = tryCatch(
      lme.zig(model_formula, data = model.in.ra, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL),
      error = function(e) NULL
    ),
    "ZIGMM_AR1" = tryCatch(
      lme.zig(model_formula, data = model.in.ra, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL, correlation = corAR1()),
      error = function(e) NULL
    ),
    "ZIGMM_CAR1" = tryCatch(
      lme.zig(model_formula, data = model.in.ra, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL, correlation = corCAR1()),
      error = function(e) NULL
    ),
    "ZIGMM_ARMA" = tryCatch(
      lme.zig(model_formula, data = model.in.ra, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL, correlation = corARMA(p = 1, q = 1)),
      error = function(e) NULL
    )
  )

  return(models)
}

# Compute metrics (excluding estimate)
compute_metrics <- function(model, observed) {
  if (is.null(model)) {
    return(c(MSE = NA, RMSE = NA, AIC = NA, BIC = NA, LogLik = NA))
  }

  predictions <- tryCatch(predict(model, type = "response"), error = function(e) NA)
  residuals_sq <- (predictions - observed)^2
  mse <- mean(residuals_sq, na.rm = TRUE)
  rmse <- sqrt(mse)

  n <- length(observed)
  p <- length(coef(model))
  sigma2 <- sum(residuals(model)^2, na.rm = TRUE) / (n - p)
  logLik <- -(n/2) * (log(2 * pi * sigma2) + 1) - sum(residuals(model)^2, na.rm = TRUE) / (2 * sigma2)
  k <- length(coef(model)) + 2

  aic <- (2 * k) - (2 * logLik)
  bic <- (log(n) * k) - (2 * logLik)

  return(c(MSE = mse, RMSE = rmse, AIC = aic, BIC = bic, LogLik = logLik))
}

# Loop over microbes
results <- list()
for (x in test.mics) {
  models <- fit_models(x, model.in, model.in.ra)

  model_metrics <- lapply(names(models), function(model_name) {
    observed <- if (grepl("GMM", model_name)) model.in.ra[[x]] else model.in[[x]]
    compute_metrics(models[[model_name]], observed)
  })

  model_metrics <- do.call(rbind, model_metrics)
  rownames(model_metrics) <- names(models)

  # Add microbe name
  model_metrics <- cbind(Microbe = x, Model = rownames(model_metrics), model_metrics)
  results[[x]] <- model_metrics

  print(paste("Microbe:", x))
  print(model_metrics)
}

# Combine all results into a single data frame
final_mic <- do.call(rbind, results)
final_mic <- as.data.frame(final_mic)

# Ensure numeric columns
num_cols <- c("MSE", "RMSE", "AIC", "BIC", "LogLik")
final_mic[num_cols] <- lapply(final_mic[num_cols], as.numeric)
```
And ZIGMM with no specified correlation structure is still clear winner 


irAE ~ microbe + (1|id)
With a binary outcome variable we shouldnt use gaussian
Need to instead you glmm with binomial family 
- use RA
- use CLR transformed RA
- use counts with library size offset 

glmmTMB reports model metrics no need to calculate

```{r}
# Function to fit models
fit_models <- function(x, model.in, model.in.ra, model.in.ra.clr) {
  model_formula <- as.formula(paste0("irAE.sample ~ ", x))
  
  models <- list(
    glmmTMB_Counts   = glmmTMB(formula = model_formula, ziformula = ~1, family = binomial, data = model.in),
    glmmTMB_RA       = glmmTMB(formula = model_formula, ziformula = ~1, family = binomial, data = model.in.ra),
    glmmTMB_CLR_RA   = glmmTMB(formula = model_formula, ziformula = ~1, family = binomial, data = model.in.ra.clr)
  )
  
  return(models)
}

# Function to extract model metrics from glmmTMB objects
extract_model_metrics <- function(model) {
  data.frame(
    AIC = AIC(model),
    BIC = BIC(model),
    LogLik = logLik(model)[1]  # extract numeric value
  )
}

# Loop over microbes
results <- list()
for (x in test.mics) {
  models <- fit_models(x, model.in, model.in.ra, model.in.ra.clr)
  
  model_metrics <- lapply(models, extract_model_metrics)
  model_metrics_df <- bind_rows(model_metrics, .id = "Model")
  model_metrics_df$Microbe <- x
  
  results[[x]] <- model_metrics_df
  print(paste("Microbe:", x))
  print(model_metrics_df)
}

final_irAE <- bind_rows(results)
```

looks like RA is best, def not counts 
