---
title: "Frequentist vs. Bayesian"
author: "C. Dravillas"
date: "2025-03-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lme4)
library(broom.mixed)
library(rlist)
library(NBZIMM)
library(nlme)
library(BhGLM)
library(Metrics)
library(glmmTMB)
```

# Load data
```{r}
load("../manuscript/data/clinical.rds")
load("../manuscript/data/microbiome_metadata.rds")
load("../manuscript/data/counts-long.rds")
```

# Prepare data

```{r}
# Let's create a pre-filter for microbes present in less than 4 samples 
pres <- counts.l %>%
  filter(sample %in% mic.meta$sample) %>%
  group_by(Taxonomy) %>%
  summarize(pres = sum(est.reads > 0)) %>%
  mutate(microbe = str_extract(Taxonomy, "[^|]+$")) %>%
  dplyr::select(-Taxonomy) 

model.in <- counts.l %>%
  mutate(microbe = str_extract(Taxonomy, "[^|]+$")) %>%
  dplyr::select(-Taxonomy) %>%
  right_join(mic.meta) %>%
  left_join(demographics) %>%
  # create binary outcome var
  mutate(response.bin = ifelse(Response == "Responder", 1, 0),
  # make days numeric
         Days = as.integer(Days)) %>%
  rename(Patient_Id = `Patient Id`) %>%
  filter(!grepl("t__", microbe)) %>%
  pivot_wider(names_from = microbe, values_from = est.reads, values_fill = 0) %>%
  filter(sample %in% mic.meta$sample)

lib.size <- counts.l %>%
  filter(Taxonomy %in% c("k__Bacteria","k__Archaea")) %>%
  group_by(sample) %>%
  summarize(library.size = sum(est.reads))

model.in <- model.in %>% 
  left_join(lib.size) 

microbes <- colnames(model.in)
microbes <- microbes[grepl("__", microbes)]
```

Zero-inflated Negative Binomial mixed models (ZINBMMs) annd NBMMs
using counts

Using ZINBMM for taxa with >= 10% zeroes 
Using NBMM for taxa with <10% zeroes 
```{r}
capture.ZINBMM <- function(modin.dat, microbes, zero_threshold = 0.1) {
  mods.list <- list()
  zero_prop <- sapply(microbes, function(x) mean(modin.dat[[x]] == 0, na.rm = TRUE))
  
  for (x in microbes) {
    tryCatch({
      model_formula <- as.formula(paste0(x, " ~ Response*Days + offset(log(library.size))"))
      
        if (zero_prop[x] >= zero_threshold) {
        # Use Zero-Inflated Negative Binomial Model (ZINBMM)
        model <- glmm.zinb(model_formula, 
                           data = modin.dat, 
                           random = ~ 1|Patient_Id, 
                           zi_fixed = ~1, 
                           zi_random = NULL,
                           correlation = corAR1()
                           )
        model_type <- "ZINBMM"
      } else {
        # Use Negative Binomial Model (NBMM)
        model <- glmm.nb(model_formula, 
                         data = modin.dat, 
                         random = ~ 1|Patient_Id,
                         correlation = corAR1()
                         )
        model_type <- "NBMM"
      }
      
      model_results <- broom.mixed::tidy(model)
      model_results <- model_results %>% mutate(microbe = x, model_type = model_type)
      
      mods.list[[x]] <- model_results
      
      
    }, error = function(e) {
      message("Model failed for ", x, ": ", e$message)
      mods.list[[x]] <- NULL  # Assign NULL to failed models
    })
  }
  
  mods.list.clean <- mods.list[!sapply(mods.list, is.null)]
  
  if (length(mods.list.clean) > 0) {
    mods.df <- bind_rows(mods.list.clean)
  } else {
    mods.df <- data.frame() 
  }
  
  return(mods.df)
}

nbmm.out <- capture.ZINBMM(model.in, microbes)
nbmm.out <- nbmm.out %>%
  filter(term == "ResponseResponder:Days") %>%
  mutate(padj = p.adjust(p.value, method = "bonferroni")) %>%
  left_join(pres)

nbmm.ar1.out <- capture.ZINBMM(model.in, microbes)
nbmm.ar1.out <- nbmm.ar1.out %>%
  filter(term == "ResponseResponder:Days") %>%
  mutate(padj = p.adjust(p.value, method = "bonferroni")) %>%
  left_join(pres)

# # test with one taxa
# mod = glmm.zinb(`s__Blautia_wexlerae` ~ Response + Days + offset(log(library.size)), data = model.in, 
#                 random = ~ 1|Patient_Id, zi_fixed = ~1, zi_random =NULL)
# 
# mod.out <- broom.mixed::tidy(mod)
```


Examine and compare

```{r}
nb.top <- nb.res %>%
  arrange(desc(pres)) %>%
  top_n(10) %>%
  filter()

ar.top <- ar.res  %>%
  arrange(desc(pres)) %>%
  top_n(10)

test.mics <- c("s__Alistipes_finegoldii",  "s__Lachnospiraceae_bacterium_AM48_27BH", "s__Intestinimonas_massiliensis","f__Barnesiellaceae", "s__Phascolarctobacterium_faecium", "g__Dialister", "s__Bifidobacterium_longum")
```

```{r}
for (x in test.mics) {
  model_formula <- as.formula(paste0(x, " ~ Response*Days + offset(log(library.size))"))

  nb.model <- glmm.zinb(model_formula, 
                           data = model.in, 
                           random = ~ 1|Patient_Id, 
                           zi_fixed = ~1, 
                           zi_random = NULL)

  ar.model <- glmm.zinb(model_formula, 
                           data = model.in, 
                           random = ~ 1|Patient_Id, 
                           zi_fixed = ~1, 
                           zi_random =  NULL,
                           correlation = corAR1())
  
  g_model <- lme.zig(model_formula, 
                     data = model.in,
                     random = ~ 1|Patient_Id,
                     zi_fixed = ~1, 
                     zi_random =  NULL) 
  
  g_ar_model <- lme.zig(model_formula, 
                     data = model.in,
                     random = ~ 1|Patient_Id,
                     zi_fixed = ~1, 
                     zi_random =  NULL,
                     correlation = corAR1()) 
  
  
  nb_tmb <- glmmTMB(as.formula(paste0(x, " ~ Response*Days + (1|Patient_Id) + offset(log(library.size))")),
                    data = model.in, 
                    family = nbinom2(),   # Negative binomial family
                    ziformula = ~ 1 )      # Zero-inflation part (adjust as needed)


  print(paste("Microbe:", x))
  
  # Compute Mean Squared Error (MSE)
  mse_ar <- mean((predict(ar.model, type = "response") - model.in[[x]])^2, na.rm = TRUE)
  mse_nb <- mean((predict(nb.model, type = "response") - model.in[[x]])^2, na.rm = TRUE)
  mse_tmb <- mean((predict(nb_tmb, type = "response") - model.in[[x]])^2, na.rm = TRUE)
  mse_g <- mean((predict(g_model, type = "response") - model.in[[x]])^2, na.rm = TRUE)
  mse_g_ar <-mean((predict(g_ar_model, type = "response") - model.in[[x]])^2, na.rm = TRUE)
  
  # Compute Root Mean Squared Error (RMSE)
  # rmse_ar <- rmse(model.in[[x]], predict(ar.model, type = "response"))
  # rmse_nb <- rmse(model.in[[x]], predict(nb.model, type = "response"))

  print(c(AR_Model_MSE = mse_ar, NB_Model_MSE = mse_nb, TMB = mse_tmb, G = mse_g, g.ar = mse_g_ar))
  #print(c(AR_Model_RMSE = rmse_ar, NB_Model_RMSE = rmse_nb))

}
```
VERY HIGH MSE...concerning


[1] "Microbe: s__Alistipes_finegoldii"
AR_Model_MSE NB_Model_MSE 
  7567996784   7567996669 
AR_Model_RMSE NB_Model_RMSE 
     86994.23      86994.23 

[1] "Microbe: s__Lachnospiraceae_bacterium_AM48_27BH"
AR_Model_MSE NB_Model_MSE 
     9166635      9166645 
AR_Model_RMSE NB_Model_RMSE 
     3027.645      3027.647 

[1] "Microbe: s__Intestinimonas_massiliensis"
AR_Model_MSE NB_Model_MSE 
    58239366     58239378 
AR_Model_RMSE NB_Model_RMSE 
     7631.472      7631.473 

[1] "Microbe: f__Barnesiellaceae"
AR_Model_MSE NB_Model_MSE 
  5927030776   5927031417 
AR_Model_RMSE NB_Model_RMSE 
     76987.21      76987.22 

[1] "Microbe: s__Phascolarctobacterium_faecium"
AR_Model_MSE NB_Model_MSE 
  1266264137   1266263964 
AR_Model_RMSE NB_Model_RMSE 
     35584.61      35584.60 

[1] "Microbe: g__Dialister"
AR_Model_MSE NB_Model_MSE 
   362496135    362496122 
AR_Model_RMSE NB_Model_RMSE 
     19039.33      19039.33 

[1] "Microbe: s__Bifidobacterium_longum"
AR_Model_MSE NB_Model_MSE 
  3542945763   3542955716 
AR_Model_RMSE NB_Model_RMSE 
     59522.65      59522.73 

Zero-inflated Gaussian mixed models (ZIGMMs) use RA
```{r}
rm(list = ls())

data(Romero)
names(Romero)

otu = Romero$OTU; dim(otu)
sam = Romero$SampleData; dim(sam)
colnames(sam)

N = sam[, "Total.Read.Counts"]  # total reads
preg = sam$pregnant; table(preg)
subject = sam[, "Subect_ID"]; table(subject)

non = nonzero(y = otu, total = N, plot = F)
nonzero.p = non[[1]]

N = sam[, "Total.Read.Counts"]        
Days = sam$GA_Days; Days = scale(Days)
Age = sam$Age; Age = scale(Age)
Race = sam$Race
preg = sam$pregnant; table(preg)
subject = sam[, "Subect_ID"]; table(subject)

y = otu[, 1]

y0 = log(y+1)
data = data.frame(y0=y0, Days=Days, Age=Age, Race=Race, preg=preg, N=N, subject=subject)
f1 = lme.zig(fixed = y0 ~ Days + Age + Race + preg + offset(log(N)), 
            random = ~ 1 | subject, data = data) 
summary(f1)
fixed(f1)
summary(f1$fit.zero)


f2 = mms(y = log(Romero$OTU+1), fixed = ~ GA_Days + Age + Race + pregnant + 
           offset(log(Total.Read.Counts)), data = Romero$SampleData,
         random = ~ 1 | subject, min.p = 0.2, method = "zig", 
         zi_fixed = ~1, zi_random = NULL)
```


