---
title: "longitudinal models"
author: "C. Dravillas"
date: "2025-02-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lme4)
library(broom.mixed)
library(rlist)
```

# Load data
```{r}
load("../manuscript/data/clinical.rds")
load("../manuscript/data/microbiome_metadata.rds")
load("../manuscript/data/ra-long.rds")
```

# Combine variables
```{r}
model.in <- ra.l %>%
  left_join(mic.meta) %>%
  left_join(demographics) %>%
  # create binary outcome var
  mutate(response.bin = ifelse(Response == "Responder", 1, 0)) %>%
  pivot_wider(names_from = microbe, values_from = RelAbun)

microbes <- unique(ra.l$microbe)
#microbes <- c("s__Blautia_luti", "g__Solobacterium")
microbes <- microbes[!grepl("t__", microbes)]
```

## Modeling

Response ~ microbe*days.since.C1D1 + (1|ptID) 
Response as 0 or 1, generalized logistic regression

We should use mixed effects to evaluate longitudinal data and account for within
subject correlation

Should we be using (Days | patient)? No, outcome doesn't change
could consider (microbe | patient)?

# ```{r}
# capture.models <- function(outcome, lfun, modin.dat){
#   mods.list <- lapply(microbes, function(x) 
#   try({
#     glmer(as.formula(paste0(outcome, " ~ `", x, "`* Days + (1 | `Patient Id`)")), 
#           family = lfun, 
#           data = modin.dat) %>%
#       broom.mixed::tidy()  
#     })
#   )
#   mods.list.clean <- list.clean(mods.list, function(x) is.null(x))
#   mods.df <- bind_rows(mods.list.clean)
#   return(mods.df)
# }
# ```

a variety of errors ...
Error : pwrssUpdate did not converge in (maxit) iterations
fixed-effect model matrix is rank deficient so dropping 1 column / coefficient
Error : (maxstephalfit) PIRLS step-halvings failed to reduce deviance in pwrssUpdate
Error in pwrssUpdate(pp, resp, tol = tolPwrss, GQmat = GQmat, compDev = compDev,  : 
  Downdated VtV is not positive definite

can't bind rows due to NULL output 

New version of function that removes NULL values and reports taxa
```{r}
capture.models <- function(outcome, lfun, modin.dat, microbes) {
  mods.list <- list()
  for (x in microbes) {
    tryCatch({
      model_formula <- as.formula(paste0(outcome, " ~ ", x, "* Days + (1 | `Patient Id`)"))
      
      model <- glmer(model_formula, 
                     family = lfun, 
                     data = modin.dat)
      
      mods.list[[x]] <- broom.mixed::tidy(model)
      
    }, error = function(e) {
      message("Model failed for ", x, ": ", e$message)
      mods.list[[x]] <- NULL  # Assign NULL to failed models
    })
  }
  
  mods.list.clean <- mods.list[!sapply(mods.list, is.null)]
  
  if (length(mods.list.clean) > 0) {
    mods.df <- bind_rows(mods.list.clean)
  } else {
    mods.df <- data.frame() 
  }
  
  return(mods.df)
}
```


Microbe:days === interpret this one
```{r, warning=FALSE}
g.res <- capture.models("response.bin", "binomial", model.in, microbes)

g.res <- g.res %>%
  # filter to just microbe and the interaction term 
  filter(grepl("__", term),
         is.na(group)) %>%
  mutate(padj = p.adjust(p.value, method = "bonferroni"))


test <- g.res %>%
  filter(grepl(":", term))
# 1622/3228 made it through the modelling
# need to further investigate the model errors 

```

```{r}
write.csv(g.res, "../data/exploratory-longitudinal-model_output.csv")
```



Response ~ microbe*days.since.C1D1^2 + days.since.C1D1 + (1|ptID)
Response as 0 or 1, generalized logistic regression
```{r}
capture.models.sq <- function(outcome, lfun, modin.dat){
  mods.list <- lapply(microbes, function(x)
    try({glmer(as.formula(paste0(outcome, " ~ `", x, "` + Days^2 + Days + (1|`Patient Id`)")), 
             family = lfun,
             data = modin.dat) %>%
        tidy()})
  )
  
  mods.list.clean <- list.clean(mods.list, function(x) is.null(x))
  mods.df <- bind_rows(mods.list.clean)
  return(mods.df)
}
```

Terms:
Days
Microbe
Microbe:days^2 === interpret this one
```{r, warning=FALSE}
g.res <- capture.models("response.bin", "family = binomial(link = \"logit\")", modin.clin)

# g.res <- g.res %>%
#   filter(!str_detect(term, "(Intercept)|Days")) %>%
#   mutate(padj = p.adjust(p.value, method = "bonferroni"))
```

