---
title: "venn diagram"
author: "C. Dravillas"
date: "2025-05-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(VennDiagram)
library(ggVennDiagram)
library(patchwork)
```

Check overlap between the various model results 

4 groups total 
ancom-bc vs longitudinal
responders and non-responders 

read in necessary results 
```{r}
load("../manuscript/data/ANCOM-BC2_response_baseline.RData")
load("../manuscript/data/ZIGMM-model-output.rds")
load("../manuscript/data/ANCOM-BC2_irae_baseline.RData")
load("../manuscript/data/ZIGMM-model-output-irAE.rds")
```

get microbe sets 
- should also check/include structural zeroes 
- what term do we want to look at ?
  - if interested in how the two models perform, we could just look at response
  - this term is dif between groups (at baseline)
  - but first we will look at the interaction term since that's what i have saved
  - baseline, linear, quadratic terms 

```{r}
ancom.r <- (res.recist %>%
  filter(q_ResponseResponder < 0.05,
         lfc_ResponseResponder > 0))$taxon
ancom.nr <- (res.recist %>%
  filter(q_ResponseResponder < 0.05,
         lfc_ResponseResponder < 0))$taxon

gmm.out <- gmm.out %>%
  mutate(taxon = substr(microbe, 4, nchar(microbe)))

# almost no point in this, only one value...this is telling 
base.r <- (gmm.out %>%
  filter(term == "ResponseResponder",
         padj < 0.05,
         estimate > 0))$taxon
base.nr <- (gmm.out %>%
  filter(term == "ResponseResponder",
         padj < 0.05,
         estimate < 0))$taxon

lin.r <- (gmm.out %>%
  filter(term == "ResponseResponder:Days",
         padj < 0.05,
         estimate > 0))$taxon 
lin.nr <- (gmm.out %>%
  filter(term == "ResponseResponder:Days",
         padj < 0.05,
         estimate < 0))$taxon 

quad.r <- (gmm.out %>%
  filter(term == "ResponseResponder:I(Days^2)",
         padj < 0.05,
         estimate > 0))$taxon 
quad.nr <- (gmm.out %>%
  filter(term == "ResponseResponder:I(Days^2)",
         padj < 0.05,
         estimate < 0))$taxon 
```


Examine overlap across groups with padj, notably sparse
```{r}
venn1 <- list(
  ANCOM_R = ancom.r,
  ANCOM_NR = ancom.nr,
  Base_R = base.r,
  Base_NR = base.nr
)

venn2 <- list(
  ANCOM_R = ancom.r,
  ANCOM_NR = ancom.nr,
  LIN_R = lin.r,
  LIN_NR = lin.nr
)

venn3 <- list(
  ANCOM_R = ancom.r,
  ANCOM_NR = ancom.nr,
  QUAD_R = quad.r,
  QUAD_NR = quad.nr
)

# Create plots
p1 <- ggVennDiagram(venn1, label = "count", label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "cornflowerblue") +
  theme(legend.position = "none")

p2 <- ggVennDiagram(venn2, label = "count", label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "darkseagreen2") +
  theme(legend.position = "none")

p3 <- ggVennDiagram(venn3, label = "count", label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "salmon2") +
  theme(legend.position = "none")

p1 + p2 + p3
```


```{r}
venn1 <- list(
  ANCOM_R = ancom.r,
  ANCOM_NR = ancom.nr,
  Base_R = base.r,
  Base_NR = base.nr
)

venn2 <- list(
  ANCOM_R = ancom.r,
  ANCOM_NR = ancom.nr,
  LIN_R = lin.r,
  LIN_NR = lin.nr
)

venn3 <- list(
  ANCOM_R = ancom.r,
  ANCOM_NR = ancom.nr,
  QUAD_R = quad.r,
  QUAD_NR = quad.nr
)

# Create plots
p1 <- ggVennDiagram(venn1, label = "count", label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "cornflowerblue") +
  theme(legend.position = "none")

p2 <- ggVennDiagram(venn2, label = "count", label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "darkseagreen2") +
  theme(legend.position = "none")

p3 <- ggVennDiagram(venn3, label = "count", label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "salmon2") +
  theme(legend.position = "none")

p1 + p2 + p3
```


Examine overlap at baseline across different p-val cutoffs
```{r}
# what if we look at unadjusted p-value instead
base.r.05 <- (gmm.out %>%
  filter(term == "ResponseResponder",
         p.value < 0.05,
         estimate > 0))$taxon
base.nr.05 <- (gmm.out %>%
  filter(term == "ResponseResponder",
         p.value < 0.05,
         estimate < 0))$taxon

base.r.1 <- (gmm.out %>%
  filter(term == "ResponseResponder",
         p.value < 0.1,
         estimate > 0))$taxon
base.nr.1 <- (gmm.out %>%
  filter(term == "ResponseResponder",
         p.value < 0.1,
         estimate < 0))$taxon

# there is still very little overlap
# what if we increase p-value cut off
base.r.2 <- (gmm.out %>%
  filter(term == "ResponseResponder",
         p.value < 0.2,
         estimate > 0))$taxon 
base.nr.2 <- (gmm.out %>%
  filter(term == "ResponseResponder",
         p.value < 0.2,
         estimate < 0))$taxon 
```

```{r}
venn4 <- list(
  ANCOM_R = ancom.r,
  ANCOM_NR = ancom.nr,
  Base_R.05 = base.r.05,
  Base_NR.05 = base.nr.05
)

venn5 <- list(
  ANCOM_R = ancom.r,
  ANCOM_NR = ancom.nr,
  Base_R.1 = base.r.1,
  Base_NR.1 = base.nr.1
)

venn6 <- list(
  ANCOM_R = ancom.r,
  ANCOM_NR = ancom.nr,
  Base_R.2 = base.r.2,
  Base_NR.2 = base.nr.2
)

# Create plots
p4 <- ggVennDiagram(venn4, label = "count", label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "cornflowerblue") +
  theme(legend.position = "none")

p5 <- ggVennDiagram(venn5, label = "count", label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "darkseagreen2") +
  theme(legend.position = "none")

p6 <- ggVennDiagram(venn6, label = "count", label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "salmon2") +
  theme(legend.position = "none")

p4 + p5 + p6

ggsave("../figures/exploratory-venn_base.png",  width = 12, height = 5, dpi = 300)
```
```{r}
venn.05_r <- list(
  ANCOM_R = ancom.r,
  Base_R.05 = base.r.05
)
  
venn.05_nr <- list(
  ANCOM_NR = ancom.nr,
  Base_NR.05 = base.nr.05
)

venn.1_r <- list(
  ANCOM_R = ancom.r,
  Base_R.1 = base.r.1
)
  
venn.1_nr <- list(
  ANCOM_NR = ancom.nr,
  Base_NR.1 = base.nr.1
)


# Venn diagrams
r.05 <- ggVennDiagram(venn.05_r, label = "count", label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "cornflowerblue") +
  theme(legend.position = "none",
        plot.margin = margin(5, 10, 5, 10))

nr.05 <- ggVennDiagram(venn.05_nr, label = "count", label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "coral") +
  theme(legend.position = "none",
        plot.margin = margin(5, 10, 5, 10))

r.1 <- ggVennDiagram(venn.1_r, label = "count", label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "seagreen") +
  theme(legend.position = "none",
        plot.margin = margin(5, 10, 5, 10))

nr.1 <- ggVennDiagram(venn.1_nr, label = "count", label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "orchid") +
  theme(legend.position = "none",
        plot.margin = margin(5, 10, 5, 10))

(r.05 | nr.05 | r.1 | nr.1) + plot_layout(guides = "collect")

```

```{r}
gmm.out <- gmm.out %>%
  mutate(taxon = substr(microbe, 4, nchar(microbe)))

ancom.r <- (res.recist %>%
  filter(q_ResponseResponder < 0.05,
         lfc_ResponseResponder > 0))$taxon
ancom.nr <- (res.recist %>%
  filter(q_ResponseResponder < 0.05,
         lfc_ResponseResponder < 0))$taxon
base.r <- (gmm.out %>%
  filter(term == "ResponseResponder",
         p.value < 0.05,
         estimate > 0))$taxon
base.nr <- (gmm.out %>%
  filter(term == "ResponseResponder",
         p.value < 0.05,
         estimate < 0))$taxon

lin.r <- (gmm.out %>%
  filter(term == "ResponseResponder:Days",
         p.value < 0.05,
         estimate > 0))$taxon 
lin.nr <- (gmm.out %>%
  filter(term == "ResponseResponder:Days",
         p.value < 0.05,
         estimate < 0))$taxon 
quad.r <- (gmm.out %>%
  filter(term == "ResponseResponder:I(Days^2)",
         p.value < 0.05,
         estimate > 0))$taxon 
quad.nr <- (gmm.out %>%
  filter(term == "ResponseResponder:I(Days^2)",
         p.value < 0.05,
         estimate < 0))$taxon 
```

```{r}
p1 <- ggVennDiagram(list(ANCOM = ancom.nr, GMM.base = base.nr), label = "count", label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "deeppink4")  +
  theme(legend.position = "none") +
  xlim(-10, 10) 
p2 <- ggVennDiagram(list(ANCOM = ancom.r, GMM.base = base.r), label = "count", label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "cornflowerblue") +
  theme(legend.position = "none") +
  xlim(-10, 10) 

p3 <- ggVennDiagram(list(ANCOM = ancom.nr, GMM.lin = lin.nr), label = "count", label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "deeppink4")  +
  theme(legend.position = "none") +
  xlim(-10, 10) 
p4 <- ggVennDiagram(list(ANCOM = ancom.r, GMM.lin = lin.r), label = "count", label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "cornflowerblue") +
  theme(legend.position = "none") +
  xlim(-10, 10) 

p5 <- ggVennDiagram(list(ANCOM = ancom.nr, GMM.quad = quad.nr), label = "count", label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "deeppink4")  +
  theme(legend.position = "none") +
  xlim(-10, 10) 
p6 <- ggVennDiagram(list(ANCOM = ancom.r, GMM.quad = quad.r), label = "count", label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "cornflowerblue") +
  theme(legend.position = "none") +
  xlim(-10, 10) 
```

```{r}
(p1 | p2 ) /
(p3 | p4 ) /
(p5 | p6)
```
