---
title: "irAE longitudinal models"
author: "C. Dravillas"
date: "2025-04-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(broom.mixed)
library(rlist)
library(NBZIMM)
```

```{r}
load("../manuscript/data/clinical.rds")
load("../manuscript/data/microbiome_metadata.rds")
load("../manuscript/data/counts-long.rds")
load("../manuscript/data/ra-long.rds")
load("../manuscript/data/events.rds")
```

From last meeting 
Irae model
- NO time
- Use the same gaussian model 
- Also try irae ~ microbe + (1|ptid) 


some other potential ideas

irAE as response var, with irAE changing over time, so pt becomes irAE yes after 
having an irAE? Forever or for some amount of time?

Could also look for known signatures? 



# Prepare data 
- lets say irAE yes if within 10 days so same as clustering def

```{r}
# define time points 
irAE.time <- events %>%
  filter(Event == "irAE") %>%
  mutate(irAE.day = Days) %>%
  dplyr::select(`Patient Id`, irAE.day) 

meta.irae <- mic.meta %>%
  left_join(demographics) %>%
  left_join(irAE.time) %>%
  mutate(irAE.day = ifelse(irAE.day < 0 | irAE.day > 100, NA, irAE.day),
         irAE.sample = ifelse(is.na(irAE.day), 0,
                       ifelse(abs(Days - irAE.day) <= 10, 1, 0))) %>%
  group_by(sample) %>%
  slice_max(irAE.sample, n = 1, with_ties = FALSE) %>%
  ungroup() 

pres <- counts.l %>%
  filter(sample %in% mic.meta$sample) %>%
  group_by(Taxonomy) %>%
  summarize(pres = sum(est.reads > 0)) %>%
  mutate(microbe = str_extract(Taxonomy, "[^|]+$")) %>%
  dplyr::select(-Taxonomy)

model.in <- counts.l %>%
  mutate(microbe = str_extract(Taxonomy, "[^|]+$")) %>%
  dplyr::select(-Taxonomy) %>%
  full_join(meta.irae) %>%
  # create binary outcome var
  mutate(Days = as.integer(Days)) %>%
  rename(Patient_Id = `Patient Id`) %>%
  filter(!grepl("t__", microbe)) %>%
  pivot_wider(names_from = microbe, values_from = est.reads, values_fill = 0) %>%
  filter(sample %in% mic.meta$sample)

lib.size <- counts.l %>%
  filter(Taxonomy %in% c("k__Bacteria","k__Archaea")) %>%
  group_by(sample) %>%
  summarize(library.size = sum(est.reads))

model.in <- model.in %>%
  left_join(lib.size)

model.in.ra <- ra.l %>%
  full_join(meta.irae) %>% 
  # create binary outcome var
  mutate(Days = as.integer(Days)) %>%
  rename(Patient_Id = `Patient Id`) %>%
  filter(!grepl("t__", microbe)) %>%
  pivot_wider(names_from = microbe, values_from = RelAbun, values_fill = 0) %>%
  filter(sample %in% mic.meta$sample)

microbes <- colnames(model.in)
microbes <- microbes[grepl("__", microbes)]
microbes.ra <- colnames(model.in.ra)
microbes.ra <- microbes.ra[grepl("__", microbes.ra)]
```

# Select dif test microbes for irAE modeling 

```{r}
test.mics <- c("f__Muribaculaceae",  "s__Bacteroides_pectinophilus", "s__Enterobacter_roggenkampii", "s__Pseudoflavonifractor_gallinarum", "g__Fournierella", "s__Blautia_sp_MCC283")
```

```{r}
# Function to fit models
fit_models <- function(x, model.in, model.in.ra) {
  model_formula_off <- as.formula(paste0("irAE.sample ~ ", x, " + offset(log(library.size))"))
  model_formula <- as.formula(paste0("irAE.sample ~ ", x))

  models <- list(
    # FZINBMM (Counts)
    "ZINBMM_Offset" = glmm.zinb(model_formula_off, data = model.in, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL),
    # FZINBMM w/ AR1
    "ZINBMM_AR1" = glmm.zinb(model_formula_off, data = model.in, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL, correlation = corAR1()),
    # FZINBMM w/ no offset
    "ZINBMM" = glmm.zinb(model_formula, data = model.in, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL),
    # ZIGMM (RA)
    "ZIGMM" = lme.zig(model_formula, data = model.in.ra, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL),
    # ZIGMM w/ AR1
    "ZIGMM_AR1" = lme.zig(model_formula, data = model.in.ra, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL, correlation = corAR1()),
    # ZIGMM w/ CAR1
    "ZIGMM_CAR1" = lme.zig(model_formula, data = model.in.ra, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL, correlation = corCAR1()),
    # ZIGMM w/ ARMA
    "ZIGMM_ARMA" = lme.zig(model_formula, data = model.in.ra, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL, correlation = corARMA(p = 1, q = 1))
  )

  return(models)
}

# Compute model metrics
compute_metrics <- function(model, observed) {
  # mse = mean of (error)^2
  predictions <- predict(model, type = "response")
  residuals_sq <- (predictions - observed)^2
  mse <- mean(residuals_sq, na.rm = TRUE)
  rmse <- sqrt(mse)

  # define n and p 
  n <- length(observed)
  p <- length(coef(model))
  
  # variance of residuals
  sigma2 <- sum(residuals(model)^2) / (n - p)
  
  # Loglik 
  logLik <- -(n/2) * (log(2*pi*sigma2) + 1) - sum(residuals(model)^2) / (2 * sigma2)

  # number of estimate params..i think k should be 6 ?
  # + random intercept and zi intercept 
  k <- length(coef(model)) + 2
  aic <- (2 * k) - (2 * logLik)
  bic <- (log(n) * k) - (2 * logLik)

  return(c(MSE = mse, RMSE = rmse, AIC = aic, BIC = bic, LogLik = logLik))
}

# Loop over microbes
results <- list()
for (x in test.mics) {
  models <- fit_models(x, model.in, model.in.ra)

  model_metrics <- lapply(names(models), function(model_name) {
    # Use the correct observed dataset
    observed <- if (grepl("GMM", model_name)) model.in.ra[[x]] else model.in[[x]]
    compute_metrics(models[[model_name]], observed)
  })

  model_metrics <- do.call(rbind, model_metrics)
  rownames(model_metrics) <- names(models)

  results[[x]] <- model_metrics
  print(paste("Microbe:", x))
  print(model_metrics)
}

# Combine all results into a single data frame
final_results <- do.call(rbind, lapply(names(results), function(m) {
  cbind(Microbe = m, results[[m]])
}))

final_irAE <- as.data.frame(final_results)
final_irAE[, -1] <- lapply(final_irAE[, -1], as.double)
final_irAE <- final_irAE %>% rownames_to_column("Model") %>%
  mutate(Model = gsub("\\.\\d+$", "", Model))

```
Confirm that ZIGMM fits data best with irAE ~ microbe

Now try microbe ~ irAE
Need to add try catch for model convergence 

```{r}
# Fit a list of models for each microbe
fit_models <- function(x, model.in, model.in.ra) {
  # Make sure x is a column in the datasets
  model.in[[x]] <- as.numeric(model.in[[x]])
  model.in.ra[[x]] <- as.numeric(model.in.ra[[x]])

  # Formulas where microbe counts are outcome
  model_formula_off <- as.formula(paste0(x, " ~ irAE.sample + offset(log(library.size))"))
  model_formula <- as.formula(paste0(x, " ~ irAE.sample"))

  models <- list(
    "ZINBMM_Offset" = tryCatch(
      glmm.zinb(model_formula_off, data = model.in, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL),
      error = function(e) NULL
    ),
    "ZINBMM_AR1" = tryCatch(
      glmm.zinb(model_formula_off, data = model.in, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL, correlation = corAR1()),
      error = function(e) NULL
    ),
    "ZINBMM" = tryCatch(
      glmm.zinb(model_formula, data = model.in, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL),
      error = function(e) NULL
    ),
    "ZIGMM" = tryCatch(
      lme.zig(model_formula, data = model.in.ra, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL),
      error = function(e) NULL
    ),
    "ZIGMM_AR1" = tryCatch(
      lme.zig(model_formula, data = model.in.ra, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL, correlation = corAR1()),
      error = function(e) NULL
    ),
    "ZIGMM_CAR1" = tryCatch(
      lme.zig(model_formula, data = model.in.ra, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL, correlation = corCAR1()),
      error = function(e) NULL
    ),
    "ZIGMM_ARMA" = tryCatch(
      lme.zig(model_formula, data = model.in.ra, random = ~ 1 | Patient_Id, zi_fixed = ~1, zi_random = NULL, correlation = corARMA(p = 1, q = 1)),
      error = function(e) NULL
    )
  )

  return(models)
}

# Compute metrics (excluding estimate)
compute_metrics <- function(model, observed) {
  if (is.null(model)) {
    return(c(MSE = NA, RMSE = NA, AIC = NA, BIC = NA, LogLik = NA))
  }

  predictions <- tryCatch(predict(model, type = "response"), error = function(e) NA)
  residuals_sq <- (predictions - observed)^2
  mse <- mean(residuals_sq, na.rm = TRUE)
  rmse <- sqrt(mse)

  n <- length(observed)
  p <- length(coef(model))
  sigma2 <- sum(residuals(model)^2, na.rm = TRUE) / (n - p)
  logLik <- -(n/2) * (log(2 * pi * sigma2) + 1) - sum(residuals(model)^2, na.rm = TRUE) / (2 * sigma2)
  k <- length(coef(model)) + 2

  aic <- (2 * k) - (2 * logLik)
  bic <- (log(n) * k) - (2 * logLik)

  return(c(MSE = mse, RMSE = rmse, AIC = aic, BIC = bic, LogLik = logLik))
}

# Loop over microbes
results <- list()
for (x in test.mics) {
  models <- fit_models(x, model.in, model.in.ra)

  model_metrics <- lapply(names(models), function(model_name) {
    observed <- if (grepl("GMM", model_name)) model.in.ra[[x]] else model.in[[x]]
    compute_metrics(models[[model_name]], observed)
  })

  model_metrics <- do.call(rbind, model_metrics)
  rownames(model_metrics) <- names(models)

  # Add microbe name
  model_metrics <- cbind(Microbe = x, Model = rownames(model_metrics), model_metrics)
  results[[x]] <- model_metrics

  print(paste("Microbe:", x))
  print(model_metrics)
}

# Combine all results into a single data frame
final_mic <- do.call(rbind, results)
final_mic <- as.data.frame(final_mic)

# Ensure numeric columns
num_cols <- c("MSE", "RMSE", "AIC", "BIC", "LogLik")
final_mic[num_cols] <- lapply(final_mic[num_cols], as.numeric)
```

And ZIGMM with no specified correlation structure is still clear winner 
