---
title: 'Exploratory: machine learning markers'
author: "Rebecca Hoyd"
date: "8/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readxl)
library(vegan)
library(randomForest)
library(ROCR)
library(dplyr)
library(tidyr)
library(ggplot2)
```


# Functions

```{r}
inputvar_config_test <- function(trainmat, testmat, trainclin, testclin, topredict){
  model.training <- randomForest(x = trainmat, y = as.factor(trainclin[[topredict]]))
  test.preds <- predict(model.training, testmat)
  testpred.mat <- table(observed = testclin[[topredict]], predicted = test.preds)
  
  prediction_for_roc_curve <- predict(model.training,testmat,type="prob")
  
  # Specify the different classes 
  classes <- levels(as.factor(testclin[[topredict]]))
  
  perf.list <- list()
  # For each class
  for (i in 1:length(classes)){
    # Define which observations belong to class[i]
    true_values <- ifelse(testclin[[topredict]]==classes[i],1,0)
    # Assess the performance of classifier for class[i]
    pred <- prediction(prediction_for_roc_curve[,i],true_values)
    perf <- performance(pred, "tpr", "fpr")
    perf.list[[i]] <- as.data.frame(cbind(x = perf@x.values[[1]], y = perf@y.values[[1]],
                                          predicted.var = classes[i]))
    
    # auc.perf <- performance(pred, measure = "auc")
    # auclist[[i]] <- auc.perf@y.values
  }
  
  out.dat <- list(model.training, testpred.mat, bind_rows(perf.list))
  names(out.dat) <- c("rForest.mod", "tested.predictions", "model.performance")
  return(out.dat)
}
```

# Load data

```{r}
raw.combined.obj <- read_excel("T:/Labs/Spakowicz/mitox/data/clinical/raw/Adjuvant Endpoint data for Lianbo.xlsx",
                               sheet = "Treatment duration")
```

# Format

## Make all columns

I'm looking for all the columns we need for the predictor variables, then we'll format the predicted variable. We want the marker columns and all the reasonable clinical variables too.
```{r}
test <- colnames(raw.combined.obj)
test

# Markers start at column 26, we want to make sure we keep all those

marker.cols <- test[26:104]

check <- raw.combined.obj[1:25]

# I've selected these based on completeness and just glancing to see if there's a spread in the variables. I could add all the 
# lymphocyte information too.
clin.cols <- c("Sex", "Age", "Stage (7th edition prior to starting therapy)", "BMI", "Smoking Status")
```

```{r}
limited.obj <- raw.combined.obj %>%
  select(`Sample  (n=58)`, clin.cols, marker.cols) %>%
  mutate(`Smoking Status` = ifelse(is.na(`Smoking Status`), "Unknown", `Smoking Status`),
         YWHAZ = ifelse(is.na(YWHAZ), median(YWHAZ, na.rm = T), YWHAZ)) %>%
  mutate_at(.vars = clin.cols, ~ as.numeric(as.factor(.)))

```

Now let's make the predicted variable. We were given these tests:

1)      Completed vs. Discontinuation for any other reason

2)      Completed vs Recurrence

3)      Completed vs Toxicity

But it looks like the later two do not have a good spread between the categories.
```{r}
table(raw.combined.obj$`Reason for discontinuation`)

therapy.continuance <- raw.combined.obj %>%
  select(`Sample  (n=58)`, `Reason for discontinuation`) %>%
  mutate(any.discontinuation = ifelse(`Reason for discontinuation` == "Completed therapy", 0, 1),
         recurrences = ifelse(`Reason for discontinuation` == "Completed therapy", 0, 
                              ifelse(`Reason for discontinuation` == "Recurrence", 1, NA)),
         toxicity = ifelse(`Reason for discontinuation` == "Completed therapy", 0, 
                              ifelse(`Reason for discontinuation` == "Toxicity" | `Reason for discontinuation` == "Toxixity",
                                     1, NA)))
  
```

## Make model input dataframes

````{r}
# Any discontinuation
set.seed(112358)
disc.ntest <- round(nrow(therapy.continuance)*.25)
disc.tsamps <- sample(therapy.continuance$`Sample  (n=58)`, disc.ntest, replace = F)

disc.trainmat <- limited.obj %>%
  filter(!`Sample  (n=58)` %in% disc.tsamps) %>%
  select(-`Sample  (n=58)`)

disc.testmat <- limited.obj %>%
  filter(`Sample  (n=58)` %in% disc.tsamps) %>%
  select(-`Sample  (n=58)`)

disc.trainclin <- therapy.continuance %>%
  filter(!`Sample  (n=58)` %in% disc.tsamps) %>%
  select(-`Sample  (n=58)`)

disc.testclin <- therapy.continuance %>%
  filter(`Sample  (n=58)` %in% disc.tsamps) %>%
  select(-`Sample  (n=58)`)

# Recurrence

reccurence.clin <- therapy.continuance %>%
  drop_na(recurrences)

set.seed(112358)
rec.ntest <- round(nrow(reccurence.clin)*.25)
rec.tsamps <- sample(reccurence.clin$`Sample  (n=58)`, rec.ntest, replace = F)

rec.trainclin <- reccurence.clin %>%
  filter(!`Sample  (n=58)` %in% rec.tsamps) 

rec.testclin <- reccurence.clin %>%
  filter(`Sample  (n=58)` %in% rec.tsamps) 
rec.trainmat <- limited.obj %>%
  filter(`Sample  (n=58)` %in% rec.trainclin$`Sample  (n=58)`) %>%
  select(-`Sample  (n=58)`)

rec.testmat <- limited.obj %>%
  filter(`Sample  (n=58)` %in% rec.testclin$`Sample  (n=58)`) %>%
  select(-`Sample  (n=58)`)

# Toxicity

toxicity.clin <- therapy.continuance %>%
  drop_na(toxicity)

set.seed(112358)
tox.ntest <- round(nrow(toxicity.clin)*.25)
tox.tsamps <- sample(toxicity.clin$`Sample  (n=58)`, tox.ntest, replace = F)

tox.trainclin <- toxicity.clin %>%
  filter(!`Sample  (n=58)` %in% tox.tsamps)

tox.testclin <- toxicity.clin %>%
  filter(`Sample  (n=58)` %in% tox.tsamps) 

tox.trainmat <- limited.obj %>%
  filter(`Sample  (n=58)` %in% tox.trainclin$`Sample  (n=58)`) %>%
  select(-`Sample  (n=58)`)

tox.testmat <- limited.obj %>%
  filter(`Sample  (n=58)` %in% tox.testclin$`Sample  (n=58)`) %>%
  select(-`Sample  (n=58)`)
```

# Tests

## Any discontinuation
```{r}
disc.res <- inputvar_config_test(disc.trainmat, disc.testmat, disc.trainclin, disc.testclin, "any.discontinuation")

disc.res$rForest.mod$importance %>%
  as.data.frame() %>%
  arrange(desc(MeanDecreaseGini))%>%
  head()

disc.res$model.performance %>%
  mutate(x =as.numeric(x),
         y = as.numeric(y)) %>%
  filter(predicted.var == 0) %>%
  ggplot(aes(x = x, y = y)) +
  geom_abline(slope = 1, intercept = 0, color = "grey") +
  geom_path() +
  labs(x = "False Positive Rate", y = "True Positive Rate") +
  theme_bw() +
  theme(axis.text= element_blank(),
        axis.ticks = element_blank()) 
```

## Recurrences

```{r}
rec.res <- inputvar_config_test(rec.trainmat, rec.testmat, rec.trainclin, rec.testclin, "recurrences")

rec.res$rForest.mod$importance %>%
  as.data.frame() %>%
  arrange(desc(MeanDecreaseGini))%>%
  head()

rec.res$model.performance %>%
  mutate(x =as.numeric(x),
         y = as.numeric(y)) %>%
  filter(predicted.var == 1) %>%
  ggplot(aes(x = x, y = y)) +
  geom_abline(slope = 1, intercept = 0, color = "grey") +
  geom_path() +
  labs(x = "False Positive Rate", y = "True Positive Rate") +
  theme_bw() +
  theme(axis.text= element_blank(),
        axis.ticks = element_blank())
```

## Toxicity

```{r}
tox.res <- inputvar_config_test(tox.trainmat, tox.testmat, tox.trainclin, tox.testclin, "toxicity")

tox.res$rForest.mod$importance %>%
  as.data.frame() %>%
  arrange(desc(MeanDecreaseGini)) %>%
  head()

tox.res$model.performance %>%
  mutate(x =as.numeric(x),
         y = as.numeric(y)) %>%
  filter(predicted.var == 1) %>%
  ggplot(aes(x = x, y = y)) +
  geom_abline(slope = 1, intercept = 0, color = "grey") +
  geom_path() +
  labs(x = "False Positive Rate", y = "True Positive Rate") +
  theme_bw() +
  theme(axis.text= element_blank(),
        axis.ticks = element_blank())
```