---
title: "longitudinal models"
author: "C. Dravillas"
date: "2025-02-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lme4)
library(broom.mixed)
library(rlist)
```

# Load data
```{r}
load("../manuscript/data/clinical.rds")
load("../manuscript/data/microbiome_metadata.rds")
load("../manuscript/data/ra-long.rds")
load("../manuscript/data/counts-long.rds")
```

# Combine variables 
```{r}
model.in <- ra.l %>%
  right_join(mic.meta) %>%
  left_join(demographics) %>%
  # create binary outcome var
  mutate(response.bin = ifelse(Response == "Responder", 1, 0),
  # make days numeric
         Days = as.integer(Days),
  # create quadratic days 
         Days2 = Days^2) %>%
  rename(Patient_Id = `Patient Id`) %>%
  pivot_wider(names_from = microbe, values_from = RelAbun) %>%
  select(where(~ !all(. == 0)))

model.in.c <- counts.l %>%
  mutate(microbe = str_extract(Taxonomy, "[^|]+$")) %>%
  select(-Taxonomy) %>%
  right_join(mic.meta) %>%
  left_join(demographics) %>%
  # create binary outcome var
  mutate(response.bin = ifelse(Response == "Responder", 1, 0),
  # make days numeric
         Days = as.integer(Days),
  # create quadratic days 
         Days2 = Days^2) %>%
  rename(Patient_Id = `Patient Id`) %>%
  pivot_wider(names_from = microbe, values_from = est.reads, values_fill = 0) %>%
  filter(sample %in% mic.meta$sample)

microbes <- colnames(model.in)
microbes <- microbes[grepl("__", microbes)]
# used for testing models 
#microbes <- c("s__Blautia_luti", "g__Solobacterium")
microbes <- microbes[!grepl("t__", microbes)]
# microbes.c <-  unique(counts.l$Taxonomy)
# microbes.c <- microbes.c[!grepl("t__", microbes.c)]
```


## Modeling

Response ~ microbe*days.since.C1D1 + (1|ptID) 
Response as 0 or 1, generalized logistic regression

We should use mixed effects to evaluate longitudinal data and account for within
subject correlation. 

Should we be using (Days | patient)? No, outcome doesn't change.
Could consider (microbe | patient)

Previous version of function couldn't bind rows due to NULL output 
New version of function that removes NULL values and reports taxa
```{r}
capture.models <- function(outcome, lfun, modin.dat, microbes) {
  mods.list <- list()
  for (x in microbes) {
    tryCatch({
      model_formula <- as.formula(paste0(outcome, " ~ ", x, "* Days + (1 | Patient_Id)"))
      
      model <- glmer(model_formula, 
                     family = lfun, 
                     data = modin.dat)
      
      mods.list[[x]] <- broom.mixed::tidy(model)
      
    }, error = function(e) {
      message("Model failed for ", x, ": ", e$message)
      mods.list[[x]] <- NULL  # Assign NULL to failed models
    })
  }
  
  mods.list.clean <- mods.list[!sapply(mods.list, is.null)]
  
  if (length(mods.list.clean) > 0) {
    mods.df <- bind_rows(mods.list.clean)
  } else {
    mods.df <- data.frame() 
  }
  
  return(mods.df)
}
```

Microbe:days === interpret this one
```{r, warning=FALSE}
res <- capture.models("response.bin", "binomial", model.in, microbes)
res.c <- capture.models("response.bin", "binomial", model.in.c, microbes)

res <- res %>%
  # filter to just microbe and the interaction term 
  filter(grepl("__", term),
         is.na(group)) %>%
         #grepl(":", term)) %>%
  mutate(padj = p.adjust(p.value, method = "bonferroni"))

res.c <- res.c %>%
  # filter to just microbe and the interaction term 
  filter(grepl("__", term),
         is.na(group)) %>%
         #grepl(":", term)) %>%
  mutate(padj = p.adjust(p.value, method = "bonferroni"))

# 1622/2644 made it through the modelling
# 556/2644 interaction terms made it through for counts
# need to further investigate the model errors 
```

Save and read in model output 
```{r}
# write.csv(res, "../data/modeling_RA_random_response.csv", row.names = F)
# write.csv(res.c, "../data/modeling_counts_random_response.csv", row.names = F)
res <- read.csv("../data/modeling_long_random_response.csv")
res.c <- read.csv("../data/modeling_counts_random_response.csv")
```

A variety of errors:
Error : pwrssUpdate did not converge in (maxit) iterations 
Error : (maxstephalfit) PIRLS step-halvings failed to reduce deviance in pwrssUpdate
Error in pwrssUpdate(pp, resp, tol = tolPwrss, GQmat = GQmat, compDev = compDev,  : 
  Downdated VtV is not positive definite
Model failed for s__Allisonella_histaminiformans: Downdated VtV is not positive definite

Also note:  
fixed-effect model matrix is rank deficient so dropping 1 column / coefficient
fixed-effect model matrix is rank deficient so dropping 2 columns / coefficient

Things to consider:
- sparsity
- abundance being constant across timepoints
- collinearity between abundance and days
- complete separation (i.e. response is always the same for abundance level or timepoint)

Could try:
- centering and scaling 
- check for sparsity and drop interaction for sparse data or use presence/absence
- check for complete separation
- log transform abundance - couldn't handle 0 values 

Conclusion: Due to inconsistent timepoints and abundance sparsity I think we can
expect to see rank deficiency. 

Sparse microbe?    ---> Skip or fit simpler model?
Complete separation? ---> Skip?
Severe collinearity? ---> Center/scale predictors?
Extreme outliers?  ---> Log-transform abundance...issues
Complex random effect with sparse data? ---> Simplify model

BUT because the response is always the same for every patient I think the random 
effect may be redundant. There are between patient differences but not within. 
I think our model might be trying to anticipate within subject variation. Should
we instead use response as a predictor of microbial abundance?

Let's try microbe ~ response*Days w/ random effect, don't think this makes sense 
```{r}
capture.models.mic <- function(modin.dat, microbes) {
  mods.list <- list()

  for (x in microbes) {
    tryCatch({
      model_formula <- as.formula(paste0(x, " ~ Response * Days + (1 | Patient_Id)"))
      model <- lmerTest::lmer(model_formula, data = modin.dat)
      
      model_summary <- broom.mixed::tidy(model, effects = "fixed", conf.int = TRUE)

      model_summary$microbe <- x

      mods.list[[x]] <- model_summary
    }, error = function(e) {
      mods.list[[x]] <- NULL
    })
  }

  mods.list.clean <- mods.list[!sapply(mods.list, is.null)]

  if (length(mods.list.clean) > 0) {
    mods.df <- bind_rows(mods.list.clean)

    mods.df <- mods.df %>%
      mutate(padj = p.adjust(p.value, method = "BH"))

  } else {
    mods.df <- data.frame()
  }

  return(mods.df)
}
```

several errors 
Warning: Computed variance-covariance matrix problem: not a positive definite matrix (and positive semidefiniteness is not checked);
returning NA matrixModel failed for s__GGB3817_SGB5182: not a positive definite matrix (and positive semidefiniteness is not checked)
- Rescale variables?Warning: Model may not have converged with 1 eigenvalue close to zero: 2.0e-10Warning: Model failed to converge with max|grad| = 0.133644 (tol = 0.002, component 1)Warning: Model is nearly unidentifiable: very large eigenvalue
Model failed for s__Lactobacillus_delbrueckii: Downdated VtV is not positive definite
Warning: convergence code -4 from nloptwrap: NLOPT_ROUNDOFF_LIMITED: Roundoff errors led to a breakdown of the optimization algorithm. In this case, the returned minimum may still be useful. (e.g. this error occurs in NEWUOA if one tries to achieve a tolerance too close to machine precision.)Warning: unable to evaluate scaled gradientWarning
boundary (singular) fit: see help('isSingular')

```{r}
res.mic <- capture.models.mic(model.in, microbes)
res.mic.c <- capture.models.mic(model.in.c, microbes)

res.mic <- res.mic %>%
  filter(term == "ResponseResponder:Days")
res.mic.c <- res.mic.c %>%
  filter(term == "ResponseResponder:Days")

# 2577/3228 made it through
```

```{r}
# write.csv(res.mic, "../data/modeling_RA_random_microbe.csv", row.names = F)
# write.csv(res.mic.c, "../data/modeling_counts_random_microbe.csv", row.names = F)
res.mic <- read.csv("../data/modeling_RA_random_microbe.csv")
res.mic.c <- read.csv("../data/modeling_counts_random_microbe.csv")
```

## Check quadratic term

First with response as outcome 

Response ~ microbe*(days.since.C1D1^2 + days.since.C1D1) + (1|ptID)
```{r}
capture.models.sq <- function(outcome, lfun, modin.dat, microbes) {
  
  mods.list <- list()
  
  for (x in microbes) {
    tryCatch({
      model_formula <- as.formula(paste0(outcome, " ~ `", x, "`* (Days + Days2) + (1 | Patient_Id)"))
      
      model <- glmer(model_formula, 
                     family = lfun, 
                     data = modin.dat)
      
      mods.list[[x]] <- broom.mixed::tidy(model)
      
    }, error = function(e) {
      message("Model failed for ", x, ": ", e$message)
      mods.list[[x]] <- NULL
    })
  }
  
  mods.list.clean <- mods.list[!sapply(mods.list, is.null)]
  
  if (length(mods.list.clean) > 0) {
    mods.df <- bind_rows(mods.list.clean)
  } else {
    mods.df <- data.frame() 
  }
  
  return(mods.df)
}

```

Terms:
Days
Microbe
Microbe:days^2 === interpret this one
```{r, warning=FALSE}
res.sq <- capture.models.sq("response.bin", "binomial", model.in, microbes)
res.sq.c <- capture.models.sq("response.bin", "binomial", model.in.c, microbes)
```

```{r}
res.sq <- res.sq %>%
  # filter to just microbe and the interaction term 
  filter(grepl(":Days2",term)) %>%
  mutate(padj = p.adjust(p.value, method = "bonferroni"))

 <- res.sq.c

# 807/3228 made it through the quadratric modelling 
# need to further investigate the model errors 
```

```{r}
# write.csv(res.sq, "../data/modeling_long_quad_response.csv", row.names = F)
res.sq <- read.csv("../data/modeling_long_quad_response.csv")
```



Now with microbial abundance as the outcome variable 
```{r}
capture.models.mic.sq <- function(modin.dat, microbes) {
  mods.list <- list()
  
  for (x in microbes) {
    tryCatch({
      model_formula <- as.formula(paste0(x, " ~ Response * (Days + Days2) + (1 | Patient_Id)"))
      model <- lmer(model_formula, data = modin.dat)
      model_summary <- broom.mixed::tidy(model, conf.int = TRUE, p.value = TRUE)
      model_summary$microbe <- x
      mods.list[[x]] <- model_summary
    }, error = function(e) {
      #message("Model failed for ", x, ": ", e$message)
      mods.list[[x]] <- NULL
    })
  }
  
  mods.list.clean <- mods.list[!sapply(mods.list, is.null)]
  
  if (length(mods.list.clean) > 0) {
    mods.df <- bind_rows(mods.list.clean)
  } else {
    mods.df <- data.frame()
  }
  
  return(mods.df)
}

```

```{r}
res.mic.sq <- capture.models.mic.sq(model.in, microbes)

res.mic.sq <- res.mic.sq %>%
  filter(term == "ResponseResponder:Days2")

# 2573/3228 made it through
```

```{r}
# write.csv(res.mic.sq, "../data/modeling_long_quad_microbe.csv", row.names = F)
res.mic.sq <- read.csv("../data/modeling_long_quad_microbe.csv")
```

