---
title: "aucpr_and_cfm_template"
author: "Noah Lepola"
date: "2025-02-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggplot2)
library(randomForest)
library(ROCR)
library(caTools)
library(openxlsx)
library(respondersig) 


```

Loading file
```{r}
source("../scripts/mitox_functions.R")
adjing.data <- openxlsx::loadWorkbook("T:/Labs/Spakowicz/mitox/Clinical info/Patient Data 7-1-22.xlsm")
irAE_aucprs <- readRDS("../data/rf_mitox_irAE_aucpr.RData")
Response_aucprs <- readRDS("../data/rf_mitox_Response_aucpr.RData")
load("../manuscript/data/ra-long.rds")
load("../manuscript/data/clinical.rds")
load("../manuscript/data/blood.rds")
load("../manuscript/data/microbiome_metadata.rds")
load("../manuscript/data/phylum-labels.rds")
load("../manuscript/data/struct0_response.rds")
load("../manuscript/data/struct0_irae.rds")
load("../manuscript/data/struct0_response.rds")
load("../manuscript/data/struct0_irae.rds")
load("../manuscript/data/ANCOM-BC2_response_baseline.Rdata")
load("../manuscript/data/ANCOM-BC2_irae_baseline.RData")


```






```{r}
seed_list <- c(12345:12369)
```




Making objects for structural zero and q value models
```{r}

#Object to limit microbiome sample to one ones taken within 10 days of treatment
baseline <- mic.meta %>%
  group_by(`Patient Id`) %>%
  arrange(Days) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  filter(abs(Days) <= 10) %>%
  subset(`Patient Id` %in% blood$`Patient Id`)

aucprdf <- data.frame(x = 1:25)

phyl_labels <- phyl_labels %>%
  mutate(taxon = substr(microbe, 4, nchar(microbe)))

#creating object for use in biome model
base.micro <- ra.l %>%
  pivot_wider(names_from = microbe, values_from = RelAbun) %>%
  subset(sample %in% baseline$sample)

#note it is dropping bacteria and archaea becaise it isn't in phyl_labels
#making object for irAE where q < 0.05
irAE_q <- res.irae %>%
  left_join(phyl_labels) %>%
  select(microbe, q_irAEYes)  %>%
  filter(q_irAEYes < 0.01) %>%
  replace_na(list( microbe = "k__Archaea"))



#structural 0 for irAE
struct0.irae.ft <- struct0.irae %>%
  filter(struct0.irae$`structural_zero (irAE = No)` == "FALSE" &
           struct0.irae$`structural_zero (irAE = Yes)` == "TRUE" |
           struct0.irae$`structural_zero (irAE = No)` == "TRUE" &
           struct0.irae$`structural_zero (irAE = Yes)` == "FALSE")  %>%
    left_join(phyl_labels) %>%
  select(microbe, 
         `structural_zero (irAE = No)`, 
         `structural_zero (irAE = Yes)`)

struct0.irae.list <- struct0.irae.ft$microbe %>%
  append("Patient Id", after = 0)


#limiting q < 0.05 to struct 0
irAE_q_zero <- irAE_q %>%
  filter(irAE_q$microbe %in% struct0.irae.ft$microbe)



#making object for response where q < 0.05
response_q <- res.recist %>%
  left_join(phyl_labels) %>%
  select(microbe, q_ResponseResponder)  %>%
  filter(q_ResponseResponder < 0.01) %>%
  replace_na(list( microbe = "k__Archaea"))

#structural 0 for response
struct0.recist.ft <- struct0.recist %>%
  filter(struct0.recist$`structural_zero (Response = Non-Responder)` == "FALSE" &
           struct0.recist$`structural_zero (Response = Responder)` == "TRUE" |
           struct0.recist$`structural_zero (Response = Non-Responder)` == "TRUE" &
           struct0.recist$`structural_zero (Response = Responder)` == "FALSE") %>%
    left_join(phyl_labels) %>%
  select(microbe, 
         `structural_zero (Response = Non-Responder)`, 
         `structural_zero (Response = Responder)`)

struct0.recist.list <- struct0.recist.ft$microbe %>%
  append("Patient Id", after = 0)

#limiting q < 0.05 to struct 0
response_q_zero <- response_q %>%
  filter(response_q$microbe %in% struct0.recist.ft$microbe)



response_testlist <- res.recist %>%
  filter(res.recist$taxon %in% struct0.recist$taxon)
irAE_testlist <- res.irae %>%
  filter(res.irae$taxon %in% struct0.irae$taxon)

irAE_list <- irAE_q$microbe %>%
  append("Patient Id", after = 0)

response_list <- response_q$microbe %>%
  append("Patient Id", after = 0) 

irAEcombinedlist <- c(irAE_list, struct0.irae.list)

```


Making responsesig objects for response
```{r}
#try splitting the dataset
utils::data("met4ra", package = "respondersig")
utils::data("metadata", package = "respondersig")

met4ra$clade_name <- gsub(".*\\|", "", met4ra$clade_name)

metadata <- metadata %>%
  filter(timepoint == "pre") %>%
  subset(metadata$sample %in% met4ra$sample) 

 metadata$response <- gsub("_", "-", metadata$response)


met4ra <- met4ra %>%
  subset(met4ra$sample %in% metadata$sample) %>%
  select(clade_name, relative_abundance, sample) %>%
  pivot_wider(names_from = clade_name, values_from = relative_abundance) %>% 
  replace(is.na(.), 0)

set.seed(42301)

#splitting data into train and test
split <- sample.split(metadata$response, SplitRatio = 0.7)

train  <- subset(metadata, split == TRUE) %>%
  subset(sample %in% met4ra$sample) %>%
  arrange(desc(sample))

test   <- subset(metadata, split == FALSE) %>%
  subset(sample %in% met4ra$sample) %>%
  arrange(desc(sample))

train.outcomes <- demographics %>%
  select(`Patient Id`, Response) %>%
  subset(`Patient Id` %in% baseline$`Patient Id`) %>%
  arrange(desc(`Patient Id`))

names(train.outcomes)[names(train.outcomes) == 'Response'] <- 'response'
names(train.outcomes)[names(train.outcomes) == 'Patient Id'] <- 'sample'


struct0.recist.rotate  <- struct0.recist.ft %>% 
  pivot_wider(names_from = microbe, values_from = `structural_zero (Response = Non-Responder)`) 
colnames(struct0.recist.rotate)[1] <- 'sample'

#combining structural zero and response, 
responsecombinedlist <- c(response_list, names(struct0.recist.rotate))


test.outcomes <- metadata %>%
  # filter(dataset == "Frankel_2017") %>%
  # filter(dataset == "Gopalakrishnan_2019") %>%
  # filter(dataset == "Matson_2019") %>%
  # filter(dataset == "McCulloch_2022") %>%
  filter(dataset == "Peters_2019") %>%
  filter(timepoint == "pre") %>%
  select(sample, response) %>%
  subset(sample %in% met4ra$sample) %>%
  arrange(desc(sample))

response_list[[1]] <- "sample"
struct0.recist.list[[1]] <- "sample"
responsecombinedlist[[1]] <- "sample"


  
#removing columns not present in respondersig or in structural zero
strct0_cols <- intersect(names(met4ra), names(struct0.recist.rotate))
q_cols <- intersect(names(met4ra), response_list)


combined_cols <- intersect(names(met4ra), responsecombinedlist)


# coi <- q_cols
# coi <- strct0_cols
coi <- combined_cols

met4ralim <- met4ra %>%
  select((coi))


microlim <- baseline %>%
  select(sample, `Patient Id`) %>%
  full_join(base.micro, baseline, by = "sample") %>%
  select(-sample)

names(microlim)[names(microlim) == 'Patient Id'] <- 'sample'

microlim <- microlim %>%
  select((coi))



train.seq <- filter(microlim,
                    (sample %in% train.outcomes$sample)) 


train.seq <- arrange(train.seq, desc(sample))



test.seq <- filter(met4ralim,
                    (sample %in% test.outcomes$sample))
test.seq <- arrange(test.seq, desc(sample))
```


Response model for resondersig
```{r}
#Response
rf_responce <- kTest(seed_list, "Non-Responder", "Responder", "response", 2000)

aucprs <- grabVals(aucprs, rf_responce, seed_list)

micro_only_respondersigPeters_2000_qzerofivestruct0 <- mean(aucprs$AUCPR)

micro_only_respondersigPeters_2000_qzerofivestruct0 <- as.data.frame(micro_only_respondersigPeters_2000_qzerofivestruct0) %>%
  rename(AUCPR = micro_only_respondersigPeters_2000_qzerofivestruct0) %>%
  add_column("name" = "micro_only_respondersigPeters_2000_qzerofivestruct0", .before = "AUCPR")

Response_aucprs <- Response_aucprs  %>%
  add_row(micro_only_respondersigPeters_2000_qzerofivestruct0)

aucprdf$micro_only_respondersigPeters_2000_qzerofivestruct0 <- aucprs$AUCPR

remove(micro_only_respondersigPeters_2000_qzerofivestruct0)


```

Response sig plot
```{r}

response_for_plot  <- Response_aucprs[c(28, 36, 47:56), ]

response_plot <-ggplot(data = response_for_plot,
                       mapping=aes(x = reorder(name, AUCPR),
                                   y = AUCPR, fill = AUCPR))+
    scale_y_continuous(limits = c(0, 1))+
    geom_bar(stat = "identity") +
    scale_fill_viridis_c(option = "magma", direction = -1) +
    coord_flip() + 
    xlab("Model")
   

# ggplot(df,aes(x,y,fill=y))+geom_bar(stat="identity")+scale_fill_gradient(low="blue",high="red")
print(response_plot)
# ggsave("../figures/responsemodelperformance.png", height = 4, width = 7)
```






Making responsesig objects for irAE
```{r}
#try splitting the dataset
utils::data("met4ra", package = "respondersig")
irAE_respondsig <- read.csv(file = "../data/cleaned-metadata-expanded.csv") %>%
  drop_na(irAE) %>%
  filter(timepoint == "pre") %>%
  select(sample, irAE)


met4ra$clade_name <- gsub(".*\\|", "", met4ra$clade_name)

irAE_respondsig <- irAE_respondsig %>%
  subset(irAE_respondsig$sample %in% met4ra$sample) 




met4ra <- met4ra %>%
  subset(met4ra$sample %in% irAE_respondsig$sample) %>%
  select(clade_name, relative_abundance, sample) %>%
  pivot_wider(names_from = clade_name, values_from = relative_abundance) %>% 
  replace(is.na(.), 0)

set.seed(42301)

#splitting data into train and test
split <- sample.split(irAE_respondsig$irAE, SplitRatio = 0.7)

train  <- subset(irAE_respondsig, split == TRUE) %>%
  subset(sample %in% met4ra$sample) %>%
  arrange(desc(sample))

test   <- subset(irAE_respondsig, split == FALSE) %>%
  subset(sample %in% met4ra$sample) %>%
  arrange(desc(sample))

train.outcomes <- demographics %>%
  select(`Patient Id`, irAE) %>%
  subset(`Patient Id` %in% baseline$`Patient Id`) %>%
  arrange(desc(`Patient Id`))

names(train.outcomes)[names(train.outcomes) == 'Patient Id'] <- 'sample'


struct0.irae.rotate  <- struct0.irae.ft %>% 
  pivot_wider(names_from = microbe, values_from = `structural_zero (irAE = No)`) 
colnames(struct0.irae.rotate)[1] <- 'sample'

#combining structural zero and irAE, 
irAEcombinedlist <- c(irAE_list, names(struct0.irae.rotate))


test.outcomes <- irAE_respondsig %>%
  select(sample, irAE) %>%
  subset(sample %in% met4ra$sample) %>%
  arrange(desc(sample))

test.outcomes$irAE[test.outcomes$irAE == 1] <- "Yes"
test.outcomes$irAE[test.outcomes$irAE == 0] <- "No"


irAE_list[[1]] <- "sample"
struct0.irae.list[[1]] <- "sample"
irAEcombinedlist[[1]] <- "sample"


  
#removing columns not present in respondersig or in structural zero
strct0_cols <- intersect(names(met4ra), names(struct0.irae.rotate))
q_cols <- intersect(names(met4ra), irAE_list)


combined_cols <- intersect(names(met4ra), irAEcombinedlist)


#making objects for q value, structural zero, or q value + structural zero
# coi <- q_cols
# coi <- strct0_cols
coi <- combined_cols

met4ralim <- met4ra %>%
  select((coi))


microlim <- baseline %>%
  select(sample, `Patient Id`) %>%
  full_join(base.micro, baseline, by = "sample") %>%
  select(-sample)

names(microlim)[names(microlim) == 'Patient Id'] <- 'sample'

microlim <- microlim %>%
  select((coi))



train.seq <- filter(microlim,
                    (sample %in% train.outcomes$sample)) 


train.seq <- arrange(train.seq, desc(sample))



test.seq <- filter(met4ralim,
                    (sample %in% test.outcomes$sample))
test.seq <- arrange(test.seq, desc(sample))


```


irAE model for resondersig
```{r}
rf_irAE <- kTest(seed_list, "No", "Yes", "irAE", 500)

aucprs <- grabVals(aucprs, rf_irAE, seed_list)

micro_only_irAEsig_500_qonestruct0 <- mean(aucprs$AUCPR)

micro_only_irAEsig_500_qonestruct0 <- as.data.frame(micro_only_irAEsig_500_qonestruct0) %>%
  rename(AUCPR = micro_only_irAEsig_500_qonestruct0) %>%
  add_column("name" = "micro_only_irAEsig_500_qonestruct0", .before = "AUCPR")

irAE_aucprs <- irAE_aucprs  %>%
  add_row(micro_only_irAEsig_500_qonestruct0)

aucprdf$micro_only_irAEsig_500_qonestruct0 <- aucprs$AUCPR

remove(micro_only_irAEsig_500_qonestruct0)


```

irAE plot
```{r}
irAE_for_plot  <- irAE_aucprs[c(5, 23, 31, 35, 36), ]

irAE_plot <-ggplot(data = irAE_for_plot,
                   mapping=aes(x = reorder(name, AUCPR),
                               y = AUCPR, fill = AUCPR))+
    scale_y_continuous(limits = c(0, 1))+
    geom_bar(stat = "identity") +
    scale_fill_viridis_c(option = "magma", direction = -1) +
    coord_flip() + 
    xlab("Model") 

# ggplot(df,aes(x,y,fill=y))+geom_bar(stat="identity")+scale_fill_gradient(low="blue",high="red")
print(irAE_plot)
```














Splitting the data into training and testing for mitox models-No respondersig 
```{r}

phyl_labels <- phyl_labels %>%
  mutate(taxon = substr(microbe, 4, nchar(microbe)))

res.irae <- res.irae %>%
  left_join(phyl_labels)

#Object to limit microbiome sample to one ones taken within 10 days of treatment
baseline <- mic.meta %>%
  group_by(`Patient Id`) %>%
  arrange(Days) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  filter(abs(Days) <= 10) %>%
  subset(`Patient Id` %in% blood$`Patient Id`)

#limiting blood samples to ones also with microbiome sample and creating object
base.gene <- blood %>%
  subset(`Patient Id` %in% baseline$`Patient Id`) %>%
  select(`Patient Id`:YWHAZ)

#creating object for use in biome model
base.micro <- ra.l %>%
  pivot_wider(names_from = microbe, values_from = RelAbun) %>%
  subset(sample %in% baseline$sample)



set.seed(42301)

#splitting data into train and test
split <- sample.split(demographics$Response, SplitRatio = 0.7)

train  <- subset(demographics, split == TRUE) %>%
  subset(`Patient Id` %in% blood$`Patient Id`) %>%
  subset(`Patient Id` %in% baseline$`Patient Id`)

test   <- subset(demographics, split == FALSE) %>%
  subset(`Patient Id` %in% blood$`Patient Id`) %>%
  subset(`Patient Id` %in% baseline$`Patient Id`)

train.outcomes <- demographics %>%
  select(`Patient Id`, irAE, Response) %>%
  subset(`Patient Id` %in% train$`Patient Id`) %>%
  subset(`Patient Id` %in% blood$`Patient Id`) %>%
  subset(`Patient Id` %in% baseline$`Patient Id`)

train.outcomes <- arrange(train.outcomes, desc(`Patient Id`))


test.outcomes <- demographics %>%
  select(`Patient Id`, irAE, Response) %>%
  subset(`Patient Id` %in% test$`Patient Id`) %>%
  subset(`Patient Id` %in% blood$`Patient Id`) %>%
  subset(`Patient Id` %in% baseline$`Patient Id`)

test.outcomes <- arrange(test.outcomes, desc(`Patient Id`))
```


Data for when using microbiome
```{r}
#Making data objects for microbe only model
micro <- baseline %>%
  select(sample, `Patient Id`) %>%
  full_join(base.micro, baseline, by = "sample") %>%
  select(-sample)

micro <- micro %>%
  # select(all_of(biome_list))
  # select(all_of(irAE_list))
  # select(all_of(response_list))
  # select(all_of(struct0.recist.list))
  # select(all_of(struct0.irae.list))
  # select(all_of(irAEcombinedlist))
  select(all_of(responsecombinedlist))

train.seq <- filter(micro,
                    (`Patient Id` %in% train.outcomes$`Patient Id`)) 

train.seq <- arrange(train.seq, desc(`Patient Id`))



test.seq <- filter(micro,
                    (`Patient Id` %in% test.outcomes$`Patient Id`))  
test.seq <- arrange(test.seq, desc(`Patient Id`))



```

Data for when using blood
```{r}
#Making data objects for blood only model
train.seq <- filter(base.gene,
                    (`Patient Id` %in% train.outcomes$`Patient Id`))
train.seq <- arrange(train.seq, desc(`Patient Id`))


test.seq <- filter(base.gene,
                    (`Patient Id` %in% test.outcomes$`Patient Id`))
test.seq <- arrange(test.seq, desc(`Patient Id`))
```


joining blood and biome data
```{r}
#Making data objects for microbe and blood model
micro <- baseline %>%
  select(sample, `Patient Id`) %>%
  full_join(base.micro, baseline, by = "sample") %>%
  select(-sample)

micro <- micro %>%
  # select(all_of(biome_list))
  # select(all_of(irAE_list))
  select(all_of(response_list))
  # select(all_of(struct0.recist.list))
  # select(all_of(struct0.irae.list))
  # select(all_of(irAEcombinedlist))
  # select(all_of(responsecombinedlist))

blood.and.biome <-  full_join(micro, base.gene, by = "Patient Id") %>%
  subset(`Patient Id` %in% baseline$`Patient Id`)

train.seq <- filter(blood.and.biome,
                    (`Patient Id` %in% train.outcomes$`Patient Id`)) 
train.seq <- arrange(train.seq, desc(`Patient Id`))


test.seq <- filter(blood.and.biome,
                    (`Patient Id` %in% test.outcomes$`Patient Id`))
test.seq <- arrange(test.seq, desc(`Patient Id`))

```


for adding BMI and age to data
```{r}



#selecting sheet with progressed/treatment completed
met_Datasheet <- readWorkbook(adjing.data, sheet = "Met Datasheet", startRow = 2)
met_Datasheet <- subset(met_Datasheet, Cohort != "Excluded") %>%
  select(`Coded.Sample.#`, `BMI.(kg/m2)`)
met_Datasheet$`Coded.Sample.#` <- gsub( " *\\(.*?\\) *", "", met_Datasheet$`Coded.Sample.#`)
met_Datasheet$`Coded.Sample.#` <- gsub( " *\\(.*?\\)*", "", met_Datasheet$`Coded.Sample.#`)
met_Datasheet$`Coded.Sample.#` <-gsub("- ", "-", met_Datasheet$`Coded.Sample.#`)
met_Datasheet$`Coded.Sample.#` <-gsub("^\\s+|\\s+$", "", met_Datasheet$`Coded.Sample.#`)
met_Datasheet <- met_Datasheet %>% rename(`MDSC #` = `Coded.Sample.#`)

demographics <- full_join(met_Datasheet, demographics)

add.data <- demographics %>%
  subset(`Patient Id` %in% blood$`Patient Id`) %>%
  subset(`Patient Id` %in% baseline$`Patient Id`) %>%
  select(`Patient Id`, Age, `BMI.(kg/m2)`) 


train.seq <- full_join(train.seq, add.data) %>%
  subset(`Patient Id` %in% train$`Patient Id`)

test.seq <- full_join(test.seq, add.data) %>%
  subset(`Patient Id` %in% test$`Patient Id`)

```




microbiome only model
```{r}
source("../mitox_functions.R")

#irAE
# rf_responce <- kTest(seed_list, "No", "Yes", "irAE", 500)
# 
# aucprs <- grabVals(aucprs, rf_responce, seed_list)
# 
# micro_only_irAE_500_qzerofive_strzero <- mean(aucprs$AUCPR)
# 
# micro_only_irAE_500_qzerofive_strzero <- as.data.frame(micro_only_irAE_500_qzerofive_strzero) %>%
#   rename(AUCPR = micro_only_irAE_500_qzerofive_strzero) %>%
#   add_column("name" = "micro_only_irAE_500_qzerofive_strzero", .before = "AUCPR")
# 
# irAE_aucprs <- irAE_aucprs  %>%
#   add_row(micro_only_irAE_500_qzerofive_strzero)

#Response
rf_responce <- kTest(seed_list, "Non-Responder", "Responder", "Response", 2000)

aucprs <- grabVals(aucprs, rf_responce, seed_list)

micro_only_Response_2000_qzerofive_strzero <- mean(aucprs$AUCPR)

micro_only_Response_2000_qzerofive_strzero <- as.data.frame(micro_only_Response_2000_qzerofive_strzero) %>%
  rename(AUCPR = micro_only_Response_2000_qzerofive_strzero) %>%
  add_column("name" = "micro_only_Response_2000_qzerofive_strzero", .before = "AUCPR")

Response_aucprs <- Response_aucprs  %>%
  add_row(micro_only_Response_2000_qzerofive_strzero)

```

blood/naonstring only model
```{r}
source("../mitox_functions.R")

#irAE
rf_responce <- kTest(seed_list, "No", "Yes", "irAE", 500)

aucprs <- grabVals(aucprs, rf_responce, seed_list)

blood_only_irAE_500 <- mean(aucprs$AUCPR)

#Response
rf_responce <- kTest(seed_list, "Non-Responder", "Responder", "Response", 2000)

aucprs <- grabVals(aucprs, rf_responce, seed_list)

blood_only_Response_2000 <- mean(aucprs$AUCPR)

```


microbiome+blood/naonstring model
```{r}
source("../mitox_functions.R")

#irAE
# rf_responce <- kTest(seed_list, "No", "Yes", "irAE", 500)
# 
# aucprs <- grabVals(aucprs, rf_responce, seed_list)
# 
# micro_and_blood_irAE_500_qzerofive_strzero <- mean(aucprs$AUCPR)
# 
# micro_and_blood_irAE_500_qzerofive_strzero <- as.data.frame(micro_and_blood_irAE_500_qzerofive_strzero) %>%
#   rename(AUCPR = micro_and_blood_irAE_500_qzerofive_strzero) %>%
#   add_column("name" = "micro_and_blood_irAE_500_qzerofive_strzero", .before = "AUCPR")
# 
# irAE_aucprs <- irAE_aucprs  %>%
#   add_row(micro_and_blood_irAE_500_qzerofive_strzero)


#Response
rf_responce <- kTest(seed_list, "Non-Responder", "Responder", "Response", 2000)

aucprs <- grabVals(aucprs, rf_responce, seed_list)

micro_and_blood_Response_2000_qzerofive_strzero <- mean(aucprs$AUCPR)

micro_and_blood_Response_2000_qzerofive_strzero <- as.data.frame(micro_and_blood_Response_2000_qzerofive_strzero) %>%
  rename(AUCPR = micro_and_blood_Response_2000_qzerofive_strzero) %>%
  add_column("name" = "micro_and_blood_Response_2000_qzerofive_strzero", .before = "AUCPR")

Response_aucprs <- Response_aucprs  %>%
  add_row(micro_and_blood_Response_2000_qzerofive_strzero)


```



```{r}
saveRDS(irAE_aucprs, file = "../data/rf_mitox_irAE_aucpr.RData")
irAE_aucprs <- readRDS("../data/rf_mitox_irAE_aucpr.RData")


saveRDS(Response_aucprs, file = "../data/rf_mitox_Response_aucpr.RData")
Response_aucprs <- readRDS("../data/rf_mitox_Response_aucpr.RData")
# Response_aucprstest <- Response_aucprs


# micro_only_Response_2000_factor <- pull(Response_aucprs[19,2])
```







