---
title: "try simulating data"
author: "Caroline Wheeler"
date: "2024-07-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(lme4)
library(mixedpower)
library(simr)
library(tidyverse)
```

```{r}
x <- read.csv("../data/2024-07-02_irAE mouse weights-cleaned.csv") 
```

give uniq num id to mice
```{r}
x <- x %>%
    mutate(subject = as.numeric(factor(mouse)))
```


```{r}
lmm.model <- lmer(weight ~ group*days + (1|subject), data = x)
summary(lmm.model)
```

```{r}
power_FLP <- mixedpower(model = lmm.model, 
                        data = x,
                        fixed_effects = c("group", "days") ,
                        simvar = "subject", 
                        steps =  c(8, 10, 12, 14, 16), # which sample sizes do we want to look at?
                        critical_value = 2, # which t/z value do we want to use to test for significance?
                        n_sim = 1000) # how many single simulations should be used to estimate power?

power_FLP
```
                      8    10    12    14    16      mode            effect
groupvehicle      0.458 0.532 0.559 0.649 0.701 databased      groupvehicle
days              0.875 0.942 0.968 0.968 0.984 databased              days
groupvehicle:days 1.000 1.000 1.000 1.000 1.000 databased groupvehicle:days

```{r}
multiplotPower(power_FLP)
```


```{r}
SESOI <- c(3.66, 0.75, -0.065, 0.09) # specify SESOI - smallest effect size of interest??

power_SESOI <- mixedpower(model = lmm.model, 
                        data = x,
                        fixed_effects = c("group", "days") ,
                        simvar = "subject", 
                        steps =  c(8, 10, 12, 14, 16), # which sample sizes do we want to look at?
                        critical_value = 2, # which t/z value do we want to use to test for significance?
                        n_sim = 1000,
                        SESOI = SESOI,
                        databased = T) # how many single simulations should be used to estimate power?

power_SESOI
```

                       8    10    12    14    16      mode            effect
groupvehicle       0.465 0.524 0.562 0.621 0.678 databased      groupvehicle
days               0.863 0.946 0.970 0.969 0.985 databased              days
groupvehicle:days  1.000 1.000 1.000 1.000 1.000 databased groupvehicle:days
groupvehicle1      0.219 0.225 0.259 0.277 0.317     SESOI      groupvehicle
days1              1.000 1.000 1.000 1.000 1.000     SESOI              days
groupvehicle:days1 1.000 1.000 1.000 1.000 1.000     SESOI groupvehicle:days


```{r}
multiplotPower(power_SESOI)
```

