---
title: "try simulating data"
author: "Caroline Wheeler"
date: "2024-07-15"
output: html_document
---

```{r setup, include=FALSE, warning=F}
knitr::opts_chunk$set(echo = TRUE)

library(lme4)
library(mixedpower)
library(simr)
library(tidyverse)
library(ggpubr)
```

```{r}
x <- read.csv("../data/2024-07-02_irAE mouse weights-cleaned.csv") 
```

give uniq num id to mice
```{r}
x <- x %>%
    mutate(subject = as.numeric(factor(mouse)))
```

Model with random intercepts 
```{r}
lmm.model <- lmer(weight ~ group*days + (1|subject), data = x)
summary(lmm.model)
```


Need to run separately for each sample size since critical value varies so much 
sample size is small 

critical values based on n-1, 95% CI on the two-tailed t-table.
https://www.scribbr.com/wp-content/uploads/2022/06/Critical-values-of-t-for-two-tailed-tests-l.webp

should this vary based on variable?
i.e. DF for each effect ??

should it be n1 + n2 - 2 OR n-1 ?
Going with n - 1 to be conservative 


Get databased and SESOI -15%
```{r}
# desired beta coefficients for all specified fixed effects
# try reducing all beta coefficients by 15% 
# leave intercept the same 
SESOI <- c(18.863, -1.147, -0.012,  0.057) 

power3 <- mixedpower(model = lmm.model, 
                        data = x,
                        fixed_effects = c("group", "days") ,
                        simvar = "subject", 
                        steps =  c(3), 
                        critical_value = 4.303, 
                        n_sim = 1000,
                        SESOI = SESOI,
                        databased = T) 

power4 <- mixedpower(model = lmm.model, 
                        data = x,
                        fixed_effects = c("group", "days") ,
                        simvar = "subject", 
                        steps =  c(4), 
                        critical_value = 3.182, 
                        n_sim = 1000,
                        SESOI = SESOI,
                        databased = T) 

power6 <- mixedpower(model = lmm.model, 
                        data = x,
                        fixed_effects = c("group", "days") ,
                        simvar = "subject", 
                        steps =  c(6), 
                        critical_value = 2.571, 
                        n_sim = 1000,
                        SESOI = SESOI,
                        databased = T) 

power8 <- mixedpower(model = lmm.model, 
                        data = x,
                        fixed_effects = c("group", "days") ,
                        simvar = "subject", 
                        steps =  c(8), 
                        critical_value = 2.365, 
                        n_sim = 1000,
                        SESOI = SESOI,
                        databased = T) 

power10 <- mixedpower(model = lmm.model, 
                        data = x,
                        fixed_effects = c("group", "days") ,
                        simvar = "subject", 
                        steps =  c(10), 
                        critical_value = 2.262, 
                        n_sim = 1000,
                        SESOI = SESOI,
                        databased = T) 

power12 <- mixedpower(model = lmm.model, 
                        data = x,
                        fixed_effects = c("group", "days") ,
                        simvar = "subject", 
                        steps =  c(12), 
                        critical_value = 2.201, 
                        n_sim = 1000,
                        SESOI = SESOI,
                        databased = T) 

```

combine
```{r}
df_list <- list(power3, power4, power6, power8, power10, power12)
power_SESOI <- Reduce(function(x, y) merge(x, y, all=TRUE, by = c("mode", "effect")), df_list, accumulate=FALSE)

power_SESOI
```

get SESOI -25%
```{r}
SESOI <- c(18.863,  -1.012, -0.011,  0.050) 

power3 <- mixedpower(model = lmm.model, 
                        data = x,
                        fixed_effects = c("group", "days") ,
                        simvar = "subject", 
                        steps =  c(3), 
                        critical_value = 4.303, 
                        n_sim = 1000,
                        SESOI = SESOI,
                        databased = F) 

power4 <- mixedpower(model = lmm.model, 
                        data = x,
                        fixed_effects = c("group", "days") ,
                        simvar = "subject", 
                        steps =  c(4), 
                        critical_value = 3.182, 
                        n_sim = 1000,
                        SESOI = SESOI,
                        databased = F) 

power6 <- mixedpower(model = lmm.model, 
                        data = x,
                        fixed_effects = c("group", "days") ,
                        simvar = "subject", 
                        steps =  c(6), 
                        critical_value = 2.571, 
                        n_sim = 1000,
                        SESOI = SESOI,
                        databased = F) 

power8 <- mixedpower(model = lmm.model, 
                        data = x,
                        fixed_effects = c("group", "days") ,
                        simvar = "subject", 
                        steps =  c(8), 
                        critical_value = 2.365, 
                        n_sim = 1000,
                        SESOI = SESOI,
                        databased = F) 

power10 <- mixedpower(model = lmm.model, 
                        data = x,
                        fixed_effects = c("group", "days") ,
                        simvar = "subject", 
                        steps =  c(10), 
                        critical_value = 2.262, 
                        n_sim = 1000,
                        SESOI = SESOI,
                        databased = F) 

power12 <- mixedpower(model = lmm.model, 
                        data = x,
                        fixed_effects = c("group", "days") ,
                        simvar = "subject", 
                        steps =  c(12), 
                        critical_value = 2.201, 
                        n_sim = 1000,
                        SESOI = SESOI,
                        databased = F) 
```
combine SESOI -25%
```{r}
df_list <- list(power3, power4, power6, power8, power10, power12)
power_SESOI2 <- Reduce(function(x, y) merge(x, y, all=TRUE, by = c("mode", "effect")), df_list, accumulate=FALSE) %>%
  mutate(mode = "SESOI-25%")

power_SESOI <- rbind(power_SESOI, power_SESOI2) 
power_SESOI
```

get SESOI -50%
```{r}
SESOI <- c(18.863, -0.675, -0.007, 0.033) 

power3 <- mixedpower(model = lmm.model, 
                        data = x,
                        fixed_effects = c("group", "days") ,
                        simvar = "subject", 
                        steps =  c(3), 
                        critical_value = 4.303, 
                        n_sim = 1000,
                        SESOI = SESOI,
                        databased = F) 

power4 <- mixedpower(model = lmm.model, 
                        data = x,
                        fixed_effects = c("group", "days") ,
                        simvar = "subject", 
                        steps =  c(4), 
                        critical_value = 3.182, 
                        n_sim = 1000,
                        SESOI = SESOI,
                        databased = F) 

power6 <- mixedpower(model = lmm.model, 
                        data = x,
                        fixed_effects = c("group", "days") ,
                        simvar = "subject", 
                        steps =  c(6), 
                        critical_value = 2.571, 
                        n_sim = 1000,
                        SESOI = SESOI,
                        databased = F) 

power8 <- mixedpower(model = lmm.model, 
                        data = x,
                        fixed_effects = c("group", "days") ,
                        simvar = "subject", 
                        steps =  c(8), 
                        critical_value = 2.365, 
                        n_sim = 1000,
                        SESOI = SESOI,
                        databased = F) 

power10 <- mixedpower(model = lmm.model, 
                        data = x,
                        fixed_effects = c("group", "days") ,
                        simvar = "subject", 
                        steps =  c(10), 
                        critical_value = 2.262, 
                        n_sim = 1000,
                        SESOI = SESOI,
                        databased = F) 

power12 <- mixedpower(model = lmm.model, 
                        data = x,
                        fixed_effects = c("group", "days") ,
                        simvar = "subject", 
                        steps =  c(12), 
                        critical_value = 2.201, 
                        n_sim = 1000,
                        SESOI = SESOI,
                        databased = F) 
```

combine SESOI -50%
```{r}
df_list <- list(power3, power4, power6, power8, power10, power12)
power_SESOI2 <- Reduce(function(x, y) merge(x, y, all=TRUE, by = c("mode", "effect")), df_list, accumulate=FALSE) %>%
  mutate(mode = "SESOI-50%")

power_SESOI <- rbind(power_SESOI, power_SESOI2) 
power_SESOI
```


```{r}
power_SESOI <- power_SESOI %>%
  mutate(mode = case_when(mode == "databased" ~ "Model",
                          mode == "SESOI" ~ "SESOI-15%",
                          mode == "SESOI-25%" ~ mode,
                          mode == "SESOI-50%" ~ mode),
         effect = case_when(effect == "groupvehicle" ~ "group",
                            effect == "groupvehicle:days" ~ "group:days",
                            effect == "days" ~ effect))
```

```{r}
mplot <- function(output_data){
  
  plot_data_all <- output_data %>%
    pivot_longer(!c("mode", "effect"), names_to = "sample_size", values_to = "value") %>%
    mutate(sample_size = as.integer(sample_size))
  # plot_data_all <- reshape2::melt(output_data,
  #                       id.vars = c("mode", "effect"),
  #                       variable.name = "sample_size")

  fixeffects <- unique(output_data$effect)

  cbPalette <- c("#882255", "#E69F00", "#6699CC","#999933")

  all_plots <- list()

  i <- 1 
  for (fixeff in fixeffects){

    plot_data <- subset(plot_data_all,  effect == fixeff,
                        select = c(mode, sample_size,value))


    p <- ggplot2::ggplot(plot_data, ggplot2::aes(x = sample_size, y = value, color=mode, group = mode, alpha = mode)) + 
      ggplot2::geom_point() + 
      ggplot2::geom_line()+
      ggplot2::xlab("mice per group") + 
      ggplot2::ylab("power") + 
      ggplot2::theme(axis.line = ggplot2::element_line(color = "black"), 
            panel.background = ggplot2::element_rect(fill = "white"), 
            legend.title = ggplot2::element_blank()) + 
      ggplot2::scale_color_manual(values = cbPalette) +
      ggplot2::expand_limits(y = c(0,1)) + 
      ggplot2::ggtitle(fixeff) +
      ggplot2::scale_alpha_discrete(range = c(1, 0.35, 0.35, 0.35)) + 
      ggplot2::scale_x_continuous(breaks=c(3, 4, 6, 8, 10, 12)) +
      aes(group=rev(mode)) 

    all_plots[[i]] <- p
    i <- i+1
  }


  effect_num = length(fixeffects)
  ncol = 3
  if (effect_num%%3 == 0) {
    nrow = effect_num/3
  } else {
    nrow = floor(effect_num/3) + 1
  }

  return(all_plots)

} 
```

```{r}
p <- mplot(power_SESOI)

ggarrange(p[[1]], p[[2]], p[[3]], ncol = 3)

ggsave("../figures/powerSimulation.png", height = 2, width = 10, dpi = 400)
```


