---
title: "core score exploration"
author: "Caroline Dravillas"
date: "2026-05-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

# Load data
```{r}
load("../../manuscript/data/counts-wide.rds")
load("../../manuscript/data/ra-wide.rds")
load("../../manuscript/data/microbiome_metadata.rds")
load("../../manuscript/data/clinical.rds")
load("../../manuscript/data/events.rds")
```

```{r}
irAE.time <- events %>%
  filter(Event == "irAE") %>%
  mutate(irAE.day = Days) %>%
  dplyr::select(`Patient Id`, irAE.day) 

# get irae samples 
meta.irae <- mic.meta %>%
  left_join(demographics) %>%
  left_join(irAE.time) %>%
  mutate(irAE.day = ifelse(irAE.day < 0 | irAE.day > 100, NA, irAE.day),
         irAE.sample = ifelse(is.na(irAE.day), 0,
                       ifelse(abs(Days - irAE.day) <= 10, 1, 0))) %>%
  group_by(sample) %>%
  slice_max(irAE.sample, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(sample, `Patient Id`, irAE.sample)

# pts with more than 2 samples 
pts <- mic.meta |>
  group_by(`Patient Id`) |>
  filter(n() > 2) |>
  ungroup() |>
  left_join(meta.irae) |>
    group_by(`Patient Id`) %>%
  # Step 1: Keep all irAE samples
  mutate(irae_keep = ifelse(irAE.sample == 1, TRUE, FALSE)) %>%
  # Step 2: Count how many slots are left per patient
  group_modify(~{
    patient_data <- .
    n_keep <- sum(patient_data$irae_keep)
    n_slots <- 3 - n_keep
    
    if (n_slots <= 0) {
      # already have 3 or more irAE samples, keep only these
      return(patient_data %>% filter(irae_keep))
    } else {
  remaining <- patient_data %>% filter(!irae_keep)

  if (nrow(remaining) <= n_slots) {

    selected_remaining <- remaining

  } else if (n_slots == 1) {

    ## selects max here b/c thats what is farther from the irAE samples we 
    ## have but this will need to be edited in future 
    selected_remaining <- remaining %>%
      slice_max(Days, n = 1, with_ties = FALSE)

  } else if (n_slots == 2) {

    selected_remaining <- remaining %>%
      slice_min(Days, n = 1, with_ties = FALSE) %>%
      bind_rows(
        remaining %>%
          slice_max(Days, n = 1, with_ties = FALSE)
      ) %>%
      distinct()

  } else if (n_slots == 3) {

    selected_remaining <- remaining %>%
      slice_min(Days, n = 1, with_ties = FALSE) %>%
      bind_rows(
        remaining %>%
          slice_max(Days, n = 1, with_ties = FALSE)
      ) %>%
      distinct()

    middle <- remaining %>%
      filter(!Days %in% selected_remaining$Days) %>%
      slice(1)

    selected_remaining <- bind_rows(selected_remaining, middle)
  }

  bind_rows(patient_data %>% filter(irae_keep), selected_remaining)
}
  }) %>%
  ungroup() %>%
  select(-irae_keep)

# n = 26
resp <- demographics |>
  select(`Patient Id`, Response, irAE) |>
  filter(`Patient Id` %in% pts$`Patient Id`)

ra.spec <- as.matrix(ra.w |>
  filter(grepl("s__", microbe)) |>
  select(microbe, pts$sample) |>
  column_to_rownames("microbe"))

all.equal(colnames(ra.spec), pts$sample)
```

Binarize the abundance matrix at species level
For each pt with > 1 sample
Drop microbes with 0’s in each sample
What fraction of the microbes are found in both (all) samples
Core.fraction = Consistent microbes / total microbes

```{r}
# binarize
mat_bin <- (ra.spec > 0) * 1
patient_ids <- unique(pts$`Patient Id`)
```

Now instead of a per patient score, I want a per species score. 

First lets do a strict version 
Core 3/3​= #people where species present in all 3​/ #people

Then a more relaxed version
Core 2/3​ = #people where species present in at least 2 of 3​/ #people
```{r}
spec_core_score <- function(mat_bin, pts, sample_set) {
  
  patient_ids <- unique(sample_set$`Patient Id`)
  
  species_core <- lapply(rownames(mat_bin), function(sp) {
    
    vals3 <- sapply(patient_ids, function(pid) {
      cols <- which(sample_set$`Patient Id` == pid)
      if (length(cols) == 0) return(NA)
      
      subvec <- mat_bin[sp, cols, drop = TRUE]
      sum(subvec) == length(cols)
    })
    
    vals2 <- sapply(patient_ids, function(pid) {
      cols <- which(sample_set$`Patient Id` == pid)
      if (length(cols) == 0) return(NA)
      
      subvec <- mat_bin[sp, cols, drop = TRUE]
      sum(subvec) >= ceiling(length(cols) * 2/3)
    })
    
    data.frame(
      species = sp,
      core_3of3 = mean(vals3, na.rm = TRUE),
      core_2of3 = mean(vals2, na.rm = TRUE)
    )
  })
  
  bind_rows(species_core) |>
    mutate(delta = core_2of3 - core_3of3) |>
    arrange(desc(core_3of3))
}
```


```{r}
species_core <- spec_core_score(mat_bin, pts, pts) |>
  mutate(group = "all pts w/ 3 samples")

head(species_core, 10)

write.csv(species_core, "../data/core-species-all3samps.csv", row.names = F)
```

people w/o irAE & 3 samples 
```{r}
pts <- pts |>
  left_join()
no.irae.pts <- resp |>
  filter(irAE == "No",
         Response == "")
```

responders w/ 3 samples 
```{r}

```


Then lets do the same with all people with 2 samples...but then i am only
interested in presence in 2/2 samples 

```{r}

```

