---
title: "aucpr_and_cfm_template"
author: "Noah Lepola"
date: "2025-02-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggplot2)
library(randomForest)
library(ROCR)
library(caTools)
library(openxlsx)
library(respondersig) 
library(PRROC)
library(glmnet)
library(caret)
```

Loading data
```{r}
source("../scripts/mitox_functions.R")
# irAE_aucprs <- readRDS("../data/rf_mitox_irAE_aucpr.RData")
# Response_aucprs <- readRDS("../data/rf_mitox_Response_aucpr.RData")
load("../../manuscript/data/ra-long.rds")
load("../../manuscript/data/clinical.rds")
load("../../manuscript/data/blood.rds")
load("../../manuscript/data/microbiome_metadata.rds")
load("../../manuscript/data/phylum-labels.rds")
load("../../manuscript/data/ANCOM-BC2_response_baseline.Rdata")
load("../../manuscript/data/ANCOM-BC2_irae_baseline.RData")
load("../../manuscript/data/core-score-min3.rds")

# iraeAUCPRavgs <- data.frame(name = NA, AUCPR = NA)
```

```{r}
seed_list <- c(12345:12369)
```

# Making objects for q value models
```{r}
#Object to limit microbiome sample to one ones taken within 10 days of treatment
baseline <- mic.meta %>%
  group_by(`Patient Id`) %>%
  arrange(Days) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  filter(abs(Days) <= 10) %>%
  subset(`Patient Id` %in% blood$`Patient Id`)

phyl_labels <- phyl_labels %>%
  mutate(taxon = substr(microbe, 4, nchar(microbe)))

#creating object for use in biome model
base.micro <- ra.l %>%
  pivot_wider(names_from = microbe, values_from = RelAbun) %>%
  subset(sample %in% baseline$sample)

full_micro_list <- colnames(base.micro)
full_micro_list[[1]] <- "Patient Id"

#note it is dropping bacteria and archaea becaise it isn't in phyl_labels
#making object for irAE where q < 0.01
irAE_q <- res.irae %>%
  left_join(phyl_labels) %>%
  dplyr::select(microbe, q_irAEYes)  %>%
  filter(q_irAEYes < 0.01) %>%
  replace_na(list( microbe = "k__Archaea"))

irae_list <- irAE_q$microbe %>%
  append("Patient Id", after = 0)

irae_list[[1]] <- "Patient Id"

#making object for Response where q < 0.01
Response_q <- res.recist %>%
  left_join(phyl_labels) %>%
  dplyr::select(microbe, q_ResponseResponder)  %>%
  filter(q_ResponseResponder < 0.01) %>%
  replace_na(list( microbe = "k__Archaea"))



response_list <- Response_q$microbe %>%
  append("Patient Id", after = 0)

response_list[[1]] <- "Patient Id"
```

# Making object for blood genes
```{r}
#limiting blood samples to ones also with microbiome sample and creating object
base.gene <- blood %>%
  subset(`Patient Id` %in% baseline$`Patient Id`) %>%
  dplyr::select(`Patient Id`:YWHAZ)

blood_list <- colnames(base.gene)
irae_blood_list <- c(irae_list, blood_list)
response_blood_list <- c(response_list, blood_list)
```

# clinical objects
```{r}
base.clinical <- demographics %>%
  dplyr::select(`Patient Id`, 
         Sex,
         Age,
         Treatment,
         Stage,
         `Total.WBC.(K/µL)`,
         `Relative.lymphocyte.count.(%)`, 
         `Relative.eosinophil.count.(%)`, 
         `LDH.(U/L)`, 
         `BMI.(kg/m2)`)

base.clinical <- base.clinical %>% 
  mutate_at(c("Sex",
              "Treatment", 
              "Stage"), as.factor)

clinical_list <- colnames(base.clinical)

irae_clinical_list <- c(irae_list, clinical_list)
blood_clinical_list <- c(blood_list, clinical_list)

response_clinical_list <- c(irae_list, clinical_list)
blood_clinical_list <- c(blood_list, clinical_list)


irae_blood_clinical_list <- c(irae_blood_list, clinical_list)
response_blood_clinical_list <- c(response_blood_list, clinical_list)
```

# Making object with all factors
```{r}
all_factors <- baseline %>%
  dplyr::select(sample, `Patient Id`) %>%
  full_join(base.micro, baseline, by = "sample") %>%
  full_join(base.gene, baseline, by = "Patient Id") %>%
  left_join(base.clinical, baseline, by = "Patient Id") %>%
  dplyr::select(-sample)
```


### irAE 

# Clinical model
```{r}
rf_clinical_irAE_list <- kTest(seed_list,
                               "No",
                               "Yes",
                               "irAE",
                               clinical_list,
                               500)

clinical_irae_aucprs <- grabVals(rf_clinical_irAE_list,
                            seed_list)
clinical_irae_aucprs$Predictors <- "Clinical"

irAE_clinical <- mean(clinical_irae_aucprs$AUCPR)
```

# Full biome model
```{r}
rf_fullbiome_irAE_list <- kTest(seed_list,
                          "No",
                          "Yes",
                          "irAE",
                          full_micro_list,
                          500)

fullbiome_irae_aucprs <- grabVals(rf_fullbiome_irAE_list,
                              seed_list)
fullbiome_irae_aucprs$Predictors <- "Full Microbiome"

irAE_fullbiome <- mean(fullbiome_irae_aucprs$AUCPR)
```

# Limited biome model
```{r}
rf_q01_irAE_list <- kTest(seed_list,
                          "No",
                          "Yes",
                          "irAE",
                          irae_list,
                          500)

biome_irae_aucprs <- grabVals(rf_q01_irAE_list,
                         seed_list)
biome_irae_aucprs$Predictors <- "Microbiome"

irAE_q01 <- mean(biome_irae_aucprs$AUCPR)
```

# Just blood model
```{r}
rf_blood_irAE_list <- kTest(seed_list,
                            "No",
                            "Yes",
                            "irAE",
                            blood_list,
                            500)

blood_irae_aucprs <- grabVals(rf_blood_irAE_list,
                         seed_list)
blood_irae_aucprs$Predictors <- "Blood"

irAE_blood <- mean(blood_irae_aucprs$AUCPR)
```

# Biome and blood model
```{r}
rf_biome_blood_irAE_list <- kTest(seed_list,
                                  "No",
                                  "Yes",
                                  "irAE",
                                  irae_blood_list,
                                  500)

biome_and_blood_irae_aucprs <- grabVals(rf_biome_blood_irAE_list,
                                   seed_list)
biome_and_blood_irae_aucprs$Predictors <- "Blood + Microbiome"

irAE_biome_and_blood <- mean(biome_and_blood_irae_aucprs$AUCPR)
```

# Blood and clinical model
```{r}
rf_blood_clinical_irAE_list <- kTest(seed_list,
                                  "No",
                                  "Yes",
                                  "irAE",
                                  blood_clinical_list,
                                  500)

blood_and_clinical_irae_aucprs <- grabVals(rf_blood_clinical_irAE_list,
                                   seed_list)
blood_and_clinical_irae_aucprs$Predictors <- "Clinical + Blood"

irAE_blood_and_clinical <- mean(blood_and_clinical_irae_aucprs$AUCPR)
```

# Biome and clinical model
```{r}
rf_biome_clinical_irAE_list <- kTest(seed_list,
                                  "No",
                                  "Yes",
                                  "irAE",
                                  irae_clinical_list,
                                  500)

biome_and_clinical_irae_aucprs <- grabVals(rf_biome_clinical_irAE_list,
                                   seed_list)
biome_and_clinical_irae_aucprs$Predictors <- "Clinical + Microbiome"

irAE_biome_and_clinical <- mean(biome_and_clinical_irae_aucprs$AUCPR)
```

# Biome blood clinical model
```{r}
rf_biome_blood_clinical_irAE_list <- kTest(seed_list,
                                           "No",
                                           "Yes",
                                           "irAE",
                                           irae_blood_clinical_list,
                                           500)

biome_blood_clinical_irae_aucprs <- grabVals(rf_biome_blood_clinical_irAE_list,
                                        seed_list)
biome_blood_clinical_irae_aucprs$Predictors <- "Clinical + Blood + Microbiome"

irAE_biome_blood_clinical <- mean(biome_blood_clinical_irae_aucprs$AUCPR)
```


## Combine and save as one df for irAE
```{r}
irae.rf <- rbind(clinical_irae_aucprs, blood_irae_aucprs, biome_irae_aucprs, fullbiome_irae_aucprs, biome_and_blood_irae_aucprs, blood_and_clinical_irae_aucprs, blood_and_clinical_irae_aucprs, biome_blood_clinical_irae_aucprs)

save(irae.rf, file = "../data/randomforest-irae-RFE.rds")
```


### Response

# Clinical model
```{r}

rf_clinical_response_list <- kTest(seed_list,
                                "Non-Responder",
                                "Responder",
                                "Response",
                                clinical_list,
                                500)

clinical_response_aucprs <- grabVals(rf_clinical_response_list,
                            seed_list)
clinical_response_aucprs$Predictors <- "Clinical"

Response_clinical <- mean(clinical_response_aucprs$AUCPR)
```

# Full biome model
```{r}
rf_fullbiome_response_list <- kTest(seed_list,
                                "Non-Responder",
                                "Responder",
                                "Response",
                                full_micro_list,
                                500)

fullbiome_response_aucprs <- grabVals(rf_fullbiome_response_list,
                              seed_list)
fullbiome_response_aucprs$Predictors <- "Full Microbiome"

Response_fullbiome <- mean(fullbiome_response_aucprs$AUCPR)
```

# Limited biome model
```{r}
rf_q01_response_list <- kTest(seed_list,
                              "Non-Responder",
                              "Responder",
                              "Response",
                              response_list,
                              500)

biome_response_aucprs <- grabVals(rf_q01_response_list,
                         seed_list)
biome_response_aucprs$Predictors <- "Microbiome"

Response_q01 <- mean(biome_response_aucprs$AUCPR)
```

# Just blood model
```{r}
rf_blood_response_list <- kTest(seed_list,
                                "Non-Responder",
                                "Responder",
                                "Response",
                                blood_list,
                                500)

blood_response_aucprs <- grabVals(rf_blood_response_list,
                         seed_list)
blood_response_aucprs$Predictors <- "Blood"

Response_blood <- mean(blood_response_aucprs$AUCPR)
```

# Biome and blood model
```{r}
rf_biome_blood_response_list <- kTest(seed_list,
                                  "Non-Responder",
                                  "Responder",
                                  "Response",
                                  response_blood_list,
                                  500)

biome_and_blood_response_aucprs <- grabVals(rf_biome_blood_response_list,
                                   seed_list)
biome_and_blood_response_aucprs$Predictors <- "Blood + Microbiome"

Response_biome_and_blood <- mean(biome_and_blood_response_aucprs$AUCPR)
```

# Biome clinical model
```{r}
rf_biome_clinical_response_list <- kTest(seed_list,
                                          "Non-Responder",
                                          "Responder",
                                          "Response",
                                          response_clinical_list,
                                          500)

biome_clinical_response_aucprs <- grabVals(rf_biome_clinical_response_list,
                                   seed_list)
biome_clinical_response_aucprs$Predictors <- "Clinical + Microbiome"

Response_biome_blood_clinical <- mean(biome_clinical_response_aucprs$AUCPR)
```

# Blood clinical model
```{r}
rf_blood_clinical_response_list <- kTest(seed_list,
                                          "Non-Responder",
                                          "Responder",
                                          "Response",
                                          blood_clinical_list,
                                          500)

blood_clinical_response_aucprs <- grabVals(rf_blood_clinical_response_list,
                                   seed_list)
blood_clinical_response_aucprs$Predictors <- "Clinical + Blood"

Response_biome_blood_clinical <- mean(blood_clinical_response_aucprs$AUCPR)
```

# Biome blood clinical model
```{r}
rf_biome_blood_clinical_response_list <- kTest(seed_list,
                                          "Non-Responder",
                                          "Responder",
                                          "Response",
                                          response_blood_clinical_list,
                                          500)

biome_blood_clinical_response_aucprs <- grabVals(rf_biome_blood_clinical_response_list,
                                   seed_list)
biome_blood_clinical_response_aucprs$Predictors <- "Clinical + Blood + Microbiome"

Response_biome_blood_clinical <- mean(biome_blood_clinical_response_aucprs$AUCPR)
```


## Combine and save as one df for Response
```{r}
response.rf <- rbind(clinical_response_aucprs, blood_response_aucprs, biome_response_aucprs, fullbiome_response_aucprs, biome_clinical_response_aucprs, blood_clinical_response_aucprs,  biome_and_blood_response_aucprs, biome_blood_clinical_response_aucprs)

save(response.rf, file = "../data/randomforest-response-RFE.rds")
```


